### 2009-11-29 doc/msc-proposal/msc-proposal.tex
added msc. proposal

\documentclass[12pt, titlepage, a4paper]{report}
% For printed document - always starts a new chapter on the right-side page (leaves blank pages):
%\documentclass[12pt, titlepage, a4paper, openright, twoside]{report}

%% Needed packages:
\usepackage{pdfpages} 		% used for introducing the title PDF page
%\usepackage{a4wide}		% spreads text over the whole width of the page
\usepackage[utf8x]{inputenc}	% encoding
\usepackage[english]{babel}	% multilingual typesetting
\usepackage{graphicx}		% adds graphics
\usepackage[pdftex]{hyperref}	% hyper-refs inside the doc
\usepackage{url}		% formats urls
\usepackage{cite}		% citations from bibtex
\usepackage{makeidx}		% index
\usepackage{robustindex}	% index

%% Add/remove packages here:
\usepackage{ucs}
\usepackage{parskip}
\usepackage{wrapfig}
\usepackage{array}
\usepackage{rotating}
\usepackage{alltt}
\usepackage{latexsym}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace} 
%\usepackage[hide]{ed}
\usepackage{color}
\usepackage{listings}
\lstset{float=htb,columns=fullflexible,frame=lines,basicstyle=\scriptsize, language=XML,
        numbers=left,stepnumber=5,numberstyle=\tiny,showstringspaces=false}

\pagestyle{headings}
\makeindex    % obligatory with robustindex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Document
\begin{document}

\begin{titlepage}
\includepdf[pages=1]{title}
\thispagestyle{empty}
\end{titlepage}

%% Acknowledgements page
\newpage
\thispagestyle{empty}

%% Abstract page
\newpage
\begin{abstract}
Abstract
\end{abstract}

\newpage
\tableofcontents
%\listoffigures 
\newpage

%%
%	Introduction
%%
\chapter{Introduction}
* LowPAN, possible applications (use cases), why it's important.\\
* The purpose of the proposed work will be to investigate smth. The proposal describes problems associated with enabling IP communication with devices in a LoWPAN.\\
* As preparation for this work, several related areas have been studied in more
depth and the collected background information; Structure of the proposal.\\

\chapter{LoWPAN Networks}
Low-power and lossy networks (LLNs) are the networks made of highly constrained nodes (limited CPU, memory, power) interconnected by a variety of "lossy" links (low-power radio links). Such networks can be characterized by low speed, low performance, low cost, and unstable connectivity.  A low-power wireless personal area network (LoWPAN) is a particular instance of an LLN, comprised devices that conform to the IEEE 802.15.4 standard \cite{ieee802.15.4}.

%%
%	Transmission of IPv6 Packets
%%
\chapter{IPv6 over LoWPAN Networks}

%%
%	Conclusion
%%
\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}

%% Bibliography
\newpage
\bibliographystyle{plain}
\bibliography{msc-proposal}
\addcontentsline{toc}{chapter}{Bibliography}

\nocite{ieee802.15.4} 
\nocite{rfc2460} 
\nocite{rfc4291} 
\nocite{rfc4443}
\nocite{rfc4861}
\nocite{rfc4862}
\nocite{rfc4919}
\nocite{rfc4944} 
\nocite{draft-usecases-05} 
\nocite{draft-hc-06} 
\nocite{draft-nd-06}

\end{document}

### 2009-11-29 doc/wrap-ups/hc-06/wrapup.tex
added msc. proposal

\documentclass[a4paper,11pt,oneside]{article}

\usepackage{parskip}                %% blank lines between paragraphs, no indent 
\usepackage[pdftex]{graphicx}        %% include graphics, preferably pdf 
\usepackage[pdftex]{hyperref}        %% many PDF options can be set here 
\usepackage{graphicx}    % needed for including graphics e.g. EPS, PS 
\usepackage {amssymb} 
\usepackage{hyperref}

\newenvironment{mylisting}
{\begin{list}{}{\setlength{\leftmargin}{1em}}\item\footnotesize\bfseries}
{\end{list}}

\newenvironment{mytinylisting}
{\begin{list}{}{\setlength{\leftmargin}{1em}}\item\tiny\bfseries}
{\end{list}}

\begin{document} 
\section{RFC 4944, Transmission of IPv6 Packets over IEEE 802.15.4 Networks}
RFC4944 \cite{rfc4944} defines a 6LoWPAN adaptation format to carry IPv6 datagrams over such constrained links, taking into account limited
bandwidth, memory, or energy resources that are expected in applications such as wireless sensor networks:
\begin{itemize}
\item stateless header compression for IPv6 datagrams (LOWPAN\_HC1 and LOWPAN\_HC2) to reduce the relatively large IPv6 and UDP headers down to (in the best case) several bytes;
\item Fragmentation header to support the IPv6 minimum MTU requirement \cite{rfc2460};
\item Mesh Addressing header to support Mesh Under sub-IP forwarding;
\end{itemize}

\section{ Internet Draft, Compression Format for IPv6 Datagrams in 6LoWPAN Networks} 
The draft \cite{draft-hc-06} specifies a header compression format that is intended
   to replace the LOWPAN\_HC1 defined in RFC4944 \cite{rfc4944}.

\subsection{Motivation}
\begin{itemize}
\item LOWPAN\_HC1 and LOWPAN\_HC2 are insufficient for most practical uses of 6LoWPAN networks.  LOWPAN\_HC1 is most effective for link-local unicast communication, where IPv6 addresses carry the link-local prefix and an Interface Identifier (IID) directly derived from IEEE 802.15.4 addresses \cite{rfc4291}.  In this case, both addresses may be completely elided.

\item Routable addresses must be used when communicating with devices
external to the LoWPAN or in a route-over configuration where IP
forwarding occurs within the LoWPAN.  For routable addresses,
LOWPAN\_HC1 requires both IPv6 source and destination addresses to
carry the prefix in-line. 
\end{itemize}

\subsection{New features} 
The draft defined two encoding formats:
\begin{itemize}
\item LOWPAN\_IPHC -- for effective compression of Unique Local, Global, and multicast IPv6
addresses based on shared state within contexts;

\item LOWPAN\_NHC -- for encoding arbitrary headers following the LOWPAN\_IPHC header.
\end{itemize}

\subsection{IPv6 Header Compression} 
The draft \cite{draft-hc-06} introduces LOWPAN\_IPHC compression format for the IPv6 header. LOWPAN\_IPHC assumes the following common case IPv6 header values for 6LoWPAN communication:
\begin{itemize}
\item Version is 6; 
\item Traffic Class and Flow Label are both zero; 
\item Payload Length can be inferred from lower layers from either the 6LoWPAN Fragmentation header or the IEEE 802.15.4 header;
\item Hop Limit will be set to a well-known value by the source; 
\item addresses assigned to 6LoWPAN interfaces will be formed using the link-local prefix or a single routable prefix assigned to the entire 6LoWPAN network; 
\item addresses assigned to 6LoWPAN interfaces are formed with an IID derived directly from either the 64-bit extended or 16-bit short IEEE 802.15.4 addresses.
\end{itemize}

The compression mechanism is adapted for these values of the header fields and  in such scenario the LOWPAN\_IPHC can compress the IPv6 header down to two octets (1-octet dispatch, 1-octet LOWPAN\_IPHC) with link-local communication.

\subsection{LOWPAN\_IPHC Encoding Format}
The IPv6 header encrypted with the LOWPAN\_IPHC encoding format consists of three parts (shown in Figure \ref{fig:lowpaniphc}):
\begin{itemize}
\item Dispatch type -- in case of LOWPAN\_IPHC is 3 bit value $110$;
\item LOWPAN\_IPHC -- encoding that describes how an IPv6 header is compressed;
\item Compressed IPv6 Header -- the IPv6 header fields that are not fully elided, either in a compressed form if the field is partially elided, or litteraly.
\end{itemize}
\begin{figure}[htp]
\centering
\begin{mylisting}
\begin{verbatim}
+----------+-------------+-----------------------+
| Dispatch | LOWPAN_IPHC | Compressed IPv6 Header|
+----------+-------------+-----------------------+
\end{verbatim}
\end{mylisting}
\caption{LOWPAN\_IPHC Header}\label{fig:lowpaniphc}
\end{figure}

The LOWPAN\_IPHC encoding (shown in Figure) \ref{fig:lowpanbe} usually utilizes 13 bits, but may be extended by another octet to support additional contexts. 
\begin{figure}[htp]
\centering
\begin{mylisting}
\begin{verbatim}
  0   1   2   3   4   5   6   7   8   9   0   1   2  
+---+---+---+---+---+---+---+---+---+---+---+---+---+
|  TF   |NH | HLIM  |CID|SAC|  SAM  | M |DAC|  DAM  |
+---+---+---+---+---+---+---+---+---+---+---+---+---+
\end{verbatim}
\end{mylisting}
\caption{LOWPAN\_IPHC Encoding}\label{fig:lowpanbe}
\end{figure}

The encoding supports 4 compression formats of the Traffic Class and Flow Label (TF field). The
NH field specifies whether the IPv6 Next header field is carried in-line or compressed using LOWPAN\_NHC (described in Section \ref{sec:ipnh}). The HLIM encodes the compression type for the Hop Limit field. The CID field indicates
whether an additional 8-bit Context Identifier Extension field immediately follows the DAM field.

The SAC, SAM, DAC and DAM fields describe how the source and destination addresses are compressed. The LOWPAN\_IPHC encoding supports stateless and stateful context-based compressions. The stateless compression makes use of the link-local prefixes (the value of those bits is the link-local prefix) and link-local addresses (derivable from the corresponding link-layer address). The stateful address compression can be applied to the source and destination IPv6 addresses when they do not statelessly match the source and destination link layer addresses.    
The stateful compression relies on a conceptual context which is shared between the node that compresses a packet and the node(s) that need to expand it. However, the draft does not specify how these shared contexts are established and how the information is maintained within them.

The M field specifies whether the Destination Address is a multicast address or not.

\subsection{IPv6 Next Header Compression} \label{sec:ipnh}
When the NH bit of LOWPAN\_IPHC is set to 1 the 6LoWPAN next header compression (LOWPAN\_NHC) is used. 
This mechanism is intented to replace the LOWPAN\_HC2 compression defined in \cite{rfc4944}. In contrast, LOWPAN\_NHC allows to define comparession formats for arbitrary next headers, whereas LOWPAN\_HC2 can be used only for UDP, TCP, and ICMPv6.

The value of IPv6 Next Header is recovered from the first bits in the LOWPAN\_NHC encoding and the remaining 
bits are specific to the IPv6 Next Header value. Compression formats for different next headers are identified by a
bit-pattern immediately following the LOWPAN\_IPHC compressed header and can have different length.
The draft document defines a set of LOWPAN\_NHC encodings for selected IPv6 Extension Headers (i.e. UDP Header Compression).
\begin{figure}[htp]
\centering
\begin{mylisting}
\begin{verbatim}
+-------------+-----------+-------------+---------------+--------
| LOWPAN_IPHC | In-line   | LOWPAN_NHC  | In-line Next  | Payload
|   Encoding  | IP Fields |   Encoding  | Header Fields |
+-------------+-----------+-------------+---------------+--------
\end{verbatim}
\end{mylisting}
\caption{Typical LOWPAN\_IPHC/LOWPAN\_NHC Header Configuration}\label{fig:lowpaniphc}
\end{figure}

\nocite{rfc2460} 
\nocite{rfc4291} 
\nocite{rfc4944} 
\nocite{draft-hc-06}

\bibliographystyle{plain} 
\bibliography{wrapup} 
\end{document}

### 2009-11-29 doc/wrap-ups/nd-06/wrapup.tex
added msc. proposal

\documentclass[a4paper,11pt,oneside]{article}

\usepackage{parskip}                %% blank lines between paragraphs, no indent 
\usepackage[pdftex]{graphicx}        %% include graphics, preferably pdf 
\usepackage[pdftex]{hyperref}        %% many PDF options can be set here 
\usepackage{graphicx}    % needed for including graphics e.g. EPS, PS 
\usepackage {amssymb} 
\usepackage{hyperref}

\newenvironment{mylisting}
{\begin{list}{}{\setlength{\leftmargin}{1em}}\item\footnotesize\bfseries}
{\end{list}}

\newenvironment{mytinylisting}
{\begin{list}{}{\setlength{\leftmargin}{1em}}\item\tiny\bfseries}
{\end{list}}

\title{6LoWPAN Neighbor Discovery}

\begin{document} 
\maketitle
\section{Summary}
\subsection{Questions}
\begin{itemize}
\item NC is sent by an ER as unicast, but there are no details what happens for the first time when the source address of the NR message is set to unspecified address.
\end{itemize}

\subsection{Recap: IPv6 ND bootstrapping}
\begin{itemize}
\item a node joins the solicited-node multicast address on the interface and then performs duplicate address detection (DAD) for the acquired link-local address by sending a solicited-node multicast message to the link;
\item after that the node sends multicast messages to the all-router address to solicit router advertisements;
\item once the host receives a valid router advertisement, it autoconfigures the IPv6 address with the advertised prefix in the router advertisement (RA);
\item the IPv6 routers usually send router advertisements periodically on the network to the all-node multicast address;
\item nodes send Neighbor Solicitation/Neighbor Advertisement messages to resolve the IPv6 address of the
destination on the link. These NS/NA messages are also often multicast messages;
\end{itemize}

\subsection{Terminology}
\begin{itemize}
\item Optimistic address -- an address that is assigned to an interface and available for use, subject to restrictions, while its uniqueness on a link is being verified. 
\item LoWPAN Edge Router -- an IPv6 router that interconnects the LoWPAN to another IP network.
\item Binding -- the association of the LoWPAN node IPv6 address and Owner Interface ID with associated Whiteboard and ND states including the remaining lifetime of that association.
\item Whiteboard -- a conceptual data structure which may be supported by Edge Routers.  The Whiteboard is used for      performing Duplicate Address Detection (DAD) and Neighbor Unreachability Detection (NUD) across the entire 	LoWPAN.  The Whiteboard contains bindings for LoWPAN nodes that contain, among others, an Owner Interface Identifier, Owner Nonce, IPv6 address, TID and a remaining lifetime of the binding.
\end{itemize}

\subsection{6LoWPAN ND Goals:}
\begin{itemize}
\item enables ND operations over an entire LoWPAN, even with non-transitive links and over multihop IP hops;
\item the efficient detection of duplicate addresses over entire LoWPANs;
\item optimizes ND operations for LoWPANs - minimize signalling by avoiding the use of multicast flooding and reducing the frequency of link scope multicast for ND messages inside the LoWPANs;
\item disseminate context information throughout the LoWPAN;
\item (??? What is it?) provides a mechanism for claim and defend addressing;
\end{itemize}

\subsection{6LoWPAN ND Assumptions:}
\begin{itemize}
\item link-local IPv6 addresses are derived from a unique identifier (e.g.  EUI-64);
\item there is typically a direct mapping between the IPv6 address IID and the link-layer address, thus address resolution is not normally required;

\end{itemize}

\section{6LoWPAN Neighbor Discovery}

IPv6 Neighbor Discovery (ND) \cite{rfc4861} provides several important functions such as Router Discovery, Address Resolution, Duplicate Address Detection, Redirect, Prefix and Parameter Discovery.

However, standard IPv6 Neighbor Discovery \cite{rfc4861}  has several problems when using it with 6LoWPAN.
LoWPAN links are non-transitive in their nature (A can reach B, and B can reach C, but A can't reach C), but
IPv6 ND was not designed for non-transitive wireless links. 
LoWPAN does not provide multicast capability at the link-layer, thus simulating multicast behavior by both using broadcast or sending a number of unicast messages, both expensive for the low-powered network and the low-processing capable nodes. Therefor multicast is not desirable in such a low-power, lossy wireless network (unnecessary waste of energy in LoWPANs), whereas IPv6 ND heavily uses it.

As a result, it was concluded that standard Neighbor Discovery is not suitable for LoWPANs as it is and a 6LoWPAN-specific ND definition would be useful. In 6LoWPAN Neighbor Discovery Internet Draft \cite{draft-nd-06} a new ND mechanism has been proposed which takes into account the characteristics of low-power, lossy wireless links. 6LoWPAN ND provides additions and optimizations to IPv6 ND specifically supporting 6LoWPAN.

\subsection{Bootstrapping}
Bootstrapping of a LoWPAN node consists of the following steps:

\begin{itemize}
\item A host first performs stateless address autoconfiguration of its link-local unicast address for each LoWPAN interface from its EUI-64. In order to join a LoWPAN, a node should listen for Route Advertisements (RA) from Edge Routers or LoWPAN Routers, or broadcast a Router Solicitation (RS) and receive RA responses from local routers. If a valid prefix is advertised in the RA, the host will also form an optimistic global unique address with stateless address autoconfiguration.

\item Next the node will attempt to perform initial registration with an Edge Router. Registration is performed by sending a unicast Node Registration (NR) message to a link-local Edge Router  or router. The NR contains the addresses the node wants to register.

\item The Edge Router replies with a Node Confirmation(NC) either  directly or through the relaying router. Routers only exist in Route Over configurations, and in pure Mesh Under configurations nodes are always within link-local scope of an Edge Router. The NC includes the set of addresses now confirmed to be bound to the Whiteboard of the ER.  The Host is now capable of using the LoWPAN fully, and the ER forwards on its behalf.
\end{itemize}

Once the bootstrapping has been accomplished the node may send packets to any IPv6 address inside or outside the LoWPAN. Address resolution is not required to be performed with neighbors, but instead the Interface Identifier (IID) part of the IPv6 address directly corresponds to a MAC address.  Neighbor Solicitation/Neighbor Advertisement (NS/NA) messages are not used.

The Whiteboard address bindings and assignments are soft, and thus must be renewed periodically as indicated by the lifetime of the binding.  This is achieved by periodically sending a new NR to the` ER.

\subsection{6LoWPAN-ND Messages}
6LoWPAN-ND  messages are all ICMPv6 \cite{rfc4443} messages and extend the capabilities of "The IPv6 Neighbor Discovery Protocol" \cite{rfc4861}.

\subsubsection{Node Registration/Confirmation Message}
The Node Registration (NR) and Node Confirmation (NC) messages are used by a node to register with an ER, and for the ER to confirm the binding. The Node Registration message is sent by the LoWPAN Node to the link-local unicast IPv6 address of an local Edge Router or LoWPAN Router. NR/NC messages may be sent over multiple IP hops within the LoWPAN by relaying routers. When relaying, a new message is created with an updated checksum, and a code is used to indicate relaying.

Address Options are included in the NR message for each IPv6 address to be registered, and included in the corresponding NC to indicate success. These options are also used for dissemination of context information when stateful compression of the IPv6 headers is used.

\subsubsection{Router Solicitation/Router Advertisement Message}
The Router Solicitation and Router Advertisement message formats are identical to to the RS and RA
message formats defined in RFC 4861 \cite{rfc4861}. The use of some flags is however defined in the 6LoWPAN context, and several additional new options are identified. RA messages are sent either to link-local all-nodes multicast, or to a link-local unicast address as a response to an RS.

\subsection{LoWPAN Node}
Instead of relying on multicast ND messages for DAD and neighbor unreachability detection, LoWPAN Nodes send unicast messages to an Edge Router in the LoWPAN which keeps a Whiteboard of all bound addresses from nodes attached to that ER. These functions are performed across the entire LoWPAN using the Whiteboard and it allows 6LoWPAN-ND to operate over asymmetric, non-transitive links and with sleeping nodes. Node complexity and energy
consumption are reduced as address resolution and the support of redirect are not required, ND traffic is reduced and nodes do not exchange NS/NA messages.

A LoWPAN node makes use of a few conceptual structures and variables, some of which are the following: 
\begin{itemize}
\item Prefix List  - the list of prefixes which are advertised in Router Advertisements;
\item Context List  - the list of context and their associated CID which are advertised in Router Advertisements;
\item Edge Router List  - the list of Edge Routers the node is registered with;
\end{itemize}

\subsection{LoWPAN Router}
LoWPAN Routers are used in a route over configuration where the network is composed of overlapping link-local scopes. As a result, the standard IPv6 ND defined in RFC4861 \cite{rfc4861} is extended to operate over such non-transitive LoWPAN links.

Network configuration parameters carried in Router Advertisements originated at edge routers and must disseminate to all routers and hosts within the LoWPAN. Every LoWPAN Router also serves as a relay for Node Registration/Confirmation messages between nodes and ERs separated by multiple IP hops.

\subsection{LoWPAN Edge Router}
Edge Routers are the routers that connect LoWPANs to an IPv6 infrastructure via backhaul or backbone links. Edge Routers support Whiteboard registration for LoWPAN Nodes within their subnets which contains the following fields for every registered node:
\begin{itemize}
\item IPv6 address --  the IPv6 address being registered.
\item Owner Interface Identifier -- the 64-bit OII of the LoWPAN Node's interface.
\item Owner Nonce -- the 32-bit nonce generated by a node upon booting is used for Duplicate OII detection.
\item Registration Age and Lifetime -- the registration age indicates how long ago the last registration flow took place.  When the age reaches the registration lifetime, the whiteboard entry is removed.
\end{itemize}

\nocite{rfc4443} 
\nocite{rfc4861} 
\nocite{draft-nd-06}

\bibliographystyle{plain} 
\bibliography{wrapup} 
\end{document}

### 2009-11-29 trunk/doc/msc-proposal/msc-proposal.tex
proposal: IEEE 802.15.4

\documentclass[12pt, titlepage, a4paper]{report}
% For printed document - always starts a new chapter on the right-side page (leaves blank pages):
%\documentclass[12pt, titlepage, a4paper, openright, twoside]{report}

%% Needed packages:
\usepackage{pdfpages} 		% used for introducing the title PDF page
%\usepackage{a4wide}		% spreads text over the whole width of the page
\usepackage[utf8x]{inputenc}	% encoding
\usepackage[english]{babel}	% multilingual typesetting
\usepackage{graphicx}		% adds graphics
\usepackage[pdftex]{hyperref}	% hyper-refs inside the doc
\usepackage{url}		% formats urls
\usepackage{cite}		% citations from bibtex
\usepackage{makeidx}		% index
\usepackage{robustindex}	% index

%% Add/remove packages here:
\usepackage{ucs}
\usepackage{parskip}
\usepackage{wrapfig}
\usepackage{array}
\usepackage{rotating}
\usepackage{alltt}
\usepackage{latexsym}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace} 
%\usepackage[hide]{ed}
\usepackage{color}
\usepackage{listings}
\lstset{float=htb,columns=fullflexible,frame=lines,basicstyle=\scriptsize, language=XML,
        numbers=left,stepnumber=5,numberstyle=\tiny,showstringspaces=false}

\pagestyle{headings}
\makeindex    % obligatory with robustindex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Document
\begin{document}

\begin{titlepage}
\includepdf[pages=1]{title}
\thispagestyle{empty}
\end{titlepage}

%% Acknowledgements page
\newpage
\thispagestyle{empty}

%% Abstract page
\newpage
\begin{abstract}
Abstract
\end{abstract}

\newpage
\tableofcontents
%\listoffigures 
\newpage

%%
%	Introduction
%%
\chapter{Introduction}
* There are a lot of applications where we need many small devices + examples; why it's important;
* LoWPAN\\
* The purpose of the proposed work will be to investigate smth. The proposal describes problems associated with enabling IP communication with devices in a LoWPAN.\\
* As preparation for this work, several related areas have been studied in more
depth and the collected background information; Structure of the proposal.\\

\chapter{LoWPAN Networks}
A low-power wireless personal area network (LoWPAN) is a simple low cost communication network that allows wireless connectivity in applications with limited power and relaxed throughput requirements. LoWPAN networks comprises devices that conform to the IEEE 802.15.4 standard \cite{ieee802.15.4}. LoWPAN devices  are usually characterized as having limited processing capability (the smallest common LoWPAN nodes have 8-bit processors with clock rates around 10 MHz), small memory capacity (common RAM sizes for LoWPAN devices consist of a few kilobytes) and battery-operated.

\section{IEEE 802.15.4}
IEEE 802.15.4 standard specifies the physical layer and media access control for low-rate wireless personal area networks. The standard intends to offer the fundamental lower network layers of a type of wireless personal area network which focuses on low-cost, low-speed communication between tiny devices (motes). Such networks are typically limited to a personal operating space (POS) of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.

The IEEE 802.15.4 standard distinguishes between two types of nodes, reduced-function devices (RFDs) and full-function devices (FFDs).                                   FFDs typically have more resources and may be mains powered.  Therefor, FFDs aid RFDs by providing functions such as network coordination and packet forwarding.

An 802.15.4 network may operate in either the star or the peer-to-peer topology. In the star topology devices communicate with a single central PAN coordinator. In the peer-to-peer topology a device can communicate with any other device as long as they are in range of one another. Based on the peer-to-peer topology more complex network formations may be constructed, such as mesh networking topology. However, the standard does not define a network layer, therefor routing is not directly supported, but such an additional layer can add support for multihop communications. All devices operating on a network of either topology have unique 64 bit extended addresses. This address can be used for direct communication within the PAN, or it can be exchanged for a short 16 bit address allocated by the PAN coordinator.

The basic unit of data transport is a frame. The standard defines four frame structures: a beacon frame, used by a coordinator to transmit beacons;
a data frame, used for all transfers of data; an acknowledgment frame, used for confirming successful frame reception; a MAC command frame, used for handling all MAC control transfers. Additionally, a superframe structure may be defined by the coordinator. In such case two beacons act as superframe limits and provide synchronization to other devices. A superframe consists of sixteen equal-length slots, which can be further divided into an active part and an inactive part, during which the coordinator may enter power saving mode. Any device wishing to communicate during the contention access period (CAP) between two beacons shall compete with other devices using a slotted CSMA-CA mechanism. For applications requiring specific data bandwidth, the PAN coordinator may dedicate portions of the active superframe, which form the contention-free period (CFP).

An important aspect of 802.15.4 is its limitation on the frame size, which is specified by the frame length 7 bit field (0-127 bytes). Taking into account the frame header, which is up to 25 octets, this leaves 102 bytes for payload of the higher layers.

The standard also defines several security services such as maintaining an access control list (ACL) and
using symmetric-key cryptography to protect transmitted frames. However, key management is not specified and must be provided by the higher layers.

%%
%	Transmission of IPv6 Packets
%%
\chapter{IPv6 over LoWPAN Networks}
* why it's good (4919 Assumptions);\\
* problems;\\
%%
%	Conclusion
%%
\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}

%% Bibliography
\newpage
\bibliographystyle{plain}
\bibliography{msc-proposal}
\addcontentsline{toc}{chapter}{Bibliography}

\nocite{ieee802.15.4} 
\nocite{rfc2460} 
\nocite{rfc4291} 
\nocite{rfc4443}
\nocite{rfc4861}
\nocite{rfc4862}
\nocite{rfc4919}
\nocite{rfc4944} 
\nocite{draft-usecases-05} 
\nocite{draft-hc-06} 
\nocite{draft-nd-06}

\end{document}

### 2009-12-01 doc/msc-proposal/msc-proposal.tex
proposal: IP

@@
 
+\title{IPv6 over Low-Power Wireless Personal Area Networks\\ MSc. Thesis Proposal}
+\author{Siarhei Kuryla}
+
+%\department{Department of Electrical Engineering and Computer Science}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{titlepage}
-\includepdf[pages=1]{title}
-\thispagestyle{empty}
+
+\begin{center}
+
+\huge IPv6 over Low-Power Wireless Personal Area Networks\\ [0.5cm]
+
+\Large MSc. Thesis Proposal\\ [0.5cm]
+Siarhei Kuryla\\ [0.5cm]
+
+\large Supervisor: J\"urgen Shoenw\"alder\\[0.5cm]
+{\large \today}
+
+%\begin{minipage}{0.4\textwidth}
+%\begin{flushright} \large
+%\emph{Supervisor:} \\
+%Dr. Mark \textsc{Brown}
+%\end{flushright}
+%\end{minipage}
+%\vfill
+
+\end{center} 
 \end{titlepage}
@@
 \chapter{Introduction}
-* LowPAN, possible applications (use cases), why it's important.\\
+* There are a lot of applications where we need many small devices + examples; why it's important;
+* LoWPAN\\
 * The purpose of the proposed work will be to investigate smth. The proposal describes problems associated with enabling IP communication with devices in a LoWPAN.\\
@@
 
-\chapter{LoWPAN Networks}
-Low-power and lossy networks (LLNs) are the networks made of highly constrained nodes (limited CPU, memory, power) interconnected by a variety of "lossy" links (low-power radio links).
-Such networks can be characterized by low speed, low performance, low cost, and unstable connectivity.
-A low-power wireless personal area network (LoWPAN) is a particular instance of an LLN, comprised devices that conform to the IEEE 802.15.4 standard \cite{ieee802.15.4}.
+\chapter{IEEE 802.15.4}
+The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks.
+The standard intends to offer the fundamental lower network layers of a type of wireless personal area network which focuses on low-cost, low-speed communication between tiny devices (motes).
+Such networks are typically limited to a personal operating space (POS) of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.
+
+The IEEE 802.15.4 standard distinguishes between two types of nodes, reduced-function devices (RFDs) and full-function devices (FFDs).
+FFDs typically have more resources and may be mains powered.
+Therefor, FFDs aid RFDs by providing functions such as network coordination and packet forwarding.
+
+An 802.15.4 network may operate in either the star or the peer-to-peer topology.
+In the star topology devices communicate with a single central PAN coordinator.
+In the peer-to-peer topology a device can communicate with any other device as long as they are in range of one another.
+Based on the peer-to-peer topology more complex network formations may be constructed, such as mesh networking topology.
+However, the standard does not define a network layer, therefor routing is not directly supported, but such an additional layer can add support for multihop communications.
+All devices operating on a network of either topology have unique 64 bit extended addresses.
+This address can be used for direct communication within the PAN, or it can be exchanged for a short 16 bit address allocated by the PAN coordinator.
+
+The basic unit of data transport is a frame. The standard defines four frame structures: a beacon frame, used by a coordinator to transmit beacons;
+a data frame, used for all transfers of data; an acknowledgment frame, used for confirming successful frame reception; a MAC command frame, used for handling all MAC control transfers.
+Additionally, a superframe structure may be defined by the coordinator.
+In such case two beacons act as superframe limits and provide synchronization to other devices.
+A superframe consists of sixteen equal-length slots, which can be further divided into an active part and an inactive part, during which the coordinator may enter power saving mode.
+Any device wishing to communicate during the contention access period (CAP) between two beacons shall compete with other devices using a slotted CSMA-CA mechanism.
+For applications requiring specific data bandwidth, the PAN coordinator may dedicate portions of the active superframe, which form the contention-free period (CFP).
+
+An important aspect of 802.15.4 is its limitation on the frame size, which is specified by the frame length 7 bit field (0-127 bytes).
+Taking into account the frame header, which is up to 25 octets, this leaves 102 bytes for the payload of the higher layers.
+
+The standard also defines several security services such as maintaining an access control list (ACL) and
+using symmetric-key cryptography to protect transmitted frames. However, key management is not specified and must be provided by the higher layers.
 
@@
 %%
-\chapter{IPv6 over LoWPAN Networks}
+\chapter{IPv6 over IEEE 802.15.4}
+A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising typically low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
+The IEEE 802.15.4-2003 standard specifies node-to-node frame delivery between the devices which are within reachable distance from each other.
+On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be specified that would provide end-to-end packet delivery including routing through intermediate hosts.
+
+The Internet Protocol (IP) is the most widely deployed network layer protocol.
+Internet Protocol Version 4 (IPv4) is still the dominant protocol of the Internet, although the successor, Internet Protocol Version 6 (IPv6) \cite{rfc2460} is being deployed actively worldwide.
+The application of IP technologies to LoWPANs would provide a number of benefits.
+IP-based technologies are well-known and proven which allows to use the existing infrastructure.
+IP-based devices can be easily connected to other IP-based networks.
 

### 2009-12-01 trunk/doc/msc-proposal/msc-proposal.tex
proposal: adaptation layer

@@
 
-\chapter{LoWPAN Networks}
-A low-power wireless personal area network (LoWPAN) is a simple low cost communication network that allows wireless connectivity in applications with limited power and relaxed throughput requirements.
-LoWPAN networks comprises devices that conform to the IEEE 802.15.4 standard \cite{ieee802.15.4}.
-LoWPAN devices  are usually characterized as having limited processing capability (the smallest common LoWPAN nodes have 8-bit processors with clock rates around 10 MHz), small memory capacity (common RAM sizes for LoWPAN devices consist of a few kilobytes) and battery-operated.
-
-\section{IEEE 802.15.4}
-IEEE 802.15.4 standard specifies the physical layer and media access control for low-rate wireless personal area networks.
+\chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
+The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks.
 The standard intends to offer the fundamental lower network layers of a type of wireless personal area network which focuses on low-cost, low-speed communication between tiny devices (motes).
@@
 An 802.15.4 network may operate in either the star or the peer-to-peer topology.
-In the star topology devices communicate with a single central PAN coordinator.
+In the star topology devices communicate with a single central                           a personal area network (PAN) coordinator.
 In the peer-to-peer topology a device can communicate with any other device as long as they are in range of one another.
@@
 An important aspect of 802.15.4 is its limitation on the frame size, which is specified by the frame length 7 bit field (0-127 bytes).
-Taking into account the frame header, which is up to 25 octets, this leaves 102 bytes for payload of the higher layers.
+Taking into account the frame header, which is up to 25 octets, this leaves 102 bytes for the payload of the higher layers.
 
@@
 %%
-\chapter{IPv6 over LoWPAN Networks}
-* why it's good (4919 Assumptions);\\
-* problems;\\
+\chapter{6LoWPAN Networks}
+A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
+The IEEE 802.15.4-2003 standard specifies node-to-node frame delivery between the devices which are within reachable distance from each other.
+On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be defined that would provide end-to-end packet delivery including routing through intermediate hosts.
+
+%The Internet Protocol (IP) is the most widely deployed network layer protocol.
+Internet Protocol Version 4 (IPv4) is still the dominant protocol of the Internet, although the successor, Internet Protocol Version 6 (IPv6) \cite{rfc2460} is being deployed actively worldwide.
+The application of IP technologies to LoWPANs would provide a number of benefits.
+IP-based technologies are well-known and proven which allows to use the existing infrastructure.
+IP-based devices can be easily connected to other IP-based networks.
+Today there exist a lot of ready-to-use tools for diagnostics and management of IP networks.
+
+6lowpan (IPv6 over Low power Wireless Personal Area Networks) is a working group within the IETF concerned with the specification of mechanisms to allow IPv6 packets to be sent to and received from over IEEE 802.15.4 based networks.
+The working group has already completed two documents: \cite{rfc4919} that documents and discusses the problem statement and \cite{rfc4944} that defines the format for the adaptation between IPv6 and 802.15.4.
+The latter document describes the frame format for transmission of IPv6 packets and the method of forming IPv6 link-local addresses and statelessly autoconfigured addresses on IEEE 802.15.4 networks.
+
+As described in chapter \ref{ch:ieee802.15.4}, IEEE 802.15.4 defines four types of frames: beacon frames, MAC command frames, acknowledgement frames, and data frames.
+IPv6 packets are carried on data frames.
+Data frames may optionally request acknowledgments to aid link-layer recovery.
+
+\section{Addressing}
+IEEE 802.15.4 defines addresses of two types: IEEE 64-bit extended addresses or 16-bit short addresses unique within the PAN. Both types are supported by 6LoWPAN.
+
+6LoWPAN supports stateless address autoconfiguration, which allows to obtain the Interface Identifier \cite{rfc4291} for an IEEE 802.15.4 interface  based on the EUI-64 identifier \cite{eui64} \cite{rfc2464} assigned to the IEEE802.15.4 device.
+Even though all 802.15.4 devices have an EUI-64 address, 16-bit short addresses also can be used for address autoconfiguration.
+In this case, a pseudo 48-bit address is formed by concatenating 16 zero bits to the 16-bit PAN ID, the resulting 32 bits are concatenated with the 16-bit short address and the interface identifier is formed from this 48-bit address:
+\begin{center}\texttt{16\_bit\_PAN\_ID:16\_zero\_bits:16\_bit\_short\_address}\end{center}
+
+The IPv6 link-local address for an IEEE 802.15.4 interface is formed by appending the Interface Identifier to the prefix \texttt{FE80::/64}.
+
+Packets with a multicast IPv6 destination address are sent to the 16-bit 802.15.4 address obtained by concatenating the 3-bit multicast prefix 101, the last 5 bits in the 15-th octet and the whole 16-th octet of the multicast IPv6 address.
+
+\section{Adaptation Layer}
+The IPv6 minimum Maximum Transmission Unit (MTU) size is defined as 1280 octets.
+As described in chapter \ref{ch:ieee802.15.4}, the maximum IEEE 802.15.4 frame size is limited to 127 octets.
+Taking into account the maximum frame overhead of 25 octets, 102 octets are left at the media access control layer.
+Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size.
+Therefor, a fragmention and reassembly adaptation layer is provided at the layer below IP.
+
+Furthermore, the IPv6 header is 40 octets long which leaves only 41 octets for upper-layer protocols, like UDP.
+Since the latter uses 8 octets in the header, application data size is limited to 33 octets.
+
+\subsection{Fragmentation Header}
+\chapter{6LoWPAN Neighbor Discovery}
+
 %%
@@
 \nocite{ieee802.15.4} 
+\nocite{eui64} 
 \nocite{rfc2460}
+\nocite{rfc2464}
 \nocite{rfc4291} 

### 2009-12-01 trunk/doc/msc-proposal/msc-proposal.tex
proposal: adaptation layer

@@
 
-As described in chapter \ref{ch:ieee802.15.4}, IEEE 802.15.4 defines four types of frames: beacon frames, MAC command frames, acknowledgement frames, and data frames.
+As described in Chapter \ref{ch:ieee802.15.4}, IEEE 802.15.4 defines four types of frames: beacon frames, MAC command frames, acknowledgement frames, and data frames.
 IPv6 packets are carried on data frames.
@@
 
-\section{Adaptation Layer}
+\section{Adaptation layer}\label{sec:adapt.layer}
 The IPv6 minimum Maximum Transmission Unit (MTU) size is defined as 1280 octets.
-As described in chapter \ref{ch:ieee802.15.4}, the maximum IEEE 802.15.4 frame size is limited to 127 octets.
+As described in Chapter \ref{ch:ieee802.15.4}, the maximum IEEE 802.15.4 frame size is limited to 127 octets.
 Taking into account the maximum frame overhead of 25 octets, 102 octets are left at the media access control layer.
 Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size.
-Therefor, a fragmention and reassembly adaptation layer is provided at the layer below IP.
+Therefor, a fragmentation and reassembly adaptation layer is provided at the layer below IP.
 
-Furthermore, the IPv6 header is 40 octets long which leaves only 41 octets for upper-layer protocols, like UDP.
-Since the latter uses 8 octets in the header, application data size is limited to 33 octets.
+IPv6 datagrams transported over IEEE 802.15.4 are prefixed by an encapsulation header stack.
+Each header in the header stack contains a header type followed by zero or more header fields.
+Three optional encapsulation headers may appear in the following order: Mesh Addressing Header, Broadcast Header and Fragmentation Header.
+These headers are described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header} and \ref{subsec:broad.header} respectively.
+The Dispatch
 
-\subsection{Fragmentation Header}
+\subsection{Fragmentation Header}\label{subsec:frag.header}
+\subsection{Mesh addressing Header}\label{subsec:mesh.header}
+\subsection{Broadcast Header}\label{subsec:broad.header}
+
+\section{Header compression}
+As described in Section \ref{sec:adapt.layer}, 81 octets are left in a IEEE 802.15.4 frame for IPv6.
+The IPv6 header is 40 octets long which leaves only 41 octets for upper-layer protocols, like UDP.
+The latter uses 8 octets in the header, which leaves 33 octets for application data.
+The fragmentation and reassembly layer, described in Section \ref{sec:adapt.layer}, will also use at least one additional octet, what makes header compression almost being unavoidable.
+
 \chapter{6LoWPAN Neighbor Discovery}

### 2009-12-01 trunk/doc/msc-proposal/msc-proposal.tex
propsal: adaptation layer

@@
 
-%\department{Department of Electrical Engineering and Computer Science}
+\newenvironment{mylisting}
+{\begin{list}{}{\setlength{\leftmargin}{1em}}\item\footnotesize}
+{\end{list}}
+
+\newenvironment{mytinylisting}
+{\begin{list}{}{\setlength{\leftmargin}{1em}}\item\tiny\bfseries}
+{\end{list}}
 
@@
 \tableofcontents
-%\listoffigures 
+\listoffigures 
 \newpage
@@
 Taking into account the maximum frame overhead of 25 octets, 102 octets are left at the media access control layer.
-Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size.
+Link-layer security imposes further overhead, which in the maximfum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size.
 Therefor, a fragmentation and reassembly adaptation layer is provided at the layer below IP.
@@
 Each header in the header stack contains a header type followed by zero or more header fields.
-Three optional encapsulation headers may appear in the following order: Mesh Addressing Header, Broadcast Header and Fragmentation Header.
-These headers are described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header} and \ref{subsec:broad.header} respectively.
-The Dispatch
+There are four encapsulation headers defined: Mesh Addressing Header, Broadcast Header and Fragmentation Header (described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header},\ref{subsec:broad.header} and \ref{subsec:dispatch.header} respectively).
+The headers must appear in the specified order.
+First three of them are optional, whereas the last one is required.
 
+\subsection{Dispatch Header}\label{subsec:dispatch.header}
+The dispatch type is defined as a zero bit as the first bit followed by a one bit.
+
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                     1                   2                   3
+   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |0 1| Dispatch  |  type-specific header
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Dispatch Type and Header}\label{fig:dispatch.header}
+\end{figure}
+The Dispatch value identifies the type of header immediately following the Dispatch Header.
+
+\subsection{Mesh addressing Header}\label{subsec:mesh.header}
 \subsection{Fragmentation Header}\label{subsec:frag.header}
-\subsection{Mesh addressing Header}\label{subsec:mesh.header}
 \subsection{Broadcast Header}\label{subsec:broad.header}

### 2009-12-02 trunk/doc/msc-proposal/msc-proposal.tex
proposal: finished the adaptation layer part

@@
 Taking into account the maximum frame overhead of 25 octets, 102 octets are left at the media access control layer.
-Link-layer security imposes further overhead, which in the maximfum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size.
+Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size.
 Therefor, a fragmentation and reassembly adaptation layer is provided at the layer below IP.
+
+Even though 802.15.4 networks are expected to commonly use mesh routing, the IEEE 802.15.4-2003 \cite{ieee802.15.4} specification does not define such capability.
+Mesh routing allows two devices  without direct reachability to communicate with each other.
+In such mesh scenarios, intermediate devices are used as forwarders towards the final destination.
+In order to support mesh routing, in addition to the hop-by-hop source and destination link-layer addresses the originator and final destination addresses need to be known.
+For this purpose a Mesh Addressing header is introduced.
 
@@
 Each header in the header stack contains a header type followed by zero or more header fields.
-There are four encapsulation headers defined: Mesh Addressing Header, Broadcast Header and Fragmentation Header (described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header},\ref{subsec:broad.header} and \ref{subsec:dispatch.header} respectively).
-The headers must appear in the specified order.
+There are four encapsulation headers defined: Mesh Addressing Header, Broadcast Header, Fragmentation Header and Dispatch Header (described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header},\ref{subsec:broad.header} and \ref{subsec:dispatch.header} respectively).
+These headers must appear in the specified order.
 First three of them are optional, whereas the last one is required.
@@
 \subsection{Dispatch Header}\label{subsec:dispatch.header}
-The dispatch type is defined as a zero bit as the first bit followed by a one bit.
+The dispatch type is defined by a zero bit as the first bit followed by a one bit. The dispatch header is shown in Figure \ref{fig:dispatch.header}.
 
@@
 \end{figure}
-The Dispatch value identifies the type of header immediately following the Dispatch Header.
+The Dispatch value identifies the type of payload immediately following the Dispatch Header.
+This can  be, for example, a compressed or uncompressed IPv6 header.
+More details can be found in \cite{rfc4944}.
 
 \subsection{Mesh addressing Header}\label{subsec:mesh.header}
+
+The mesh header type is defined by a one bit and zero bit as the first two bits. The mesh header is shown in Figure \ref{fig:mesh.header}:
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                       1                   2                   3
+   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |1 0|V|F|HopsLft| originator address, final address
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Mesh Addressing Type and Header}\label{fig:mesh.header}
+\end{figure}
+
+The \texttt{V} and \texttt{F} bits indicate whether an IEEE extended 64-bit address (EUI-64) or a short 16-bit addresses are used for the originator and the final destination addresses respectively.
+The \texttt{HopsLft} field is used as a hop limit and decremented by every forwarding node.
+The left part of the mesh header specifies the originator and final destination link-layer addresses.
+
 \subsection{Fragmentation Header}\label{subsec:frag.header}
+As previously discussed, an entire IPv6 datagram may not fit within a single 802.15.4 frame.
+In such case it is broken into link fragments.
+The first link fragment contains the first fragment header shown in Figure \ref{fig:first.fragment}.
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                      1                   2                   3
+  0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |1 1 0 0 0|    datagram_size    |         datagram_tag          |
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{First Fragment}\label{fig:first.fragment}
+\end{figure}
+
+The second and subsequent link fragments contains a fragmentation header that conforms to the format shown in Figure \ref{fig:sub.fragment}.
+
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                     1                   2                   3
+   1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |1 1 1 0 0|    datagram_size    |         datagram_tag          |
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |datagram_offset|
+  +-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Subsequent Fragments}\label{fig:sub.fragment}
+\end{figure}
+
+The first and subsequent fragment header types are specified by the binary sequences \texttt{11000} and \texttt{11100} respectively.
+The  \texttt{datagram\_size} 11-bit field encodes the size of the entire IP packet before link-layer fragmentation.
+The \texttt{datagram\_tag} field is the same for all link fragments of a datagram and incremented by the sender for successive fragmented datagrams.
+The \texttt{datagram\_offset} field is present only in the second and subsequent link fragments and specifies the offset of the fragment from the beginning of the payload datagram.
+
 \subsection{Broadcast Header}\label{subsec:broad.header}
+Broadcast is an additional mesh routing functionality. For that a broadcast header is defined the format of which is shown in Figure \ref{fig:broad.header}.
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                       1
+   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |0|1|0|1|0|0|0|0|Sequence Number|
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Broadcast Header
+}\label{fig:broad.header}
+\end{figure}
+
+The broadcast header consists of a dispatch type \texttt{01010000} followed by a sequence number field, which is used to detect duplicate packets.
+This field is incremented by the originator whenever it sends a new mesh broadcast or multicast packet.
 

### 2009-12-07 doc/msc-proposal/msc-proposal.tex
proposal: header compression

@@
 * LoWPAN\\
-* The purpose of the proposed work will be to investigate smth. The proposal describes problems associated with enabling IP communication with devices in a LoWPAN.\\
+* The purpose of the proposed work will be to investigate something. The proposal describes problems associated with enabling IP communication with devices in a LoWPAN.\\
 * As preparation for this work, several related areas have been studied in more
@@
 
-\chapter{IEEE 802.15.4}
+\chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
 The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks.
@@
 %%
-\chapter{IPv6 over IEEE 802.15.4}
-A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising typically low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
+\chapter{6LoWPAN Networks}
+A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
 The IEEE 802.15.4-2003 standard specifies node-to-node frame delivery between the devices which are within reachable distance from each other.
-On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be specified that would provide end-to-end packet delivery including routing through intermediate hosts.
+On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be defined that would provide end-to-end packet delivery including routing through intermediate hosts.
 
-The Internet Protocol (IP) is the most widely deployed network layer protocol.
+%The Internet Protocol (IP) is the most widely deployed network layer protocol.
 Internet Protocol Version 4 (IPv4) is still the dominant protocol of the Internet, although the successor, Internet Protocol Version 6 (IPv6) \cite{rfc2460} is being deployed actively worldwide.
@@
 IP-based devices can be easily connected to other IP-based networks.
+Today there exist a lot of ready-to-use tools for diagnostics and management of IP networks.
+
+6lowpan (IPv6 over Low power Wireless Personal Area Networks) is a working group within the IETF concerned with the specification of mechanisms to allow IPv6 packets to be sent to and received from over IEEE 802.15.4 based networks.
+The working group has already completed two documents: \cite{rfc4919} that documents and discusses the problem statement and \cite{rfc4944} that defines the format for the adaptation between IPv6 and 802.15.4.
+The latter document describes the frame format for transmission of IPv6 packets and the method of forming IPv6 link-local addresses and statelessly autoconfigured addresses on IEEE 802.15.4 networks.
+
+As described in Chapter \ref{ch:ieee802.15.4}, IEEE 802.15.4 defines four types of frames: beacon frames, MAC command frames, acknowledgement frames, and data frames.
+IPv6 packets are carried on data frames.
+Data frames may optionally request acknowledgments to aid link-layer recovery.
+
+\section{Addressing}
+IEEE 802.15.4 defines addresses of two types: IEEE 64-bit extended addresses or 16-bit short addresses unique within the PAN. Both types are supported by 6LoWPAN.
+
+6LoWPAN supports stateless address autoconfiguration, which allows to obtain the Interface Identifier \cite{rfc4291} for an IEEE 802.15.4 interface  based on the EUI-64 identifier \cite{eui64} \cite{rfc2464} assigned to the IEEE802.15.4 device.
+Even though all 802.15.4 devices have an EUI-64 address, 16-bit short addresses also can be used for address autoconfiguration.
+In this case, a pseudo 48-bit address is formed by concatenating 16 zero bits to the 16-bit PAN ID, the resulting 32 bits are concatenated with the 16-bit short address and the interface identifier is formed from this 48-bit address:
+\begin{center}\texttt{16\_bit\_PAN\_ID:16\_zero\_bits:16\_bit\_short\_address}\end{center}
+
+The IPv6 link-local address for an IEEE 802.15.4 interface is formed by appending the Interface Identifier to the prefix \texttt{FE80::/64}.
+
+Packets with a multicast IPv6 destination address are sent to the 16-bit 802.15.4 address obtained by concatenating the 3-bit multicast prefix 101, the last 5 bits in the 15-th octet and the whole 16-th octet of the multicast IPv6 address.
+
+\section{Adaptation Layer}\label{sec:adapt.layer}
+The IPv6 minimum Maximum Transmission Unit (MTU) size is defined as 1280 octets.
+As described in Chapter \ref{ch:ieee802.15.4}, the maximum IEEE 802.15.4 frame size is limited to 127 octets.
+Taking into account the maximum frame overhead of 25 octets, 102 octets are left at the media access control layer.
+Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size.
+Therefor, a fragmentation and reassembly adaptation layer is provided at the layer below IP.
+
+Even though 802.15.4 networks are expected to commonly use mesh routing, the IEEE 802.15.4-2003 \cite{ieee802.15.4} specification does not define such capability.
+Mesh routing allows two devices  without direct reachability to communicate with each other.
+In such mesh scenarios, intermediate devices are used as forwarders towards the final destination.
+In order to support mesh routing, in addition to the hop-by-hop source and destination link-layer addresses the originator and final destination addresses need to be known.
+For this purpose a Mesh Addressing header is introduced.
+
+IPv6 datagrams transported over IEEE 802.15.4 are prefixed by an encapsulation header stack.
+Each header in the header stack contains a header type followed by zero or more header fields.
+There are four encapsulation headers defined: Mesh Addressing Header, Broadcast Header, Fragmentation Header and Dispatch Header (described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header},\ref{subsec:broad.header} and \ref{subsec:dispatch.header} respectively).
+These headers must appear in the specified order.
+First three of them are optional, whereas the last one is required.
+
+\subsection{Dispatch Header}\label{subsec:dispatch.header}
+The dispatch type is defined by a zero bit as the first bit followed by a one bit. The dispatch header is shown in Figure \ref{fig:dispatch.header}.
+
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                     1                   2                   3
+   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |0 1| Dispatch  |  type-specific header
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Dispatch Type and Header}\label{fig:dispatch.header}
+\end{figure}
+The Dispatch value identifies the type of payload immediately following the Dispatch Header.
+This can  be, for example, a compressed or uncompressed IPv6 header.
+More details can be found in \cite{rfc4944}.
+
+\subsection{Mesh addressing Header}\label{subsec:mesh.header}
+
+The mesh header type is defined by a one bit and zero bit as the first two bits. The mesh header is shown in Figure \ref{fig:mesh.header}:
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                       1                   2                   3
+   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |1 0|V|F|HopsLft| originator address, final address
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Mesh Addressing Type and Header}\label{fig:mesh.header}
+\end{figure}
+
+The \texttt{V} and \texttt{F} bits indicate whether an IEEE extended 64-bit address (EUI-64) or a short 16-bit addresses are used for the originator and the final destination addresses respectively.
+The \texttt{HopsLft} field is used as a hop limit and decremented by every forwarding node.
+The left part of the mesh header specifies the originator and final destination link-layer addresses.
+
+\subsection{Fragmentation Header}\label{subsec:frag.header}
+As previously discussed, an entire IPv6 datagram may not fit within a single 802.15.4 frame.
+In such case it is broken into link fragments.
+The first link fragment contains the first fragment header shown in Figure \ref{fig:first.fragment}.
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                      1                   2                   3
+  0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |1 1 0 0 0|    datagram_size    |         datagram_tag          |
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{First Fragment}\label{fig:first.fragment}
+\end{figure}
+
+The second and subsequent link fragments contains a fragmentation header that conforms to the format shown in Figure \ref{fig:sub.fragment}.
+
+\begin{figure}[htp]
+\centering
+\begin{mylisting}
+\begin{verbatim}
+                     1                   2                   3
+   1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |1 1 1 0 0|    datagram_size    |         datagram_tag          |
+  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+  |datagram_offset|
+  +-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Subsequent Fragments}\label{fig:sub.fragment}
+\end{figure}
+
+The first and subsequent fragment header types are specified by the binary sequences \texttt{11000} and \texttt{11100} respectively.
+The  \texttt{datagram\_size} 11-bit field encodes the size of the entire IP packet before link-layer fragmentation.
+The \texttt{datagram\_tag} field is the same for all link fragments of a datagram and incremented by the sender for successive fragmented datagrams.
+The \texttt{datagram\_offset} field is present only in the second and subsequent link fragments and specifies the offset of the fragment from the beginning of the payload datagram.
+
+\subsection{Broadcast Header}\label{subsec:broad.header}
+Broadcast is an additional mesh routing functionality. For that a broadcast header is defined the format of which is shown in Figure \ref{fig:broad.header}.
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                                       1
+                   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5
+                  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+                  |0|1|0|1|0|0|0|0|Sequence Number|
+                  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Broadcast Header
+}\label{fig:broad.header}
+\end{figure}
+
+The broadcast header consists of a dispatch type \texttt{01010000} followed by a sequence number field, which is used to detect duplicate packets.
+This field is incremented by the originator whenever it sends a new mesh broadcast or multicast packet.
+
+\section{Header Compression}
+As described in Section \ref{sec:adapt.layer}, 81 octets are left in a IEEE 802.15.4 frame for IPv6.
+The IPv6 header is 40 octets long which leaves only 41 octets for upper-layer protocols, like UDP.
+The latter uses 8 octets in the header, which leaves 33 octets for application data.
+The fragmentation and reassembly layer, described in Section \ref{sec:adapt.layer}, will also use at least one additional octet, what makes header compression almost being unavoidable.
+
+RFC 4944 \cite{rfc4944} defines stateless header compression mechanisms for IPv6 datagrams (LOWPAN\_HC1 and LOWPAN\_HC2) to reduce the relatively large IPv6 and UDP headers.
+However, LOWPAN\_HC1 and LOWPAN\_HC2 are insufficient for most practical uses of 6LoWPAN networks, since they are effective only for link-local unicast communication, where IPv6 addresses derived directly from IEEE 802.15.4 addresses.
+When communicating with devices external to the LoWPAN or in a route-over configuration where IP forwarding occurs within the LoWPAN, LOWPAN\_HC1 carries both IPv6 source and destination addresses in-line.
+Internet Draft document \cite{draft-hc-06} defines an encoding format, LOWPAN\_IPHC, for effective compression of unique local, global, and multicast IPv6 addresses based on shared contexts.
+In addition, it also defines an encoding format, LOWPAN\_NHC, for arbitrary next header compression.
+
+\subsection{HC1 Header Encoding}
+HC1 compression format does not use any shared context and compresses headers relying on information pertaining to the entire link.
+
+\subsection{HC2 Header Encoding}
+
+\subsection{IPHC Header Encoding}
+
+\subsection{NHC Header Encoding}
+
+\chapter{6LoWPAN Neighbor Discovery}
 

### 2009-12-07 doc/msc-proposal/msc-proposal.tex
proposal: hc1

@@
 \end{figure}
-The Dispatch value identifies the type of payload immediately following the Dispatch Header.
+The dispatch value identifies the type of payload immediately following the Dispatch Header.
 This can  be, for example, a compressed or uncompressed IPv6 header.
@@
 \subsection{HC1 Header Encoding}
-HC1 compression format does not use any shared context and compresses headers relying on information pertaining to the entire link.
+The HC1 compression format does not use any shared context and compresses headers relying on information pertaining to the entire link.
+The following IPv6 header values are expected to be common on 6LoWPAN network: Version is IPv6; both IPv6 source and destination addresses are link local;  the IPv6 interface identifiers for the source or destination addresses can be inferred from the layer two source and destination addresses; the packet length can be inferred either from layer two or from the "datagram\_size" field in the fragment header; both the Traffic Class and the Flow Label are zero; and the Next Header is UDP, ICMP or TCP.
+The Hop Limit (8 bits) field is the only one that always needs to be carried.
+Using the HC1 encoding allows to compress a 40 byte IPv6 header to 2 bytes.
+
+The HC1 encoding uses a dispatch value of $000010$ to specify that the payload following the dispatch header, described in Section \ref{subsec:dispatch.header}, is a HC1 compressed IPv6 header.
+The HC1 header is 8 bits.
+Bits 0 and 1 specify the encoding for the IPv6 source prefix and interface identifier respectively: $0$ -- the field is carried in-line; $1$ -- link-local address is assumed.
+Bits 2 and 3 specify how the destination address is encoded (the same encoding as for the source address).
+If bit 4 is set then  Traffic Class and Flow Label fields are zero, otherwise their values are not compressed and carried inline.
+Bits 5 and 6 indicate the Next Header: $00$ -- the field is not compressed and carried inlie, $01$ corresponds to UDP, $10$ -- ICMP, $11$ -- TCP.
+The last bit specifies whether HC1 header is immediately followed by more header compression bits of the HC2 encoding format.
+
+By using this encoding an IPv6 header can be compressed to different degrees and only non-compressed fields need to be sent.
+The only field that is always present is the Hop Limit (8 bits).
+It follows the HC1 encoding fields.
+Other non-compressed fields follow the Hop Limit.
 

### 2009-12-07 trunk/doc/msc-proposal/msc-proposal.tex
proposal: HC2

@@
 
-\section{Adaptation layer}\label{sec:adapt.layer}
+\section{Adaptation Layer}\label{sec:adapt.layer}
 The IPv6 minimum Maximum Transmission Unit (MTU) size is defined as 1280 octets.
@@
 \begin{figure}[htp]
-\centering
 \begin{mylisting}
@@
 
-\section{Header compression}
+\section{Header Compression}
 As described in Section \ref{sec:adapt.layer}, 81 octets are left in a IEEE 802.15.4 frame for IPv6.
@@
 The fragmentation and reassembly layer, described in Section \ref{sec:adapt.layer}, will also use at least one additional octet, what makes header compression almost being unavoidable.
+
+RFC 4944 \cite{rfc4944} defines stateless header compression mechanisms for IPv6 datagrams (LOWPAN\_HC1 and LOWPAN\_HC2) to reduce the relatively large IPv6 and UDP headers.
+However, LOWPAN\_HC1 and LOWPAN\_HC2 are insufficient for most practical uses of 6LoWPAN networks, since they are effective only for link-local unicast communication, where IPv6 addresses derived directly from IEEE 802.15.4 addresses.
+When communicating with devices external to the LoWPAN or in a route-over configuration where IP forwarding occurs within the LoWPAN, LOWPAN\_HC1 carries both IPv6 source and destination addresses in-line.
+Internet Draft document \cite{draft-hc-06} defines an encoding format, LOWPAN\_IPHC, for effective compression of unique local, global, and multicast IPv6 addresses based on shared contexts.
+In addition, it also defines an encoding format, LOWPAN\_NHC, for arbitrary next header compression.
+
+\subsection{HC1 Header Encoding}
+The HC1 compression format does not use any shared context and compresses headers relying on information pertaining to the entire link.
+The following IPv6 header values are expected to be common on 6LoWPAN network: Version is IPv6; both IPv6 source and destination addresses are link local;  the IPv6 interface identifiers for the source or destination addresses can be inferred from the layer two source and destination addresses; the packet length can be inferred either from layer two or from the "datagram\_size" field in the fragment header; both the Traffic Class and the Flow Label are zero; and the Next Header is UDP, ICMP or TCP.
+The Hop Limit (8 bits) field is the only one that always needs to be carried.
+Using the HC1 encoding allows to compress a 40 byte IPv6 header to 2 bytes.
+
+The HC1 encoding uses a dispatch value of $000010$ to specify that the payload following the dispatch header, described in Section \ref{subsec:dispatch.header}, is a HC1 compressed IPv6 header.
+The HC1 header is 8 bits.
+Bits 0 and 1 specify the encoding for the IPv6 source prefix and interface identifier respectively: $0$ -- the field is carried in-line; $1$ -- link-local address is assumed.
+Bits 2 and 3 specify how the destination address is encoded (the same encoding as for the source address).
+If bit 4 is set then  Traffic Class and Flow Label fields are zero, otherwise their values are not compressed and carried in-line.
+Bits 5 and 6 indicate the Next Header: $00$ -- the field is not compressed and carried inlie, $01$ corresponds to UDP, $10$ -- ICMP, $11$ -- TCP.
+The last bit specifies whether HC1 header is immediately followed by more header compression bits of the HC2 encoding format.
+
+By using this encoding an IPv6 header can be compressed to different degrees and only non-compressed fields need to be sent.
+The only field that is always present is the Hop Limit (8 bits).
+It follows the HC1 encoding fields.
+Other non-compressed fields follow the Hop Limit.
+
+\subsection{HC\_UDP Header Encoding}
+The HC\_UDP encoding is used for compression of a UDP header.
+It is applied only if bits $5$ and $6$ in HC1 specify that the protocol that follows the IPv6 header is UDP and bit 7 indicates that the HC2 encoding is used.
+The HC\_UDP encoding allows to compress the UDP header to 4 octets instead of the original 8 octets.
+
+The first $8$ bits of the HC\_UDP encoding specify how the UDP header is compressed.
+Both the source and destination port fields may be compressed to $4$ bits.
+It is indicated by bits $0$ and $1$ of the HC\_UDP encoding for the source and destination ports respectively.
+In such case, the actual $16$-bit port is obtained by calculating $61616$ + short port value, which is a $4$ bit value carried in-line.
+If bit $2$ of the HC\_UDP encoding is set then the UDP length header field is computed from IPv6 header length information, otherwise it's carried in-line.
+HC\_UDP encoding bits 3-7 are reserved for future use.
+The UDP header's checksum field is not compressed and is therefore carried in full.
+
+The non-compressed or partially compressed fields in the UDP header always follow the IPv6 header.
+All present UDP header in-line fields always appear in the same order as the corresponding fields appear in a normal UDP header \cite{rfc780}, e.g., source port, destination port, length, and checksum.
+
+\subsection{IPHC Header Encoding}
+
+\subsection{NHC Header Encoding}
 
@@
 
-%% Bibliography
+% Bibliography
 \newpage
@@
 \nocite{eui64} 
+\nocite{rfc780}
 \nocite{rfc2460}

### 2009-12-08 trunk/doc/msc-proposal/msc-proposal.tex
proposal: IPHC

@@
 \subsection{IPHC Header Encoding}
+The IPHC encoding format enables effective compression of the IPv6 header relying on information pertaining to the entire 6LoWPAN network.
+The encoding assumes the following common case parameters for 6LoWPAN communication: Version is $6$; Traffic Class and Flow Label are both zero; Payload Length can be inferred from lower layers from either the 6LoWPAN Fragmentation header or the IEEE 802.15.4 header; Hop Limit will be set to a well-known value by the source; addresses assigned to 6LoWPAN interfaces will be formed using the link-local prefix or a single routable prefix assigned to the entire 6LoWPAN network; addresses assigned to 6LoWPAN interfaces are formed with an IID derived directly from either the 64-bit extended or 16-bit short IEEE 802.15.4 addresses.
+The compression mechanism is adapted for these values of the header fields and in such scenario the LOWPAN IPHC can compress the IPv6 header down to two octets (1-octet dispatch, 1-octet LOWPAN\_IPHC encoding) with link-local communication.
+When routing over multiple IP hops, IPHC can compress the IPv6 header down to 7 octets (1-octet dispatch, 1-octet IPHC encoding, 1-octet Hop Limit, 2-octet Source Address, and 2-octet Destination Address).
+In this case, the Hop Limit may not be compressed because it needs to be decremented at each hop and stateful address compression is applied to the source and destination IPv6 addresses because they do not statelessly match the source and destination link layer addresses on intermediate hops.
+
+The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$, a 13 bit IPHC encoding, describing how the IPv6 header is compressed, followed by compressed IPv6 header fields.
+The IPHC encoding may be extended by another octet to support additional contexts.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
++------------------+---------------+------------------------+
+|    Dispatch      | IPHC encoding | Compressed IPv6 Header |
+| (3 bits - 110)   |   (13 bits)   |                        |
++------------------+---------------+------------------------+
+\end{verbatim}
+\end{mylisting}
+\caption{IPHC Header}\label{fig:iphc.header}
+\end{figure}
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+  0   1   2   3   4   5   6   7   8   9   0   1   2   3   4   5
++---+---+---+---+---+---+---+---+---+---+---+---+---+---+---+---+
+| 0 | 1 | 1 |  TF   |NH | HLIM  |CID|SAC|  SAM  | M |DAC|  DAM  |
++---+---+---+---+---+---+---+---+---+---+---+---+---+---+---+---+
+\end{verbatim}
+\end{mylisting}
+\caption{IPHC Encoding}\label{fig:iphc.enc}
+\end{figure}
 

### 2009-12-08 trunk/doc/msc-proposal/msc-proposal.tex
proposal: IPHC

@@
 If bit 4 is set then  Traffic Class and Flow Label fields are zero, otherwise their values are not compressed and carried in-line.
-Bits 5 and 6 indicate the Next Header: $00$ -- the field is not compressed and carried inlie, $01$ corresponds to UDP, $10$ -- ICMP, $11$ -- TCP.
+Bits 5 and 6 indicate the Next Header: $00$ -- the field is not compressed and carried in-line, $01$ corresponds to UDP, $10$ -- ICMP, $11$ -- TCP.
 The last bit specifies whether HC1 header is immediately followed by more header compression bits of the HC2 encoding format.
@@
 
-The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$, a 13 bit IPHC encoding, describing how the IPv6 header is compressed, followed by compressed IPv6 header fields.
+The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$, a 13 bit IPHC base encoding (shown in Figure \ref{fig:iphc.format}), describing how the IPv6 header is compressed, followed by compressed IPv6 header fields.
 The IPHC encoding may be extended by another octet to support additional contexts.
@@
     |    Dispatch      | IPHC encoding | Compressed IPv6 Header |
-| (3 bits - 110)   |   (13 bits)   |                        |
+    | (3 bits - 110)   |   (13 bits)   |      Fields            |
     +------------------+---------------+------------------------+
@@
 \begin{verbatim}
-  0   1   2   3   4   5   6   7   8   9   0   1   2   3   4   5
-+---+---+---+---+---+---+---+---+---+---+---+---+---+---+---+---+
-| 0 | 1 | 1 |  TF   |NH | HLIM  |CID|SAC|  SAM  | M |DAC|  DAM  |
-+---+---+---+---+---+---+---+---+---+---+---+---+---+---+---+---+
+       0   1   2   3   4   5   6   7   8   9   0   1   2  
+     +---+---+---+---+---+---+---+---+---+---+---+---+---+
+     |  TF   |NH | HLIM  |CID|SAC|  SAM  | M |DAC|  DAM  |
+     +---+---+---+---+---+---+---+---+---+---+---+---+---+
 \end{verbatim}
 \end{mylisting}
-\caption{IPHC Encoding}\label{fig:iphc.enc}
+\caption{IPHC Base Encoding}\label{fig:iphc.format}
 \end{figure}
 
-\subsection{NHC Header Encoding}
+\subsubsection{Traffic Class and  Flow Label Compression}
+The IPHC supports four compression formats of the Traffic Class and Flow Label (TF).
+The two bit TF field in the IPHC Base Encoding, shown in Figure \ref{fig:iphc.format}, indicates whether these fields are carried in-line in the compressed IPv6 header.
+The Traffic Class field in the IPv6 header comprises 6 bits of Differentiated Services Code Point (DSCP) Field \cite{rfc2474} and 2 bits of Explicit Congestion Notification (ECN) \cite{rfc3168}.
+
+The TF field value $00$ indicates that the Traffic Class and Flow Label are carried in-line with additional 4 bits included to maintain byte-alignment.
+In this case, shown in Figure \ref{fig:tf.00}, Traffic Class and  Flow Label fields are represented by 4 octets in the IPv6 header.
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                        1                   2                   3
+    0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |ECN|   DSCP    |  rsv  |             Flow Label                |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{ TF = 00: Traffic Class and Flow Label carried in-line.}\label{fig:tf.00}
+\end{figure}
+
+The TF value $01$ corresponds to the 3 byte encoding, shown in Figure \ref{fig:tf.01}, when ECN and Flow Label fields are carried in-line, but DSCP field is elided.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                          1                   2
+      0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3
+     +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+     |ECN|rsv|             Flow Label                |
+     +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{TF = 01: Flow Label carried in-line.}\label{fig:tf.01}
+\end{figure}
+
+When TF is $10$ the Flow Label is elided. In this case, shown in Figure \ref{fig:tf.10}, the Traffic Class and Flow Label fields are compressed down to 1 octet.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                        0 1 2 3 4 5 6 7
+                       +-+-+-+-+-+-+-+-+
+                       |ECN|   DSCP    |
+                       +-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{TF = 10: Traffic Class carried in-line.}\label{fig:tf.10}
+\end{figure}
+
+The TF value $00$ is used when Traffic Class and Flow Label are elided.
+
+\subsubsection{Next Header, Hop Limit and Context Identifier Extension}
+The Next Header (NH) field can be carried in-line, if NH field in the IPHC Base Encoding is set to $0$ (see Figure \ref{fig:iphc.format}), or, otherwise, compressed using NHC encoding, discussed in Section \ref{sec:nhc}.
+
+The IPHC encoding allows for compression of some commonly-used IPv6 Hop Limit (HLIM) values.
+If the LoWPAN is a mesh-under stub, a Hop Limit of 1 for inbound and a default value such as 64 for outbound are usually enough for application layer data traffic.
+Additionally, a hop-limit value of 255 is often used.
+This encoding enables to compress the IPv6 Hop Limit field in those cases and they are identified by the HLIM field values $01$, $10$ and $11$ respectively in the IPHC Base Encoding (shown in Figure \ref{fig:iphc.format}).
+The value $00$ specifies that Hop Limit field is carried in-line.
+
+Stateful address compression expects that a conceptual context is shared between the node that compresses a packet and the node(s) that need to expand it.
+However, the Internet Draft \cite{draft-hc-06} does not specify how the contexts are shared, maintained and what information is contained within a context.
+If the Context Identifier (CID) field (see Figure \ref{fig:iphc.format}) is set to 1 then an additional 8-bit Context Identifier Extension field is used and immediately follows the IPHC Base Encoding.
+This field identifies the pair of contexts used for the IPv6 source and destination address compression.
+The context identifiers are 4 bits long.
+
+\subsubsection{Source and Destination Addresses}
+The IPHC encoding supports both stateless and stateful context-based compressions.
+The stateless compression makes use of the link-local prefixes and link-local addresses (derivable from the corresponding link-layer address).
+The stateful address compression can be applied to the source and destination IPv6 addresses when they do not statelessly match the source and destination link layer addresses.
+The stateful compression relies on a conceptual context which is shared between nodes.
+The type of compression used is identified by the Source Address Compression (SAC) and Destination Address Compression (DAC) bits in the IPHC Base Encoding (see Figure \ref{fig:iphc.format}) for the source and destination addresses respectively.
+The value of 0 corresponds to stateless compression, otherwise the stateful compression is applied.
+
+The M bit in the IPHC Base Encoding (see Figure \ref{fig:iphc.format}) is set when the Destination address is a multicast address.
+The Source Address Mode (SAM) and Destination Address Mode (DAM) identify the used addressing schemes.
+The available schemes allow to carry the address in-line, to compress it down to either 64 bits or 16 bits, or to fully elide it.
+
+\subsection{NHC Header Encoding}\label{sec:nhc}
 
@@
 \nocite{rfc2464}
+\nocite{rfc2474}
+\nocite{rfc3168}
 \nocite{rfc4291} 

### 2009-12-13 doc/msc-proposal/msc-proposal.tex
proposal: NHC IPv6 Extened Header

@@
 Bits 2 and 3 specify how the destination address is encoded (the same encoding as for the source address).
-If bit 4 is set then  Traffic Class and Flow Label fields are zero, otherwise their values are not compressed and carried inline.
-Bits 5 and 6 indicate the Next Header: $00$ -- the field is not compressed and carried inlie, $01$ corresponds to UDP, $10$ -- ICMP, $11$ -- TCP.
+If bit 4 is set then  Traffic Class and Flow Label fields are zero, otherwise their values are not compressed and carried in-line.
+Bits 5 and 6 indicate the Next Header: $00$ -- the field is not compressed and carried in-line, $01$ corresponds to UDP, $10$ -- ICMP, $11$ -- TCP.
 The last bit specifies whether HC1 header is immediately followed by more header compression bits of the HC2 encoding format.
@@
 
-\subsection{HC2 Header Encoding}
+\subsection{HC\_UDP Header Encoding}
+The HC\_UDP encoding is used for compression of a UDP header.
+It is applied only if bits $5$ and $6$ in HC1 specify that the protocol that follows the IPv6 header is UDP and bit 7 indicates that the HC2 encoding is used.
+The HC\_UDP encoding allows to compress the UDP header to 4 octets instead of the original 8 octets.
+
+The first $8$ bits of the HC\_UDP encoding specify how the UDP header is compressed.
+Both the source and destination port fields may be compressed to $4$ bits.
+It is indicated by bits $0$ and $1$ of the HC\_UDP encoding for the source and destination ports respectively.
+In such case, the actual $16$-bit port is obtained by calculating $61616$ + short port value, which is a $4$ bit value carried in-line.
+If bit $2$ of the HC\_UDP encoding is set then the UDP length header field is computed from IPv6 header length information, otherwise it's carried in-line.
+HC\_UDP encoding bits 3-7 are reserved for future use.
+The UDP header's checksum field is not compressed and is therefore carried in full.
+
+The non-compressed or partially compressed fields in the UDP header always follow the IPv6 header.
+All present UDP header in-line fields always appear in the same order as the corresponding fields appear in a normal UDP header \cite{rfc780}, e.g., source port, destination port, length, and checksum.
 
 \subsection{IPHC Header Encoding}
+The IPHC encoding format enables effective compression of the IPv6 header relying on information pertaining to the entire 6LoWPAN network.
+The encoding assumes the following common case parameters for 6LoWPAN communication: Version is $6$; Traffic Class and Flow Label are both zero; Payload Length can be inferred from lower layers from either the 6LoWPAN Fragmentation header or the IEEE 802.15.4 header; Hop Limit will be set to a well-known value by the source; addresses assigned to 6LoWPAN interfaces will be formed using the link-local prefix or a single routable prefix assigned to the entire 6LoWPAN network; addresses assigned to 6LoWPAN interfaces are formed with an IID derived directly from either the 64-bit extended or 16-bit short IEEE 802.15.4 addresses.
+The compression mechanism is adapted for these values of the header fields and in such scenario the LOWPAN IPHC can compress the IPv6 header down to two octets (1-octet dispatch, 1-octet LOWPAN\_IPHC encoding) with link-local communication.
+When routing over multiple IP hops, IPHC can compress the IPv6 header down to 7 octets (1-octet dispatch, 1-octet IPHC encoding, 1-octet Hop Limit, 2-octet Source Address, and 2-octet Destination Address).
+In this case, the Hop Limit may not be compressed because it needs to be decremented at each hop and stateful address compression is applied to the source and destination IPv6 addresses because they do not statelessly match the source and destination link layer addresses on intermediate hops.
 
-\subsection{NHC Header Encoding}
+The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$ and a 13 bit IPHC base encoding (shown in Figure \ref{fig:iphc.format}), describing how the IPv6 header is compressed, followed by compressed IPv6 header fields.
+The IPHC encoding may be extended by another octet to support additional contexts.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+    +------------------+---------------+------------------------+
+    |    Dispatch      | IPHC encoding | Compressed IPv6 Header |
+    | (3 bits - 110)   |   (13 bits)   |      Fields            |
+    +------------------+---------------+------------------------+
+\end{verbatim}
+\end{mylisting}
+\caption{IPHC Header}\label{fig:iphc.header}
+\end{figure}
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+       0   1   2   3   4   5   6   7   8   9   0   1   2  
+     +---+---+---+---+---+---+---+---+---+---+---+---+---+
+     |  TF   |NH | HLIM  |CID|SAC|  SAM  | M |DAC|  DAM  |
+     +---+---+---+---+---+---+---+---+---+---+---+---+---+
+\end{verbatim}
+\end{mylisting}
+\caption{IPHC Base Encoding}\label{fig:iphc.format}
+\end{figure}
+
+\subsubsection{Traffic Class and  Flow Label Compression}
+The IPHC supports four compression formats of the Traffic Class and Flow Label (TF).
+The two bit TF field in the IPHC Base Encoding, shown in Figure \ref{fig:iphc.format}, indicates whether these fields are carried in-line in the compressed IPv6 header.
+The Traffic Class field in the IPv6 header comprises 6 bits of Differentiated Services Code Point (DSCP) Field \cite{rfc2474} and 2 bits of Explicit Congestion Notification (ECN) \cite{rfc3168}.
+
+The TF field value $00$ indicates that the Traffic Class and Flow Label are carried in-line with additional 4 bits included to maintain byte-alignment.
+In this case, shown in Figure \ref{fig:tf.00}, Traffic Class and  Flow Label fields are represented by 4 octets in the IPv6 header.
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                        1                   2                   3
+    0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |ECN|   DSCP    |  rsv  |             Flow Label                |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{ TF = 00: Traffic Class and Flow Label carried in-line.}\label{fig:tf.00}
+\end{figure}
+
+The TF value $01$ corresponds to the 3 byte encoding, shown in Figure \ref{fig:tf.01}, when ECN and Flow Label fields are carried in-line, but DSCP field is elided.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                          1                   2
+      0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3
+     +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+     |ECN|rsv|             Flow Label                |
+     +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{TF = 01: Flow Label carried in-line.}\label{fig:tf.01}
+\end{figure}
+
+When TF is $10$ the Flow Label is elided. In this case, shown in Figure \ref{fig:tf.10}, the Traffic Class and Flow Label fields are compressed down to 1 octet.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                        0 1 2 3 4 5 6 7
+                       +-+-+-+-+-+-+-+-+
+                       |ECN|   DSCP    |
+                       +-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{TF = 10: Traffic Class carried in-line.}\label{fig:tf.10}
+\end{figure}
+
+The TF value $00$ is used when Traffic Class and Flow Label are elided.
+
+\subsubsection{Next Header, Hop Limit and Context Identifier Extension}
+The Next Header (NH) field can be carried in-line, if NH field in the IPHC Base Encoding is set to $0$ (see Figure \ref{fig:iphc.format}), or, otherwise, compressed using NHC encoding, discussed in Section \ref{sec:nhc}.
+
+The IPHC encoding allows for compression of some commonly-used IPv6 Hop Limit (HLIM) values.
+If the LoWPAN is a mesh-under stub, a Hop Limit of 1 for inbound and a default value such as 64 for outbound are usually enough for application layer data traffic.
+Additionally, a hop-limit value of 255 is often used.
+This encoding enables to compress the IPv6 Hop Limit field in those cases and they are identified by the HLIM field values $01$, $10$ and $11$ respectively in the IPHC Base Encoding (shown in Figure \ref{fig:iphc.format}).
+The value $00$ specifies that Hop Limit field is carried in-line.
+
+Stateful address compression expects that a conceptual context is shared between the node that compresses a packet and the node(s) that need to expand it.
+However, the Internet Draft \cite{draft-hc-06} does not specify how the contexts are shared, maintained and what information is contained within a context.
+If the Context Identifier (CID) field (see Figure \ref{fig:iphc.format}) is set to 1 then an additional 8-bit Context Identifier Extension field is used and immediately follows the IPHC Base Encoding.
+This field identifies the pair of contexts used for the IPv6 source and destination address compression.
+The context identifiers are 4 bits long.
+
+\subsubsection{Source and Destination Addresses}
+The IPHC encoding supports both stateless and stateful context-based compressions.
+The stateless compression makes use of the link-local prefixes and link-local addresses (derivable from the corresponding link-layer address).
+The stateful address compression can be applied to the source and destination IPv6 addresses when they do not statelessly match the source and destination link layer addresses.
+The stateful compression relies on a conceptual context which is shared between nodes.
+The type of compression used is identified by the Source Address Compression (SAC) and Destination Address Compression (DAC) bits in the IPHC Base Encoding (see Figure \ref{fig:iphc.format}) for the source and destination addresses respectively.
+The value of 0 corresponds to stateless compression, otherwise the stateful compression is applied.
+
+The M bit in the IPHC Base Encoding (see Figure \ref{fig:iphc.format}) is set when the Destination address is a multicast address.
+The Source Address Mode (SAM) and Destination Address Mode (DAM) identify the used addressing schemes.
+The available schemes allow to carry the address in-line, to compress it down to either 64 bits or 16 bits, or to fully elide it.
+
+\subsection{NHC Header Encoding}\label{sec:nhc}
+The Next Header field in the IPHC encoding indicates whether the following header is compressed using NHC format.
+If so, the compressed IPv6 header is immediately followed by the NHC encoding.
+In contrast to HC2, the NHC encoding provides a flexible and extensible mechanism for arbitrary header compression, and not only for UDP, TCP, and ICMPv6.
+Compression formats for different next headers are identified by a variable-length bit-pattern immediately following the IPHC compressed header.
+
+\subsubsection{IPv6 Extension Header Compression}
+The NHC encoding for IPv6 Extension Headers is composed of a single octet followed by the IPv6 Extension Header.
+IPv6 Extension Headers use  $1110$ bit pattern as the first 4 bits in the NHC encoding.
+The latter 3 bits is an IPv6 Extension Header ID (EID) which identifies the IPv6 Extension Header immediately following the NHC octet.
+Currently, the following values of the EID field have been specified:  $0$ -- IPv6 Hop-by-Hop Options Header \cite{rfc2460}; $1$ -- IPv6 Routing Header \cite{rfc2460};  $2$ -- IPv6 Fragment Header \cite{rfc2460}; $3$ -- IPv6 Destination Options Header \cite{rfc2460}; $4$ -- IPv6 Mobility Header \cite{rfc3775}; $5$ -- Reserved; $6$ -- Reserved; $7$ -- IPv6 Header.
+The last bit of the NHC encoding indicates whether the NHC compression is used for the following header.
+
+\subsubsection{UDP Header Compression}
+   Using NHC, this document defines a compression mechanism for
+   UDP.  While [RFC4944] defines a compression mechanism for UDP, that
+   mechanism does not enable checksum compression when rendered possible
+   by additional upper layer mechanisms such as upper layer Message
+   Integrity Check (MIC).  This specification adds the capability to
+   elide the UDP checksum over the LoWPAN, which enables to save an
+   additional pair of octets.
+
+   Also using NHC, this document defines encoding formats for
+   IPv6-in-IPv6 encapsulation as well as IPv6 Extension Headers.  With
+   HC1 and HC2, chains of next headers can not be encoded
+   efficiently.
 
@@
 \nocite{eui64} 
+\nocite{rfc780}
 \nocite{rfc2460}
 \nocite{rfc2464}
+\nocite{rfc2474}
+\nocite{rfc3168}
+\nocite{rfc3775}
 \nocite{rfc4291} 

### 2009-12-14 doc/msc-proposal/msc-proposal.tex
proposal: NHC UDP

@@
 
-\subsection{HC\_UDP Header Encoding}
+\subsection{HC\_UDP Header Encoding}\label{sec:udp.header.enc}
 The HC\_UDP encoding is used for compression of a UDP header.
@@
 HC\_UDP encoding bits 3-7 are reserved for future use.
-The UDP header's checksum field is not compressed and is therefore carried in full.
+The UDP header's checksum field is not compressed and, therefore, carried in full.
 
@@
 If so, the compressed IPv6 header is immediately followed by the NHC encoding.
-In contrast to HC2, the NHC encoding provides a flexible and extensible mechanism for arbitrary header compression, and not only for UDP, TCP, and ICMPv6.
-Compression formats for different next headers are identified by a variable-length bit-pattern immediately following the IPHC compressed header.
+In contrast to HC2, the NHC encoding provides a flexible and extensible mechanism for arbitrary header compression, not only for UDP, TCP, and ICMPv6.
+With NHC, chains of next headers can be encoded efficiently, which is not possible with HC1 and HC2.
+NHC compression formats for different next headers are identified by a variable-length bit-pattern immediately following the IPHC compressed header.
 
 \subsubsection{IPv6 Extension Header Compression}
-The NHC encoding for IPv6 Extension Headers is composed of a single octet followed by the IPv6 Extension Header.
-IPv6 Extension Headers use  $1110$ bit pattern as the first 4 bits in the NHC encoding.
-The latter 3 bits is an IPv6 Extension Header ID (EID) which identifies the IPv6 Extension Header immediately following the NHC octet.
+The NHC encoding for IPv6 Extension Headers (shown in Figure \ref{fig:ipv6.ext.header}) is composed of a single octet followed by the IPv6 Extension Header.
+IPv6 Extension Headers use  $1110$ bit pattern as bits $0$ through $3$ in the NHC encoding.
+The latter 3 bits represent an IPv6 Extension Header ID (EID) which identifies the IPv6 Extension Header immediately following the NHC octet.
 Currently, the following values of the EID field have been specified:  $0$ -- IPv6 Hop-by-Hop Options Header \cite{rfc2460}; $1$ -- IPv6 Routing Header \cite{rfc2460};  $2$ -- IPv6 Fragment Header \cite{rfc2460}; $3$ -- IPv6 Destination Options Header \cite{rfc2460}; $4$ -- IPv6 Mobility Header \cite{rfc3775}; $5$ -- Reserved; $6$ -- Reserved; $7$ -- IPv6 Header.
@@
 
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                0   1   2   3   4   5   6   7
+              +---+---+---+---+---+---+---+---+
+              | 1 | 1 | 1 | 0 |    EID    |NH |
+              +---+---+---+---+---+---+---+---+
+\end{verbatim}
+\end{mylisting}
+\caption{IPv6 Extension Header Encoding}\label{fig:ipv6.ext.header}
+\end{figure}
+
 \subsubsection{UDP Header Compression}
-   Using NHC, this document defines a compression mechanism for
-   UDP.  While [RFC4944] defines a compression mechanism for UDP, that
-   mechanism does not enable checksum compression when rendered possible
-   by additional upper layer mechanisms such as upper layer Message
-   Integrity Check (MIC).  This specification adds the capability to
-   elide the UDP checksum over the LoWPAN, which enables to save an
-   additional pair of octets.
+The NHC format, shown in Figure \ref{fig:udp.header.enc}, provides a compression mechanism for a UDP header.
+Bits $0$ through $4$ of the encoding are the NHC ID and the value $11110$ indicates the UDP header compression encoding.
 
-   Also using NHC, this document defines encoding formats for
-   IPv6-in-IPv6 encapsulation as well as IPv6 Extension Headers.  With
-   HC1 and HC2, chains of next headers can not be encoded
-   efficiently.
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                0   1   2   3   4   5   6   7
+              +---+---+---+---+---+---+---+---+
+              | 1 | 1 | 1 | 1 | 0 | C |   P   |
+              +---+---+---+---+---+---+---+---+
+\end{verbatim}
+\end{mylisting}
+\caption{UDP Header Encoding}\label{fig:udp.header.enc}
+\end{figure}
+
+The UDP Length field is always elided and inferred from lower layers.
+If the upper layer protocol provides message integrity check, the Checksum field may be elided from the UDP header.
+Such behaviour is enabled by the compression format and when C bit is set the Checksum is not included into the UDP header.
+
+This format also allows to compress down well-known ports (0xF0Bx).	When the P field has the value $01$, all 16 bits for the Source Port are carried in-line, whereas first 8 bits of the Destination Port is 0xF0 and elided, and the latest 8 bits are carried in-line.
+The value $10$ of the P field indicates that all 16 bits for the Destination Port are carried in-line and the Source Port is encoded in the same way as the Destination port in the previous case.
+If the P field is $11$, first 12 bits of both the Source Port and the Destination Port are 0xF0B and elided.
+The remaining 4 bits for each are carried in-line.
+In case of both ports are carried in-line, is $00$.
+Fields carried in-line appear in the same order as they do in the UDP header format.
 

### 2009-12-16 trunk/doc/msc-proposal/msc-proposal.tex
proposal: nd overview

@@
 %%
-\chapter{6LoWPAN Networks}
+\chapter{6LoWPAN Networks}\label{ch:6lowpan}
 A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
@@
 
-6lowpan (IPv6 over Low power Wireless Personal Area Networks) is a working group within the IETF concerned with the specification of mechanisms to allow IPv6 packets to be sent to and received from over IEEE 802.15.4 based networks.
+6lowpan (Internet Protocol Version 6 (IPv6) over Low power Wireless Personal Area Networks) is a working group within the IETF concerned with the specification of mechanisms to allow IPv6 packets to be sent to and received from over IEEE 802.15.4 based networks.
 The working group has already completed two documents: \cite{rfc4919} that documents and discusses the problem statement and \cite{rfc4944} that defines the format for the adaptation between IPv6 and 802.15.4.
@@
 
-As described in Chapter \ref{ch:ieee802.15.4}, IEEE 802.15.4 defines four types of frames: beacon frames, MAC command frames, acknowledgement frames, and data frames.
+%As described in Chapter \ref{ch:ieee802.15.4}, IEEE 802.15.4 defines four types of frames: beacon frames, MAC command frames, acknowledgement frames, and data frames.
 IPv6 packets are carried on data frames.
 Data frames may optionally request acknowledgments to aid link-layer recovery.
+
+This chapter describes the Adaptation Layer and Header Compression Formats that have been introduced to allow the transmission of IPv6 packets over 802.15.4 links.
 
@@
 
-\section{Header Compression}
+\section{Header Compression}\label{sec:hc}
 As described in Section \ref{sec:adapt.layer}, 81 octets are left in a IEEE 802.15.4 frame for IPv6.
@@
 
-The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$, a 13 bit IPHC base encoding (shown in Figure \ref{fig:iphc.format}), describing how the IPv6 header is compressed, followed by compressed IPv6 header fields.
+The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$ and a 13 bit IPHC base encoding (shown in Figure \ref{fig:iphc.format}), describing how the IPv6 header is compressed, followed by compressed IPv6 header fields.
 The IPHC encoding may be extended by another octet to support additional contexts.
@@
 \subsection{NHC Header Encoding}\label{sec:nhc}
+The Next Header field in the IPHC encoding indicates whether the following header is compressed using NHC format.
+If so, the compressed IPv6 header is immediately followed by the NHC encoding.
+In contrast to HC2, the NHC encoding provides a flexible and extensible mechanism for arbitrary header compression, not only for UDP, TCP, and ICMPv6.
+With NHC, chains of next headers can be encoded efficiently, which is not possible with HC1 and HC2.
+NHC compression formats for different next headers are identified by a variable-length bit-pattern immediately following the IPHC compressed header.
 
-\chapter{6LoWPAN Neighbor Discovery}
+\subsubsection{IPv6 Extension Header Compression}
+The NHC encoding for IPv6 Extension Headers (shown in Figure \ref{fig:ipv6.ext.header}) is composed of a single octet followed by the IPv6 Extension Header.
+IPv6 Extension Headers use  $1110$ bit pattern as bits $0$ through $3$ in the NHC encoding.
+The latter 3 bits represent an IPv6 Extension Header ID (EID) which identifies the IPv6 Extension Header immediately following the NHC octet.
+Currently, the following values of the EID field have been specified:  $0$ -- IPv6 Hop-by-Hop Options Header \cite{rfc2460}; $1$ -- IPv6 Routing Header \cite{rfc2460};  $2$ -- IPv6 Fragment Header \cite{rfc2460}; $3$ -- IPv6 Destination Options Header \cite{rfc2460}; $4$ -- IPv6 Mobility Header \cite{rfc3775}; $5$ -- Reserved; $6$ -- Reserved; $7$ -- IPv6 Header.
+The last bit of the NHC encoding indicates whether the NHC compression is used for the following header.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                0   1   2   3   4   5   6   7
+              +---+---+---+---+---+---+---+---+
+              | 1 | 1 | 1 | 0 |    EID    |NH |
+              +---+---+---+---+---+---+---+---+
+\end{verbatim}
+\end{mylisting}
+\caption{IPv6 Extension Header Encoding}\label{fig:ipv6.ext.header}
+\end{figure}
+
+\subsubsection{UDP Header Compression}
+The NHC format, shown in Figure \ref{fig:udp.header.enc}, provides a compression mechanism for a UDP header.
+Bits $0$ through $4$ of the encoding are the NHC ID and the value $11110$ indicates the UDP header compression encoding.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+                0   1   2   3   4   5   6   7
+              +---+---+---+---+---+---+---+---+
+              | 1 | 1 | 1 | 1 | 0 | C |   P   |
+              +---+---+---+---+---+---+---+---+
+\end{verbatim}
+\end{mylisting}
+\caption{UDP Header Encoding}\label{fig:udp.header.enc}
+\end{figure}
+
+The UDP Length field is always elided and inferred from lower layers.
+If the upper layer protocol provides message integrity check, the Checksum field may be elided from the UDP header.
+Such behaviour is enabled by the compression format and when C bit is set the Checksum is not included into the UDP header.
+
+This format also allows to compress down well-known ports (0xF0Bx).	When the P field has the value $01$, all 16 bits for the Source Port are carried in-line, whereas first 8 bits of the Destination Port is 0xF0 and elided, and the latest 8 bits are carried in-line.
+The value $10$ of the P field indicates that all 16 bits for the Destination Port are carried in-line and the Source Port is encoded in the same way as the Destination port in the previous case.
+If the P field is $11$, first 12 bits of both the Source Port and the Destination Port are 0xF0B and elided.
+The remaining 4 bits for each are carried in-line.
+In case of both ports are carried in-line, is $00$.
+Fields carried in-line appear in the same order as they do in the UDP header format.
+
+\chapter{Neighbor Discovery Protocol}\label{ch:nd}
+The Neighbor Discovery (ND) protocol for IPv6 \cite{rfc4861} provides for basic bootstrapping and network operation.
+Nodes (hosts and routers) use ND to determine the link-layer addresses for neighbors known to reside on attached links.
+Hosts also use ND to find neighboring routers that are willing to forward packets.
+In the previous Chapter \ref{ch:6lowpan} a mechanism which allows to carry IPv6 packets over IEEE 802.15.4 networks with the help of an adaptation layer has been described.
+However, the standard IPv6 Neighbor Discovery has several problems when using it with 6LoWPANs.
+IPv6 ND heavily uses multicast capabilities, whereas it is very expensive and not desirable in a low-power, lossy wireless network.
+Therefore, the classic ND is not suitable for 6LoWPANs.
+The Internet Draft \cite{draft-nd-07} specifies a Neighbor Discovery mechanism optimized for LoWPAN networks.
+This chapter describes the 6LoWPAN ND protocol.
+
+\section{Overview}\label{Overview}
+6LoWPAN Neighbor Discovery (6LoWPAN-ND) provides an optimized neighbor discovery mechanism sufficient for LoWPAN operation.
+6LoWPAN-ND introduces a node registration mechanism optimizing the node-router interface.
+This mechanism requires no flooding and reduces link-local multicast frequency.
+Nodes in the LoWPAN register with Routers and Edge Routers (IPv6 routers that interconnect the LoWPAN to another IP network), creating state about nodes attached to that router (binding table) and about all IPv6 addresses in the LoWPAN.
+This state information is stored within a conceptual data structure, Whiteboard, located at Edge Routers (ERs).
+The Whiteboard makes use of soft bindings that contain an Owner Interface Identifier, Owner Nonce, IPv6 address and a remaining lifetime of the binding.
+Nodes send periodic registration messages in order to maintain their bindings.
+The detection of duplicate addresses (DAD) is performed across the entire LoWPAN using a lookup on the Whiteboard using the node registration mechanism.
+
+6LoWPAN-ND makes use of Router Solicitation (RS) / Router Advertisement (RA) message exchanges similar to classic ND.
+In 6LoWPAN-ND RA messages may carry additional options for context dissemination and are reduced in size.
+In addition to RS and RA messages, 6LoWPAN-ND defines two new ICMP packet types: Node Registration (NR), which is sent by a node to an Edge Router to register a binding for an IPv6 address in the Whiteboard, and Node Confirmation (NC) by which an Edge Router replies to the registering node.
 
@@
 %%
-\chapter*{Conclusion}
-\addcontentsline{toc}{chapter}{Conclusion}
+\chapter*{Proposed work}
+\addcontentsline{toc}{chapter}{Proposed work}
 
@@
 \nocite{rfc3168}
+\nocite{rfc3775}
 \nocite{rfc4291} 
@@
 \nocite{draft-hc-06} 
-\nocite{draft-nd-06}
+\nocite{draft-nd-07}
 

### 2009-12-16 trunk/doc/msc-proposal/msc-nd.tex
proposal: added structure

\chapter{6LoWPAN Neighbor Discovery Protocol}\label{ch:nd}
The Neighbor Discovery (ND) protocol for IPv6 \cite{rfc4861} provides for basic bootstrapping and network operation. Nodes (hosts and routers) use ND to determine the link-layer addresses for neighbors known to reside on attached links. Hosts also use ND to find neighboring routers that are willing to forward packets. In the previous Chapter \ref{ch:6lowpan} a mechanism which allows to carry IPv6 packets over IEEE 802.15.4 networks with the help of an adaptation layer has been described. However, the standard IPv6 Neighbor Discovery has several problems when using it with 6LoWPANs. IPv6 ND heavily uses multicast capabilities, whereas it is very expensive and not desirable in a low-power, lossy wireless network. Therefore, the classic ND is not suitable for 6LoWPANs. The Internet Draft \cite{draft-nd-07} specifies an optimized neighbor discovery mechanism sufficient for LoWPAN operation. This chapter describes the 6LoWPAN ND protocol.

\section{Basic features}\label{nd.basic}
6LoWPAN Neighbor Discovery (6LoWPAN-ND) introduces a node registration mechanism optimizing the node-router interface.  This mechanism requires no flooding and reduces link-local multicast frequency. Nodes in the LoWPAN register with Routers and Edge Routers (IPv6 routers that interconnect the LoWPAN to another IP network), creating state about nodes attached to that router (binding table) and about all IPv6 addresses in the LoWPAN. A list of all IPv6 addresses in the LoWPAN are stored within a conceptual data structure, Whiteboard, located at Edge Routers (ERs). The Whiteboard makes use of soft bindings that contain an Owner Interface Identifier, Owner Nonce, IPv6 address and a remaining lifetime of the binding. Nodes send periodic registration messages in order to maintain their bindings in the Router binding tables and Edge Router Whiteboard.

6LoWPAN-ND makes use of Router Solicitation (RS)/Router Advertisement (RA) message exchanges similar to classic ND. In 6LoWPAN-ND RA messages may carry additional options for context dissemination and are reduced in size. In addition to RS and RA messages, 6LoWPAN-ND defines two new ICMP packet types: Node Registration (NR), which is sent by a node to an Edge Router to register a binding for an IPv6 address in the Whiteboard, and Node Confirmation (NC) by which an Edge Router replies to the registering node.

\section{Bootstrapping}
Bootstrapping of a LoWPAN node consists of several steps.  At first, a node is required to autoconfigure at least one address, a link-local address, which is derived from the IEEE 64-bit extended MAC. In order to receive RAs from routers, a node joins the all-nodes multicast address and, if the node is a router, the all-routers multicast address. Once the interfaces have been initialized, a node listens for Router Advertisements (RA) from Edge Routers or LoWPAN Routers, or broadcasts a Router Solicitation (RS). Upon receipt of the RA, the node forms an optimistic global unique address with stateless address autoconfiguration and chooses one or more default routers.

The constructed global address is tentative or optimistic as long as the binding is not confirmed by the ER. To accomplish this, the node performs initial registration with the ER by sending a unicast Node Registration (NR) message. The destination address of the NR message is the link-local unicast address of the ER, while the IPv6 unspecified address is used as the source address.  The NR message includes an Address Option for each address to be registered.

The ER replies with a Node Confirmation(NC), which includes the set of addresses confirmed to be bound to the Whiteboard of the ER.  The
source of the packet is the link-local address of LoWPAN Router and the destination address is the link-local address of the node. Once the node has received the NC, it is capable to send packets to any IPv6 address inside or outside the LoWPAN. The detection of duplicate addresses (DAD) is performed as part of the node registration process by the ER across the entire LoWPAN using a lookup on the Whiteboard.

This information about link-local addresses is collected during the node registration process. Nodes store the information about router link-local addresses in their default router list, while routers keep the information about nodes in their binding tables.

### 2009-12-16 trunk/doc/msc-proposal/msc-proposal.tex
proposal: added structure

@@
 %%
-\chapter{6LoWPAN Networks}\label{ch:6lowpan}
+\chapter{Transmission of IPv6 Packets over IEEE 802.15.4}\label{ch:6lowpan}
 A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
@@
 
-\chapter{Neighbor Discovery Protocol}\label{ch:nd}
-The Neighbor Discovery (ND) protocol for IPv6 \cite{rfc4861} provides for basic bootstrapping and network operation.
-Nodes (hosts and routers) use ND to determine the link-layer addresses for neighbors known to reside on attached links.
-Hosts also use ND to find neighboring routers that are willing to forward packets.
-In the previous Chapter \ref{ch:6lowpan} a mechanism which allows to carry IPv6 packets over IEEE 802.15.4 networks with the help of an adaptation layer has been described.
-However, the standard IPv6 Neighbor Discovery has several problems when using it with 6LoWPANs.
-IPv6 ND heavily uses multicast capabilities, whereas it is very expensive and not desirable in a low-power, lossy wireless network.
-Therefore, the classic ND is not suitable for 6LoWPANs.
-The Internet Draft \cite{draft-nd-07} specifies a Neighbor Discovery mechanism optimized for LoWPAN networks.
-This chapter describes the 6LoWPAN ND protocol.
+\input{msc-routing}
 
-\section{Overview}\label{Overview}
-6LoWPAN Neighbor Discovery (6LoWPAN-ND) provides an optimized neighbor discovery mechanism sufficient for LoWPAN operation.
-6LoWPAN-ND introduces a node registration mechanism optimizing the node-router interface.
-This mechanism requires no flooding and reduces link-local multicast frequency.
-Nodes in the LoWPAN register with Routers and Edge Routers (IPv6 routers that interconnect the LoWPAN to another IP network), creating state about nodes attached to that router (binding table) and about all IPv6 addresses in the LoWPAN.
-This state information is stored within a conceptual data structure, Whiteboard, located at Edge Routers (ERs).
-The Whiteboard makes use of soft bindings that contain an Owner Interface Identifier, Owner Nonce, IPv6 address and a remaining lifetime of the binding.
-Nodes send periodic registration messages in order to maintain their bindings.
-The detection of duplicate addresses (DAD) is performed across the entire LoWPAN using a lookup on the Whiteboard using the node registration mechanism.
-
-6LoWPAN-ND makes use of Router Solicitation (RS) / Router Advertisement (RA) message exchanges similar to classic ND.
-In 6LoWPAN-ND RA messages may carry additional options for context dissemination and are reduced in size.
-In addition to RS and RA messages, 6LoWPAN-ND defines two new ICMP packet types: Node Registration (NR), which is sent by a node to an Edge Router to register a binding for an IPv6 address in the Whiteboard, and Node Confirmation (NC) by which an Edge Router replies to the registering node.
-
+\input{msc-nd}
 %%

### 2009-12-16 trunk/doc/msc-proposal/msc-routing.tex
proposal: added structure

\chapter{6LoWPAN Routing}\label{ch:routing}
Neither the IEEE 802.15.4 \cite{ieee802.15.4} standard nor the 6LoWPAN format specification \cite{rfc4944} define how mesh topologies could be obtained and maintained. In 6LoWPAN routing can be perforem either in the IP-layer, using a Route Over approach, or in the adaptation layer described in Section \ref{subsec:mesh.header}, using the Mesh Under approach.

### 2009-12-16 trunk/doc/msc-proposal/msc-routing.tex
proposal: ND

@@
 Neither the IEEE 802.15.4 \cite{ieee802.15.4} standard nor the 6LoWPAN format specification \cite{rfc4944} define how mesh topologies could be obtained and maintained.
-In 6LoWPAN routing can be perforem either in the IP-layer, using a Route Over approach, or in the adaptation layer described in Section \ref{subsec:mesh.header}, using the Mesh Under approach.
+In 6LoWPAN routing can be performed either in the IP-layer, using a Route Over approach, or in the adaptation layer described in Section \ref{subsec:mesh.header}, using the Mesh Under approach.
+The Mesh Under approach performs the multi-hop communication below the IP link and therefore a 6LoWPAN is seen as a single IP link.

### 2009-12-16 doc/msc-proposal/msc-nd.tex
proposal: ND

--- 2009-12-16 trunk/doc/msc-proposal/msc-nd.tex
@@
 The Neighbor Discovery (ND) protocol for IPv6 \cite{rfc4861} provides for basic bootstrapping and network operation.
-Nodes (hosts and routers) use ND to determine the link-layer addresses for neighbors known to reside on attached links.
-Hosts also use ND to find neighboring routers that are willing to forward packets.
-In the previous Chapter \ref{ch:6lowpan} a mechanism which allows to carry IPv6 packets over IEEE 802.15.4 networks with the help of an adaptation layer has been described.
+Nodes use ND to determine the link-layer addresses for neighbors known to reside on attached links.
+ND also allows to find neighboring routers that are willing to forward packets.
 However, the standard IPv6 Neighbor Discovery has several problems when using it with 6LoWPANs.
 IPv6 ND heavily uses multicast capabilities, whereas it is very expensive and not desirable in a low-power, lossy wireless network.
+Moreover, in a Route Over configuration LoWPAN links are non-transitive and composed of a large amount of overlapping radio ranges, but the classic ND was not designed for non-transitive links.
 Therefore, the classic ND is not suitable for 6LoWPANs.
 The Internet Draft \cite{draft-nd-07} specifies an optimized neighbor discovery mechanism sufficient for LoWPAN operation.
-This chapter describes the 6LoWPAN ND protocol.
+The 6LoWPAN Neighbor Discovery (6LoWPAN-ND) supports both the Mesh Under and Route Over routing solutions for multihop forwarding.
+This chapter describes the 6LoWPAN-ND protocol.
 
-\section{Basic features}\label{nd.basic}
-6LoWPAN Neighbor Discovery (6LoWPAN-ND) introduces a node registration mechanism optimizing the node-router interface.
+\section{Basics}\label{nd.basic}
+A 6LoWPAN network comprises nodes which are either hosts or routers.
+A host is a node that only sources or sinks IPv6 datagrams, whereas a router forwards datagrams between arbitrary source-destination pairs.
+Routers are present only in a Route Over configuration, where the network is composed of overlapping link-local scopes, to overcome non-transitive nature of links.
+6LoWPAN-ND also introduces a notion of Edge Routers that interconnect the LoWPAN to another IP network.
+
+6LoWPAN-ND introduces a node registration mechanism optimizing the node-router interface.
 This mechanism requires no flooding and reduces link-local multicast frequency.
-Nodes in the LoWPAN register with Routers and Edge Routers (IPv6 routers that interconnect the LoWPAN to another IP network), creating state about nodes attached to that router (binding table) and about all IPv6 addresses in the LoWPAN.
+Nodes in the LoWPAN register with Routers and Edge Routers, creating state about nodes attached to that router and about all IPv6 addresses in the LoWPAN.
+
+(binding table) \\
+Routers only in the Route Over\\
+(IPv6 routers that interconnect the LoWPAN to another IP network)\\
+
 A list of all IPv6 addresses in the LoWPAN are stored within a conceptual data structure, Whiteboard, located at Edge Routers (ERs).
@@
 Nodes send periodic registration messages in order to maintain their bindings in the Router binding tables and Edge Router Whiteboard.
+
+* Hosts;\\
+* Routers;\\
+* Edge Routers;\\
+* Simple ER; \\
+* Extended ER;
+
+\section{Message Formats}
 

### 2009-12-18 doc/msc-proposal/msc-proposal.tex
proposal: ND

@@
 %%
-\chapter{6LoWPAN Networks}
+\chapter{Transmission of IPv6 Packets over 802.15.4}\label{ch:6lowpan}
 A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
@@
 
-\chapter{6LoWPAN Neighbor Discovery}
+\input{msc-routing}
 
+\input{msc-nd}
 %%

### 2009-12-18 doc/msc-proposal/msc-routing.tex
proposal: ND

\chapter{6LoWPAN Routing}\label{ch:routing}
Neither the IEEE 802.15.4 \cite{ieee802.15.4} standard nor the 6LoWPAN format specification \cite{rfc4944} define how mesh topologies could be obtained and maintained. In 6LoWPAN routing can be performed either in the IP-layer, using a Route Over approach, or in the adaptation layer, described in Section \ref{subsec:mesh.header}, using the Mesh Under approach. The Mesh Under approach performs the multi-hop communication below the IP link and therefore a 6LoWPAN is seen as a single IP link.

A 6LoWPAN network comprises nodes which are either hosts or routers. A host is a node that only sources or sinks IPv6 datagrams, whereas a router forwards datagrams between arbitrary source-destination pairs. Routers are present only in a Route Over configuration, where the network is composed of overlapping link-local scopes, to overcome non-transitive nature of links.

### 2009-12-18 trunk/doc/msc-proposal/msc-nd.tex
proposal: ND

@@
 The Neighbor Discovery (ND) protocol for IPv6 \cite{rfc4861} provides for basic bootstrapping and network operation.
-Nodes (hosts and routers) use ND to determine the link-layer addresses for neighbors known to reside on attached links.
-Hosts also use ND to find neighboring routers that are willing to forward packets.
-In the previous Chapter \ref{ch:6lowpan} a mechanism which allows to carry IPv6 packets over IEEE 802.15.4 networks with the help of an adaptation layer has been described.
+Nodes use ND to determine the link-layer addresses for neighbors known to reside on attached links, as well as to find neighboring routers that are willing to forward packets.
 However, the standard IPv6 Neighbor Discovery has several problems when using it with 6LoWPANs.
 IPv6 ND heavily uses multicast capabilities, whereas it is very expensive and not desirable in a low-power, lossy wireless network.
+Moreover, in a Route Over configuration LoWPAN links are non-transitive and composed of a large amount of overlapping radio ranges, but the classic ND was not designed for such links.
 Therefore, the classic ND is not suitable for 6LoWPANs.
+
 The Internet Draft \cite{draft-nd-07} specifies an optimized neighbor discovery mechanism sufficient for LoWPAN operation.
-This chapter describes the 6LoWPAN ND protocol.
+The 6LoWPAN Neighbor Discovery (6LoWPAN-ND) introduces a node registration mechanism optimizing the node-router interface, which requires no flooding and reduces link-local multicast frequency.
+The concept of a LoWPAN Whiteboard located at Edge Routers (routers that connect a LoWPAN to another IP network) is introduced, which allows for Duplicate Address Detection for the entire LoWPAN.
+The solution supports both Mesh Under and Route Over configurations for multihop forwarding.
+This chapter describes the specified 6LoWPAN-ND protocol.
 
-\section{Basic features}\label{nd.basic}
-6LoWPAN Neighbor Discovery (6LoWPAN-ND) introduces a node registration mechanism optimizing the node-router interface.
-This mechanism requires no flooding and reduces link-local multicast frequency.
-Nodes in the LoWPAN register with Routers and Edge Routers (IPv6 routers that interconnect the LoWPAN to another IP network), creating state about nodes attached to that router (binding table) and about all IPv6 addresses in the LoWPAN.
-A list of all IPv6 addresses in the LoWPAN are stored within a conceptual data structure, Whiteboard, located at Edge Routers (ERs).
-The Whiteboard makes use of soft bindings that contain an Owner Interface Identifier, Owner Nonce, IPv6 address and a remaining lifetime of the binding.
+%All nodes in the LoWPAN register with Routers and Edge Routers, though Routers are present only in a Route Over configuration.
+Each Router maintains a set of information about nodes that are currently registered through it, called the binding table.
+All IPv6 addresses in the LoWPAN are stored within a conceptual data structure, Whiteboard, located at Edge Routers (ERs).
 Nodes send periodic registration messages in order to maintain their bindings in the Router binding tables and Edge Router Whiteboard.
+
+\section{Bootstrapping and Basic Operation}
+Bootstrapping of a LoWPAN node consists of several steps.
+At first, a node is required to autoconfigure at least one address, a link-local address, which is derived from the IEEE 64-bit extended MAC.
+In order to receive RAs from routers, a node joins the all-nodes multicast address and, if the node is a router, the all-routers multicast address.
+Once the interfaces have been initialized, a node listens for Router Advertisements (RA) from Edge Routers or LoWPAN Routers, or broadcasts a Router Solicitation (RS).
+Upon receipt of the RA, the node forms an optimistic global unique address with stateless address autoconfiguration and chooses one or more default routers (routers are present only in a Route Over configuration).
+
+The constructed global address is tentative as long as the binding is not confirmed by the ER.
+To accomplish this, the node performs initial registration with the ER by sending a unicast Node Registration (NR) message with a list of all addresses it wants to register.
+The ER replies with a Node Confirmation(NC), which includes the set of addresses confirmed to be bound to the Whiteboard of the ER.
+Once the node has received the NC message, it is capable to send packets to any IPv6 address inside or outside the LoWPAN.
+Each router maintains a set of information about nodes that are currently registered through it, called the binding table.
+Nodes send periodic registration messages in order to maintain their bindings in the router binding tables and Edge Router Whiteboard.
+The detection of duplicate addresses (DAD) is also performed as part of the node registration process by the ER using a lookup on the Whiteboard.
+
+This information about link-local addresses is collected during the node registration process.
+Nodes store the information about router link-local addresses in their default router list, while routers keep the information about nodes in their binding tables.
+
+\section{Message Formats}
 
@@
 In addition to RS and RA messages, 6LoWPAN-ND defines two new ICMP packet types: Node Registration (NR), which is sent by a node to an Edge Router to register a binding for an IPv6 address in the Whiteboard, and Node Confirmation (NC) by which an Edge Router replies to the registering node.
-
-\section{Bootstrapping}
-Bootstrapping of a LoWPAN node consists of several steps.
-At first, a node is required to autoconfigure at least one address, a link-local address, which is derived from the IEEE 64-bit extended MAC.
-In order to receive RAs from routers, a node joins the all-nodes multicast address and, if the node is a router, the all-routers multicast address.
-Once the interfaces have been initialized, a node listens for Router Advertisements (RA) from Edge Routers or LoWPAN Routers, or broadcasts a Router Solicitation (RS).
-Upon receipt of the RA, the node forms an optimistic global unique address with stateless address autoconfiguration and chooses one or more default routers.
-
-The constructed global address is tentative or optimistic as long as the binding is not confirmed by the ER.
-To accomplish this, the node performs initial registration with the ER by sending a unicast Node Registration (NR) message.
-The destination address of the NR message is the link-local unicast address of the ER, while the IPv6 unspecified address is used as the source address.
-The NR message includes an Address Option for each address to be registered.
-
-The ER replies with a Node Confirmation(NC), which includes the set of addresses confirmed to be bound to the Whiteboard of the ER.  The
-source of the packet is the link-local address of LoWPAN Router and the destination address is the link-local address of the node.
-Once the node has received the NC, it is capable to send packets to any IPv6 address inside or outside the LoWPAN.
-The detection of duplicate addresses (DAD) is performed as part of the node registration process by the ER across the entire LoWPAN using a lookup on the Whiteboard.
-
-This information about link-local addresses is collected during the node registration process.
-Nodes store the information about router link-local addresses in their default router list, while routers keep the information about nodes in their binding tables.

### 2009-12-18 trunk/doc/msc-proposal/msc-nd.tex
proposa: ND

@@
 In order to receive RAs from routers, a node joins the all-nodes multicast address and, if the node is a router, the all-routers multicast address.
-Once the interfaces have been initialized, a node listens for Router Advertisements (RA) from Edge Routers or LoWPAN Routers, or broadcasts a Router Solicitation (RS).
+Once the interfaces have been initialized, a node listens for Router Advertisements (RA), or broadcasts a Router Solicitation (RS).
 Upon receipt of the RA, the node forms an optimistic global unique address with stateless address autoconfiguration and chooses one or more default routers (routers are present only in a Route Over configuration).
@@
 The constructed global address is tentative as long as the binding is not confirmed by the ER.
-To accomplish this, the node performs initial registration with the ER by sending a unicast Node Registration (NR) message with a list of all addresses it wants to register.
+To accomplish this, the node performs initial registration with a link-local ER  or LoWPAN Router by sending a unicast Node Registration (NR) message with a list of all addresses it wants to register.
+Registering directly with an ER is preferred, although all LoWPAN Routers have the ability to relay NR/NC messages on behalf of a node to the ER.
 The ER replies with a Node Confirmation(NC), which includes the set of addresses confirmed to be bound to the Whiteboard of the ER.
@@
 
-This information about link-local addresses is collected during the node registration process.
-Nodes store the information about router link-local addresses in their default router list, while routers keep the information about nodes in their binding tables.
+Next-hop determination assumes destinations are non-local and forwarded to a default router except for link-local scope addresses.
+The information about link-local addresses is collected during the node registration process.
+Nodes store router link-local addresses in their default router list, whereas routers resolve the link-layer addresses of the attached nodes by looking up on the binding table.
 
-\section{Message Formats}
+\section{Message Types}
 
-6LoWPAN-ND makes use of Router Solicitation (RS)/Router Advertisement (RA) message exchanges similar to classic ND.
-In 6LoWPAN-ND RA messages may carry additional options for context dissemination and are reduced in size.
-In addition to RS and RA messages, 6LoWPAN-ND defines two new ICMP packet types: Node Registration (NR), which is sent by a node to an Edge Router to register a binding for an IPv6 address in the Whiteboard, and Node Confirmation (NC) by which an Edge Router replies to the registering node.
+6LoWPAN-ND makes use of Router Solicitation (RS) and Router Advertisement (RA) message exchanges in a similar way to classic ND.
+However, 6LoWPAN-ND RA messages may carry additional options for context dissemination and are reduced in size.
+In addition to RS and RA messages, 6LoWPAN-ND defines two new ICMP packet types, Node Registration (NR) and Node Confirmation (NC), used by a node to register with an ER, and by the ER to confirm the
+binding.
+
+\subsection{Router Solicitation Message}
+The format of RS message for 6LoWPAN is identical to the classic RS message \cite{rfc4861}.
+If a node has only the addresses that are not yet confirmed by an ER, then the IPv6 unspecified address is used as the IPv6 source address.
+The Source Link-Layer Address Option \cite{rfc4861} used in IPv6 ND is not included in the RSs and the Owner Interface Identifier Option, described in Section \ref{sec:nd.oiio}, is used instead.
+
+\subsection{Router Advertisement Message}
+The format of RA message for 6LoWPAN is identical to the classic \cite{rfc4861} RA message.
+However, the use of a 2-bit  Default Router Preference (Prf) flag \cite{rfc4191} is defined in the context of 6LoWPAN.
+LoWPAN Edge Routers set Prf to $01$ indicating their high preference, whereas LoWPAN Routers with and without ER availability use the values $00$ and $01$ respectively for normal and low preference.
+RAs are sent either to all-nodes multicast or to a link-local unicast address as a response to an RS.
+RA messages use additional 6LoWPAN Information and 6LoWPAN Summary Options described in Sections \ref{nd.option.info} and \ref{nd.option.summary} respectively.
+
+\subsection{Neighbor Solicitation/Advertisement Message}
+Neighbor Solicitation(NS) and Neighbor Advertisement(NA) messages are employed only between ERs on the backbone link in Extended LoWPAN.
+Extended LoWPAN (shown in Figure \ref{fig:ext.lowpan})) allows to form a single subnet out of multiple LoWPANs interconnected by a backbone link via Edge Routers.
+The format of NS/NA messages is identical to the classic NS/NA messages \cite{rfc4861} with additional use of Owner Interface Identifier Option, described in Section \ref{sec:nd.oiio}.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+               +-----+                 +-----+
+               |     | Router          |     | Host
+               |     |                 |     |
+               +-----+                 +-----+
+                  |                       |
+                  |     Backbone link     |
+            +--------------------+------------------+
+            |                    |                  |
+         +-----+             +-----+             +-----+
+         |     | Edge        |     | Edge        |     | Edge
+         |     | Router      |     | Router      |     | Router
+         +-----+             +-----+             +-----+
+            o         o       o   o  o      o        o o
+        o o   o  o  o  o  o o   o  o  o  o  o   o  o  o  o
+       o  o o  o o   o    o   o  o  o  o     o   o  o  o o
+       o   o  o  o     o    o    o  o     o      o  o   o
+         o   o o     o          o  o      o    o       o
+\end{verbatim}
+\end{mylisting}
+\caption{Extended LoWPAN}\label{fig:ext.lowpan}
+\end{figure}
+
+\subsection{Node Registration/Confirmation Message}
+In order to register for the first time, the node sends an NR message to the link-local unicast address of a local ER or LoWPAN Router.
+Since the node still has only an optimistic address, not yet confirmed by an Edge Router, the IPv6 unspecified address is used as the source address of the NR message.
+To renew bindings the subsequent NR messages use the link-local IPv6 address of the sender as the source address.
+
+The message format for NR/NC messages is shown in Figure \ref{fig:nr.nc.format}.
+The values for the Type field will be defined in the future to distinguish Node Registration and Node Confirmation messages.
+The Status field is used only in NC messages and specifies the result of the registation procedure.
+A 4-bit unsigned Code provides additional details for the message, e.g.
+it may indicate whether the NR/NC is sent directly or relayed by a router.
+The Checksum field carries the ICMP checksum.
+A unique Transaction ID (TID) is used to match replies and a NC TID corresponds to the NR TID.
+Its value is incremented upon each new registration.
+A 1-bit Primary flag (P) indicates whether the ER is primary and may represent the node on the backbone in a Extended LoWPAN.
+A Router flag indicates the role of the node sending the NR message (0 is used by hosts and 1 by routers).
+The Reserved field is unused and initialized to zero.
+The Binding Lifetime is the amount of time in minutes remaining before the binding of the owner interface identifier expires.
+The Advertising Lifetime field indicates the amount of time in units of 10 seconds the node will advertise itself to its local router using a NR local refresh.
+The Owner Nonce field is a 32-bit value generated randomly by the node upon booting.
+And the Owner Interface Identifier (OII) is a globally unique identifier for the requesting host's interface.
+
+A NR message includes a 6LoWPAN  Address Option, described in Section \ref{nd.option.address}, for each address the host wants to bind for the interface, and a NC message may carry a 6LoWPAN Information Option, described in Section \ref{nd.option.info}, containing information about the prefixes of the LoWPAN along with other context information.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+    0                   1                   2                   3
+    0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |     Type      |Status | Code  |           Checksum            |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |      TID      |P|R|                Reserved                   |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |        Binding Lifetime       |     Advertising Interval      |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |                         Owner Nonce                           |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |                                                               |
+   +                  Owner Interface Identifier                   +
+   |                                                               |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |   Registration option(s)...
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Node Registration/Confirmation message format}\label{fig:nr.nc.format}
+\end{figure}
+
+\subsection{Message Options}
+\subsubsection{6LoWPAN Address Option}\label{nd.option.address}
+
+\subsubsection{6LoWPAN Information Option}\label{nd.option.info}
+
+\subsubsection{6LoWPAN Summary Option}\label{nd.option.summary}
+
+\subsubsection{Owner Interface Identifier Option}\label{sec:nd.oiio}

### 2009-12-18 trunk/doc/msc-proposal/msc-proposal.tex
proposa: ND

@@
 %%
-\chapter{Transmission of IPv6 Packets over IEEE 802.15.4}\label{ch:6lowpan}
+\chapter{Transmission of IPv6 Packets over 802.15.4}\label{ch:6lowpan}
 A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
@@
 \nocite{rfc3775}
+\nocite{rfc4191}
 \nocite{rfc4291} 

### 2009-12-18 trunk/doc/msc-proposal/msc-nd.tex
proposal: ND options

@@
 
-A NR message includes a 6LoWPAN  Address Option, described in Section \ref{nd.option.address}, for each address the host wants to bind for the interface, and a NC message may carry a 6LoWPAN Information Option, described in Section \ref{nd.option.info}, containing information about the prefixes of the LoWPAN along with other context information.
+A NR message includes a 6LoWPAN  Address Option, described in Section \ref{nd.option.address}, for each address the host wants to bind for the interface, and a NC message may carry a 6LoWPAN Information Option, described in Section \ref{nd.option.info}.
 
@@
 \subsection{Message Options}
+The 6LoWPAN-ND introduces four new message options described in this section.
+
 \subsubsection{6LoWPAN Address Option}\label{nd.option.address}
+The 6LoWPAN Address Option (6AO) is used in NR and NC messages to indicate the address which a node wants to register with an ER and to get the result of registration back.
+A NR/NC message can include multiple Address Options.
+The fields of the option allow to indicate the success or failure of the binding in an NC, as well as to compress the carried IPv6 address to different extent.
 
 \subsubsection{6LoWPAN Information Option}\label{nd.option.info}
+The 6LoWPAN Information Option (6IO) is used to carry prefix information.
+A RA message includes one Information Option for each listed prefix.
+The option also allows to disseminate contexts used in 6LoWPAN address compression and identified by a Context Identifier (CID).
 
 \subsubsection{6LoWPAN Summary Option}\label{nd.option.summary}
+The 6LoWPAN Summary Option (6SO) associates the current prefix options with a sequence number.
+It allows prefix options themselves to be sent only periodically in unsolicited RAs, reducing the message size.
+When the sequence number of this option has a new value, then the prefix information has likely changed, and, in this case, a node requests the prefix information with an RS.
+An RA sent in response to a unicast RS always includes the full set of prefix information.
 
 \subsubsection{Owner Interface Identifier Option}\label{sec:nd.oiio}
+The Owner Interface Identifier Option (OIIO) is used with classic NS and NA messages between ERs over a backbone link to identify bindings and whiteboard entries.

### 2009-12-19 trunk/doc/msc-proposal/msc-routing.tex
proposal: ND

--- 2009-12-16 trunk/doc/msc-proposal/msc-routing.tex
@@
 Neither the IEEE 802.15.4 \cite{ieee802.15.4} standard nor the 6LoWPAN format specification \cite{rfc4944} define how mesh topologies could be obtained and maintained.
-In 6LoWPAN routing can be perforem either in the IP-layer, using a Route Over approach, or in the adaptation layer described in Section \ref{subsec:mesh.header}, using the Mesh Under approach.
+In 6LoWPAN routing can be performed either in the IP-layer, using a route over approach, or in the adaptation layer, described in Section \ref{subsec:mesh.header}, using the mesh under approach.
+The mesh under approach performs the multi-hop communication below the IP link and therefore a 6LoWPAN is seen as a single IP link.
+
+A 6LoWPAN network comprises nodes which are either hosts or routers.
+A host is a node that only sources or sinks IPv6 datagrams, whereas a router forwards datagrams between arbitrary source-destination pairs.
+Routers are present only in a route over configuration, where the network is composed of overlapping link-local scopes, to overcome non-transitive nature of links.

### 2009-12-19 doc/msc-proposal/msc-nd.tex
proposal: ND

--- 2009-12-16 trunk/doc/msc-proposal/msc-nd.tex
@@
 The Neighbor Discovery (ND) protocol for IPv6 \cite{rfc4861} provides for basic bootstrapping and network operation.
-Nodes (hosts and routers) use ND to determine the link-layer addresses for neighbors known to reside on attached links.
-Hosts also use ND to find neighboring routers that are willing to forward packets.
-In the previous Chapter \ref{ch:6lowpan} a mechanism which allows to carry IPv6 packets over IEEE 802.15.4 networks with the help of an adaptation layer has been described.
+Nodes use ND to determine the link-layer addresses for neighbors known to reside on attached links, as well as to find neighboring routers that are willing to forward packets.
 However, the standard IPv6 Neighbor Discovery has several problems when using it with 6LoWPANs.
 IPv6 ND heavily uses multicast capabilities, whereas it is very expensive and not desirable in a low-power, lossy wireless network.
+Moreover, in a route over configuration LoWPAN links are non-transitive and composed of a large amount of overlapping radio ranges, but the classic ND was not designed for such links.
 Therefore, the classic ND is not suitable for 6LoWPANs.
+
 The Internet Draft \cite{draft-nd-07} specifies an optimized neighbor discovery mechanism sufficient for LoWPAN operation.
-This chapter describes the 6LoWPAN ND protocol.
+The 6LoWPAN Neighbor Discovery (6LoWPAN-ND) introduces a node registration mechanism optimizing the node-router interface, which requires no flooding and reduces link-local multicast frequency.
+The concept of a LoWPAN Whiteboard located at Edge Routers (routers that connect a LoWPAN to another IP network) is introduced, which allows for Duplicate Address Detection for the entire LoWPAN.
+The solution supports both Mesh Under and route over configurations for multihop forwarding.
+This chapter describes the specified 6LoWPAN-ND protocol.
 
-\section{Basic features}\label{nd.basic}
-6LoWPAN Neighbor Discovery (6LoWPAN-ND) introduces a node registration mechanism optimizing the node-router interface.
-This mechanism requires no flooding and reduces link-local multicast frequency.
-Nodes in the LoWPAN register with Routers and Edge Routers (IPv6 routers that interconnect the LoWPAN to another IP network), creating state about nodes attached to that router (binding table) and about all IPv6 addresses in the LoWPAN.
-A list of all IPv6 addresses in the LoWPAN are stored within a conceptual data structure, Whiteboard, located at Edge Routers (ERs).
-The Whiteboard makes use of soft bindings that contain an Owner Interface Identifier, Owner Nonce, IPv6 address and a remaining lifetime of the binding.
+%All nodes in the LoWPAN register with Routers and Edge Routers, though Routers are present only in a route over configuration.
+Each Router maintains a set of information about nodes that are currently registered through it, called the binding table.
+All IPv6 addresses in the LoWPAN are stored within a conceptual data structure, Whiteboard, located at Edge Routers (ERs).
 Nodes send periodic registration messages in order to maintain their bindings in the Router binding tables and Edge Router Whiteboard.
 
-6LoWPAN-ND makes use of Router Solicitation (RS)/Router Advertisement (RA) message exchanges similar to classic ND.
-In 6LoWPAN-ND RA messages may carry additional options for context dissemination and are reduced in size.
-In addition to RS and RA messages, 6LoWPAN-ND defines two new ICMP packet types: Node Registration (NR), which is sent by a node to an Edge Router to register a binding for an IPv6 address in the Whiteboard, and Node Confirmation (NC) by which an Edge Router replies to the registering node.
-
-\section{Bootstrapping}
+\section{Bootstrapping and Basic Operation}
 Bootstrapping of a LoWPAN node consists of several steps.
 At first, a node is required to autoconfigure at least one address, a link-local address, which is derived from the IEEE 64-bit extended MAC.
+Therefore,  knowledge of the 64-bit address of a node is enough to derive its link-local address and reach it on the same link.
 In order to receive RAs from routers, a node joins the all-nodes multicast address and, if the node is a router, the all-routers multicast address.
-Once the interfaces have been initialized, a node listens for Router Advertisements (RA) from Edge Routers or LoWPAN Routers, or broadcasts a Router Solicitation (RS).
-Upon receipt of the RA, the node forms an optimistic global unique address with stateless address autoconfiguration and chooses one or more default routers.
+Once the interfaces have been initialized, a node listens for Router Advertisements (RA), or broadcasts a Router Solicitation (RS).
+Upon receipt of the RA, the node forms an optimistic global unique address with stateless address autoconfiguration and chooses one or more default routers (routers are present only in a route over configuration).
 
-The constructed global address is tentative or optimistic as long as the binding is not confirmed by the ER.
-To accomplish this, the node performs initial registration with the ER by sending a unicast Node Registration (NR) message.
-The destination address of the NR message is the link-local unicast address of the ER, while the IPv6 unspecified address is used as the source address.
-The NR message includes an Address Option for each address to be registered.
+The constructed global address is tentative as long as the binding is not confirmed by the ER.
+To accomplish this, the node performs initial registration with a link-local ER  or LoWPAN Router by sending a unicast Node Registration (NR) message with a list of all addresses it wants to register.
+Registering directly with an ER is preferred, although all LoWPAN Routers have the ability to relay NR/NC messages on behalf of a node to the ER.
+The ER replies with a Node Confirmation(NC), which includes the set of addresses confirmed to be bound to the Whiteboard of the ER.
+Once the node has received the NC message, it is capable to send packets to any IPv6 address inside or outside the LoWPAN.
+Each router maintains a set of information about nodes that are currently registered through it, called the binding table.
+Nodes send periodic registration messages in order to maintain their bindings in the router binding tables and Edge Router Whiteboard.
+The detection of duplicate addresses (DAD) is also performed as part of the node registration process by the ER using a lookup on the Whiteboard.
 
-The ER replies with a Node Confirmation(NC), which includes the set of addresses confirmed to be bound to the Whiteboard of the ER.  The
-source of the packet is the link-local address of LoWPAN Router and the destination address is the link-local address of the node.
-Once the node has received the NC, it is capable to send packets to any IPv6 address inside or outside the LoWPAN.
-The detection of duplicate addresses (DAD) is performed as part of the node registration process by the ER across the entire LoWPAN using a lookup on the Whiteboard.
+Next-hop determination assumes destinations are non-local and forwarded to a default router except for link-local scope addresses.
+The information about link-local addresses is collected during the node registration process.
+Nodes store router link-local addresses in their default router list, whereas routers resolve the link-layer addresses of the attached nodes by looking up on the binding table.
 
-This information about link-local addresses is collected during the node registration process.
-Nodes store the information about router link-local addresses in their default router list, while routers keep the information about nodes in their binding tables.
+\section{Message Types}
+
+6LoWPAN-ND makes use of Router Solicitation (RS) and Router Advertisement (RA) message exchanges in a similar way to classic ND.
+However, 6LoWPAN-ND RA messages may carry additional options for context dissemination and are reduced in size.
+In addition to RS and RA messages, 6LoWPAN-ND defines two new ICMP packet types, Node Registration (NR) and Node Confirmation (NC), used by a node to register with an ER, and by the ER to confirm the
+binding.
+
+\subsection{Router Solicitation Message}
+The format of RS message for 6LoWPAN is identical to the classic RS message \cite{rfc4861}.
+If a node has only the addresses that are not yet confirmed by an ER, then the IPv6 unspecified address is used as the IPv6 source address.
+The Source Link-Layer Address Option \cite{rfc4861} used in IPv6 ND is not included in the RSs and the Owner Interface Identifier Option, described in Section \ref{sec:nd.oiio}, is used instead.
+
+\subsection{Router Advertisement Message}
+The format of RA message for 6LoWPAN is identical to the classic \cite{rfc4861} RA message.
+However, the use of a 2-bit  Default Router Preference (Prf) flag \cite{rfc4191} is defined in the context of 6LoWPAN.
+LoWPAN Edge Routers set Prf to $01$ indicating their high preference, whereas LoWPAN Routers with and without ER availability use the values $00$ and $01$ respectively for normal and low preference.
+RAs are sent either to all-nodes multicast or to a link-local unicast address as a response to an RS.
+RA messages use additional 6LoWPAN Information and 6LoWPAN Summary Options described in Sections \ref{nd.option.info} and \ref{nd.option.summary} respectively.
+
+\subsection{Neighbor Solicitation/Advertisement Message}
+Neighbor Solicitation(NS) and Neighbor Advertisement(NA) messages are employed only between ERs on the backbone link in Extended LoWPAN.
+Extended LoWPAN (shown in Figure \ref{fig:ext.lowpan})) allows to form a single subnet out of multiple LoWPANs interconnected by a backbone link via Edge Routers.
+The format of NS/NA messages is identical to the classic NS/NA messages \cite{rfc4861} with additional use of Owner Interface Identifier Option, described in Section \ref{sec:nd.oiio}.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+               +-----+                 +-----+
+               |     | Router          |     | Host
+               |     |                 |     |
+               +-----+                 +-----+
+                  |                       |
+                  |     Backbone link     |
+            +--------------------+------------------+
+            |                    |                  |
+         +-----+             +-----+             +-----+
+         |     | Edge        |     | Edge        |     | Edge
+         |     | Router      |     | Router      |     | Router
+         +-----+             +-----+             +-----+
+            o         o       o   o  o      o        o o
+        o o   o  o  o  o  o o   o  o  o  o  o   o  o  o  o
+       o  o o  o o   o    o   o  o  o  o     o   o  o  o o
+       o   o  o  o     o    o    o  o     o      o  o   o
+         o   o o     o          o  o      o    o       o
+\end{verbatim}
+\end{mylisting}
+\caption{Extended LoWPAN}\label{fig:ext.lowpan}
+\end{figure}
+
+\subsection{Node Registration/Confirmation Message}
+In order to register for the first time, the node sends an NR message to the link-local unicast address of a local ER or LoWPAN Router.
+Since the node still has only an optimistic address, not yet confirmed by an Edge Router, the IPv6 unspecified address is used as the source address of the NR message.
+To renew bindings the subsequent NR messages use the link-local IPv6 address of the sender as the source address.
+
+The message format for NR/NC messages is shown in Figure \ref{fig:nr.nc.format}.
+The values for the Type field will be defined in the future to distinguish Node Registration and Node Confirmation messages.
+The Status field is used only in NC messages and specifies the result of the registration procedure.
+A 4-bit unsigned Code provides additional details for the message, e.g.
+it may indicate whether the NR/NC is sent directly or relayed by a router.
+The Checksum field carries the ICMP checksum.
+A unique Transaction ID (TID) is used to match replies and a NC TID corresponds to the NR TID.
+Its value is incremented upon each new registration.
+A 1-bit Primary flag (P) indicates whether the ER is primary and may represent the node on the backbone in a Extended LoWPAN.
+A Router flag indicates the role of the node sending the NR message (0 is used by hosts and 1 by routers).
+The Reserved field is unused and initialized to zero.
+The Binding Lifetime is the amount of time in minutes remaining before the binding of the owner interface identifier expires.
+The Advertising Lifetime field indicates the amount of time in units of 10 seconds the node will advertise itself to its local router using a NR local refresh.
+The Owner Nonce field is a 32-bit value generated randomly by the node upon booting.
+And the Owner Interface Identifier (OII) is a globally unique identifier for the requesting host's interface.
+
+A NR message includes a 6LoWPAN  Address Option, described in Section \ref{nd.option.address}, for each address the host wants to bind for the interface, and a NC message may carry a 6LoWPAN Information Option, described in Section \ref{nd.option.info}.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+    0                   1                   2                   3
+    0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |     Type      |Status | Code  |           Checksum            |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |      TID      |P|R|                Reserved                   |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |        Binding Lifetime       |     Advertising Interval      |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |                         Owner Nonce                           |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |                                                               |
+   +                  Owner Interface Identifier                   +
+   |                                                               |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |   Registration option(s)...
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+
+\end{verbatim}
+\end{mylisting}
+\caption{Node Registration/Confirmation message format}\label{fig:nr.nc.format}
+\end{figure}
+
+\subsection{Message Options}
+The 6LoWPAN-ND introduces four new message options described in this section.
+
+\subsubsection{6LoWPAN Address Option}\label{nd.option.address}
+The 6LoWPAN Address Option (6AO) is used in NR and NC messages to indicate the address which a node wants to register with an ER and to get the result of registration back.
+A NR/NC message can include multiple Address Options.
+The fields of the option allow to indicate the success or failure of the binding in an NC, as well as to compress the carried IPv6 address to different extent.
+
+\subsubsection{6LoWPAN Information Option}\label{nd.option.info}
+The 6LoWPAN Information Option (6IO) is used to carry prefix information.
+A RA message includes one Information Option for each listed prefix.
+The option also allows to disseminate contexts used in 6LoWPAN address compression and identified by a Context Identifier (CID).
+
+\subsubsection{6LoWPAN Summary Option}\label{nd.option.summary}
+The 6LoWPAN Summary Option (6SO) associates the current prefix options with a sequence number.
+It allows prefix options themselves to be sent only periodically in unsolicited RAs, reducing the message size.
+When the sequence number of this option has a new value, then the prefix information has likely changed, and, in this case, a node requests the prefix information with an RS.
+An RA sent in response to a unicast RS always includes the full set of prefix information.
+
+\subsubsection{Owner Interface Identifier Option}\label{sec:nd.oiio}
+The Owner Interface Identifier Option (OIIO) is used with classic NS and NA messages between ERs over a backbone link to identify the entries within a binding table or whiteboard.
+
+\section{Conceptual Data Structures }
+\subsection{LoWPAN Node}
+Every node in a LoWPAN keeps a list of prefixes advertised in Router Advertisements.
+The entries of the list are associated with the sequence numbers last advertised in the 6LoWPAN Summary Option, which allows to detect the changes in prefix information.
+In order to support stateful address compression, a node holds a list of context.
+As in the case of prefixes, each context entry is associated with the 6LoWPAN Summary Option sequence number.
+The list of Edge Routers the node is registered with an associated timeout and primary flag is stored within the Edge Router List.
+The Default Router List data structure contains the routers to which packets may be forwarded from the node.
+
+In addition to the mentioned data structures, a node also keeps a set of conceptual variables for each interface, such as the default hop limit and the binding lifetime.
+
+\subsection{LoWPAN Router}
+LoWPAN Routers are used only in a route over configuration.
+In addition to the classic ND conceptual variables defined in Section 6.2.1 of \cite{rfc4861}, routers maintain binding tables with the information about nodes that are currently registered through them.
+A binding table entry  contains the registered node's OII, link-local IPv6 address and the advertisement interval from the last NR.
+
+\subsection{LoWPAN Edge Router}
+Edge Routers implement LoWPAN Router features and extend them with Whiteboard registration for LoWPAN Nodes within their subnets.
+For every registered node an ER maintains the IPv6 address, Owner Interface Identifier, .
+There is no link-layer information stored in the Whiteboard.
+In an Extended LoWPAN the full registry of all the LoWPAN Nodes in a subnet is distributed between the Edge Routers.

### 2009-12-20 doc/msc-proposal/msc-nd.tex
proposal: routing requirements

@@
 
-In addition to the mentioned data structures, a node also keeps a set of conceptual variables for each interface, such as the default hop limit and the binding lifetime.
+In addition to the mentioned data structures, a node also keeps a set of conceptual variables for each interface, e.g. the default hop limit and the binding lifetime.
 
@@
 Edge Routers implement LoWPAN Router features and extend them with Whiteboard registration for LoWPAN Nodes within their subnets.
-For every registered node an ER maintains the IPv6 address, Owner Interface Identifier, .
+An ER maintains information about every registered node, e.g.
+the IPv6 address and Owner Interface Identifier of the node.
 There is no link-layer information stored in the Whiteboard.

### 2009-12-20 doc/msc-proposal/msc-routing.tex
proposal: routing requirements

@@
 Neither the IEEE 802.15.4 \cite{ieee802.15.4} standard nor the 6LoWPAN format specification \cite{rfc4944} define how mesh topologies could be obtained and maintained.
-In 6LoWPAN routing can be performed either in the IP-layer, using a Route Over approach, or in the adaptation layer, described in Section \ref{subsec:mesh.header}, using the Mesh Under approach.
-The Mesh Under approach performs the multi-hop communication below the IP link and therefore a 6LoWPAN is seen as a single IP link.
+In 6LoWPAN routing can be performed either in the IP-layer, using a route over approach, or in the adaptation layer, described in Section \ref{subsec:mesh.header}, using the mesh under approach.
+The mesh under configuration performs the multi-hop routing below the IP link and therefore the characteristics of IEEE 802.15.4 directly affect the 6LoWPAN routing mechanisms.
+In this approach a 6LoWPAN is seen as a single IP link and the IPv6 link-local scope covers all nodes in the LoWPAN.
+In the route over configuration intermediate nodes become LoWPAN Routers and perform standard Layer 3 (IP) routing.
+Therefore, in this case the link-local scope includes the set of nodes within symmetric radio range of a node and routing can be accomplished over various types of interconnected links.
 
-A 6LoWPAN network comprises nodes which are either hosts or routers.
-A host is a node that only sources or sinks IPv6 datagrams, whereas a router forwards datagrams between arbitrary source-destination pairs.
-Routers are present only in a Route Over configuration, where the network is composed of overlapping link-local scopes, to overcome non-transitive nature of links.
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+  +-----------------------------+    +-----------------------------+
+  |  Application Layer          |    |  Application Layer          |
+  +-----------------------------+    +-----------------------------+
+  |  Transport Layer (TCP/UDP)  |    |  Transport Layer (TCP/UDP)  |
+  +-----------------------------+    +-----------------------------+
+  |  Network Layer (IPv6)       |    |  Network       +---------+  |
+  +-----------------------------+    |  Layer         | Routing |  |
+  |  6LoWPAN       +---------+  |    |  (IPv6)        +---------+  |
+  |  Adaptation    | Routing*|  |    +-----------------------------+
+  |  Layer         +---------+  |    |  6LoWPAN Adaptation Layer   |
+  +-----------------------------+    +-----------------------------+
+  |  IEEE 802.15.4 (MAC)        |    |  IEEE 802.15.4 (MAC)        |
+  +-----------------------------+    +-----------------------------+
+  |  IEEE 802.15.4 (PHY)        |    |  IEEE 802.15.4 (PHY)        |
+  +-----------------------------+    +-----------------------------+
+\end{verbatim}
+\end{mylisting}
+\caption{Mesh Under (left) and Route Over routing (right)}\label{fig:routing}
+\end{figure}
+
+This chapter discusses a list of requirements for 6LoWPAN routing \cite{draft-routing-04} in Section \ref{sec:rout.req}.
+Section \ref{sec:rout.rpl} describes the Internet Draft \cite{draft-rpl-04} which specifies the IPv6 Routing Protocol for Low power and Lossy networks (LLNs).
+
+\section{6LoWPAN Routing Requirements}\label{sec:rout.req}
+A LoWPAN has to support multiple device types and roles such as host nodes drawing their power from primary batteries, mains-powered nodes and power-affluent gateways.
+Battery-operated devices need to last from several months to a few years with a single AA battery.
+Therefore 6LoWPAN routing protocols have to cause minimal power consumption by the efficient use of control packets, minimization of expensive IP multicast which causes link broadcast to the entire LoWPAN, and by the efficient routing of data packets.
+Control messages have to fit a single IEEE 802.15.4 frame size in order to avoid packet fragmentation and the overhead for reassembly.
+The design of 6LoWPAN routing protocols should be scalable to support from a few nodes to millions of them.
+
+6LoWPAN devices are unreliable due to limited system capabilities and unpredictable environment where they can be deployed.
+6LoWPAN routing protocols have to be robust to dynamic loss caused by link failure or device unavailability.
+Moreover, some of the links may be asymmetric, when the probability of successful transmission between two nodes is significantly higher in one direction than in the other one.
+6LoWPAN routing protocols have to be designed to correctly operate in the presence of such links.
+In addition, latency and successful end-to-end packet delivery ratio requirements of applications must be taken into account.
+
+6LoWPAN devices have small memory sizes, therefore 6LoWPAN routing protocols require implementation with small code size and low routing state to fit the typical 6LoWPAN node capacity.
+The code size is limited to available flash memory size, and the routing table is bounded by RAM size.
+
+\section{Existing Routing Protocols}
+
+\section{Routing Protocol for Low power and Lossy networks}\label{sec:rout.rpl}
+The ROLL WG is now working on route over approaches for Low power and Lossy Networks (LLNs), not specifically for 6LoWPAN. 
+\subsection{Routing Metrics}
+Historically, IGP such as OSPF ([RFC2328]) and IS-IS ([RFC1195]) have used quantitative static link metric

### 2009-12-20 trunk/doc/msc-proposal/msc-proposal.tex
proposal: routing requirements

@@
 \chapter{Transmission of IPv6 Packets over 802.15.4}\label{ch:6lowpan}
-A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
-The IEEE 802.15.4-2003 standard specifies node-to-node frame delivery between the devices which are within reachable distance from each other.
+As described in Chapter \ref{ch:ieee802.15.4}, the IEEE 802.15.4-2003 standard specifies the physical layer and media access control for low-rate wireless personal area network providing node-to-node frame delivery between the devices within reachable distance from each other.
 On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be defined that would provide end-to-end packet delivery including routing through intermediate hosts.
@@
 \nocite{draft-nd-07}
+\nocite{draft-routing-04}
+\nocite{draft-protocols-07}
+\nocite{draft-rpl-04}
+\nocite{draft-routing-metrics-04}
 

### 2009-12-20 trunk/doc/msc-proposal/msc-routing.tex
proposal: routing

--- 2009-12-16 trunk/doc/msc-proposal/msc-routing.tex
@@
 Neither the IEEE 802.15.4 \cite{ieee802.15.4} standard nor the 6LoWPAN format specification \cite{rfc4944} define how mesh topologies could be obtained and maintained.
-In 6LoWPAN routing can be perforem either in the IP-layer, using a Route Over approach, or in the adaptation layer described in Section \ref{subsec:mesh.header}, using the Mesh Under approach.
+In 6LoWPAN routing can be performed either in the IP-layer, using a route over approach, or in the adaptation layer, described in Section \ref{subsec:mesh.header}, using the mesh under approach.
+The mesh under configuration performs the multi-hop routing below the IP link and therefore the characteristics of IEEE 802.15.4 directly affect the 6LoWPAN routing mechanisms.
+In this approach a 6LoWPAN is seen as a single IP link and the IPv6 link-local scope covers all nodes in the LoWPAN.
+In the route over configuration intermediate nodes become LoWPAN Routers and perform standard Layer 3 (IP) routing.
+Therefore, in this case, the link-local scope includes the set of nodes within symmetric radio range of a node and routing can be accomplished over various types of interconnected links.
+
+The IETF Routing Over Low power and Lossy networks Working Group  (ROLL WG) is focused on routing issues for Low power and Lossy networks (LLNs)  composed of many embedded devices with limited power, memory, and processing resources interconnected by a variety of links, such as IEEE 802.15.4 or Low Power WiFi.
+The aim of the group is to provide an IPv6 architectural framework for routing and path selection in LLNs.
+So far several Internet Drafts have been published, such as Overview of Existing Routing Protocols for Low Power and Lossy Networks \cite{draft-protocols-07},  IPv6 Routing Protocol for Low power and Lossy Networks \cite{draft-rpl-04}, and Routing Metrics used for Path Calculation in Low Power and Lossy Networks \cite{draft-routing-metrics-04}.
+
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+  +-----------------------------+    +-----------------------------+
+  |  Application Layer          |    |  Application Layer          |
+  +-----------------------------+    +-----------------------------+
+  |  Transport Layer (TCP/UDP)  |    |  Transport Layer (TCP/UDP)  |
+  +-----------------------------+    +-----------------------------+
+  |  Network Layer (IPv6)       |    |  Network       +---------+  |
+  +-----------------------------+    |  Layer         | Routing |  |
+  |  6LoWPAN       +---------+  |    |  (IPv6)        +---------+  |
+  |  Adaptation    | Routing*|  |    +-----------------------------+
+  |  Layer         +---------+  |    |  6LoWPAN Adaptation Layer   |
+  +-----------------------------+    +-----------------------------+
+  |  IEEE 802.15.4 (MAC)        |    |  IEEE 802.15.4 (MAC)        |
+  +-----------------------------+    +-----------------------------+
+  |  IEEE 802.15.4 (PHY)        |    |  IEEE 802.15.4 (PHY)        |
+  +-----------------------------+    +-----------------------------+
+\end{verbatim}
+\end{mylisting}
+\caption{Mesh Under (left) and Route Over routing (right)}\label{fig:routing}
+\end{figure}
+
+Further in this chapter, the list of requirements for 6LoWPAN routing defined in the Internet Draft \cite{draft-routing-04} is discussed in Section \ref{sec:rout.req}.
+The evaluation of existing  routing protocol for LLN scenarious is presented in Section \ref{sec:rout.protocols}.
+And,  finally, Section \ref{sec:rout.rpl} describes the Internet Draft \cite{draft-rpl-04} which specifies the IPv6 Routing Protocol for LLNs.
+
+\section{6LoWPAN Routing Requirements}\label{sec:rout.req}
+A LoWPAN has to support multiple device types and roles such as host nodes drawing their power from primary batteries, mains-powered nodes and power-affluent gateways.
+Battery-operated devices need to last from several months to a few years with a single AA battery.
+Therefore 6LoWPAN routing protocols have to cause minimal power consumption by the efficient use of control packets, minimization of expensive IP multicast which causes link broadcast to the entire LoWPAN, and by the efficient routing of data packets.
+Control messages have to fit a single IEEE 802.15.4 frame size in order to avoid packet fragmentation and the overhead for reassembly.
+The design of 6LoWPAN routing protocols should be scalable to support from a few nodes to millions of them.
+
+6LoWPAN devices are unreliable due to limited system capabilities and unpredictable environment where they can be deployed.
+6LoWPAN routing protocols have to be robust to dynamic loss caused by link failure or device unavailability.
+Moreover, some of the links may be asymmetric, when the probability of successful transmission between two nodes is significantly higher in one direction than in the other one.
+6LoWPAN routing protocols have to be designed to correctly operate in the presence of such links.
+In addition, latency and successful end-to-end packet delivery ratio requirements of applications must be taken into account.
+
+6LoWPAN devices have small memory sizes, therefore 6LoWPAN routing protocols require implementation with small code size and low routing state to fit the typical 6LoWPAN node capacity.
+The code size is limited to available flash memory size, and the routing table is bounded by RAM size.
+
+\section{Existing Routing Protocols}\label{sec:rout.protocols}
+
+\section{Routing Protocol for Low power and Lossy networks}\label{sec:rout.rpl}
+
+\subsection{Routing Metrics}
+Historically, IGP such as OSPF ([RFC2328]) and IS-IS ([RFC1195]) have used quantitative static link metric

### 2010-01-12 doc/msc-proposal/msc-proposal.tex
notepad

@@
 
-\large Supervisor: J\"urgen Shoenw\"alder\\[0.5cm]
+\large Supervisor: J\"urgen Sh\"onw\"alder\\[0.5cm]
 {\large \today}
@@
 \tableofcontents
-\listoffigures 
+%\listoffigures
 \newpage
@@
 \chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
-The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks.
-The standard intends to offer the fundamental lower network layers of a type of wireless personal area network which focuses on low-cost, low-speed communication between tiny devices (motes).
-Such networks are typically limited to a personal operating space (POS) of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.
+The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks providing nodo-to-node frame dilivery between devices within reachable distance from each other.
+Such networks are typically limited to a personal operating space of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.
 
@@
 An 802.15.4 network may operate in either the star or the peer-to-peer topology.
-In the star topology devices communicate with a single central                           a personal area network (PAN) coordinator.
+In the star topology devices communicate with a single central                            personal area network (PAN) coordinator.
 In the peer-to-peer topology a device can communicate with any other device as long as they are in range of one another.
@@
 \chapter{Transmission of IPv6 Packets over 802.15.4}\label{ch:6lowpan}
-A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard.
-The IEEE 802.15.4-2003 standard specifies node-to-node frame delivery between the devices which are within reachable distance from each other.
-On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be defined that would provide end-to-end packet delivery including routing through intermediate hosts.
+As described in Chapter \ref{ch:ieee802.15.4}, the IEEE 802.15.4-2003 standard specifies the physical layer and media access control for low-rate wireless personal area network providing node-to-node frame delivery between the devices within reachable distance from each other.
+On top of IEEE 802.15.4, a network layer (layer 3 of the seven-layer OSI model) has to be defined that provides end-to-end packet delivery including routing through intermediate nodes.
 
@@
 
-6lowpan (Internet Protocol Version 6 (IPv6) over Low power Wireless Personal Area Networks) is a working group within the IETF concerned with the specification of mechanisms to allow IPv6 packets to be sent to and received from over IEEE 802.15.4 based networks.
-The working group has already completed two documents: \cite{rfc4919} that documents and discusses the problem statement and \cite{rfc4944} that defines the format for the adaptation between IPv6 and 802.15.4.
+The 6lowpan (Internet Protocol Version 6 (IPv6) over Low power Wireless Personal Area Networks) working group of the IETF is concerned with the specification of mechanisms to allow IPv6 packet transmission over IEEE 802.15.4 based networks.
+The working group has already completed two documents: "IPv6 over Low-Power Wireless Personal Area Networks (6LoWPANs): Overview, Assumptions, Problem Statement, and Goals" \cite{rfc4919} that documents and discusses the problem statement and "Transmission of IPv6 Packets over IEEE 802.15.4 Networks" \cite{rfc4944} that defines the format for the adaptation between IPv6 and 802.15.4.
 The latter document describes the frame format for transmission of IPv6 packets and the method of forming IPv6 link-local addresses and statelessly autoconfigured addresses on IEEE 802.15.4 networks.
@@
 
-6LoWPAN supports stateless address autoconfiguration, which allows to obtain the Interface Identifier \cite{rfc4291} for an IEEE 802.15.4 interface  based on the EUI-64 identifier \cite{eui64} \cite{rfc2464} assigned to the IEEE802.15.4 device.
+6LoWPAN supports stateless address autoconfiguration, which allows to obtain the interface identifier \cite{rfc4291} for an IEEE 802.15.4 interface  based on the EUI-64 identifier \cite{eui64} \cite{rfc2464} assigned to the IEEE 802.15.4 device.
 Even though all 802.15.4 devices have an EUI-64 address, 16-bit short addresses also can be used for address autoconfiguration.
-In this case, a pseudo 48-bit address is formed by concatenating 16 zero bits to the 16-bit PAN ID, the resulting 32 bits are concatenated with the 16-bit short address and the interface identifier is formed from this 48-bit address:
+In this case, a pseudo 48-bit address is formed by concatenating 16 zero bits to the 16-bit PAN ID, the resulting 32 bits are concatenated with the 16-bit short address.
+The interface identifier is formed from this 48-bit address as defined in the "IPv6 over Ethernet" specification \cite{rfc2464}:
 \begin{center}\texttt{16\_bit\_PAN\_ID:16\_zero\_bits:16\_bit\_short\_address}\end{center}
 
-The IPv6 link-local address for an IEEE 802.15.4 interface is formed by appending the Interface Identifier to the prefix \texttt{FE80::/64}.
+The IPv6 link-local address for an IEEE 802.15.4 interface is formed by appending the interface identifier to the prefix \texttt{FE80::/64}.
 
@@
 As described in Chapter \ref{ch:ieee802.15.4}, the maximum IEEE 802.15.4 frame size is limited to 127 octets.
-Taking into account the maximum frame overhead of 25 octets, 102 octets are left at the media access control layer.
+Taking into account the maximum frame overhead of 25 octets (disabled security), 102 octets are left at the media access control layer.
 Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size.
-Therefor, a fragmentation and reassembly adaptation layer is provided at the layer below IP.
+Therefore, a fragmentation and reassembly adaptation layer is provided at the layer below IP.
 
@@
 In such mesh scenarios, intermediate devices are used as forwarders towards the final destination.
-In order to support mesh routing, in addition to the hop-by-hop source and destination link-layer addresses the originator and final destination addresses need to be known.
+In order to support mesh routing, in addition to the hop-by-hop source and destination link-layer addresses, the originator and final destination addresses need to be known.
 For this purpose a Mesh Addressing header is introduced.
@@
 These headers must appear in the specified order.
-First three of them are optional, whereas the last one is required.
+The first three of them are optional, whereas the last one is required.
 
@@
 
-The first and subsequent fragment header types are specified by the binary sequences \texttt{11000} and \texttt{11100} respectively.
+The first and subsequent fragment header types are identified by the binary sequences \texttt{11000} and \texttt{11100} respectively.
 The  \texttt{datagram\_size} 11-bit field encodes the size of the entire IP packet before link-layer fragmentation.
@@
 \subsection{Broadcast Header}\label{subsec:broad.header}
-Broadcast is an additional mesh routing functionality. For that a broadcast header is defined the format of which is shown in Figure \ref{fig:broad.header}.
+Broadcast is an additional mesh routing functionality. For that a broadcast header is defined, the format of which is shown in Figure \ref{fig:broad.header}.
 \begin{figure}[htp]
@@
 \section{Header Compression}\label{sec:hc}
-As described in Section \ref{sec:adapt.layer}, 81 octets are left in a IEEE 802.15.4 frame for IPv6.
+As described in Section \ref{sec:adapt.layer}, 81 octets are left in an IEEE 802.15.4 frame for IPv6.
 The IPv6 header is 40 octets long which leaves only 41 octets for upper-layer protocols, like UDP.
 The latter uses 8 octets in the header, which leaves 33 octets for application data.
-The fragmentation and reassembly layer, described in Section \ref{sec:adapt.layer}, will also use at least one additional octet, what makes header compression almost being unavoidable.
+The fragmentation and reassembly layer, described in Section \ref{sec:adapt.layer}, will also use at least one additional octet. This makes header compression almost unavoidable.
 
-RFC 4944 \cite{rfc4944} defines stateless header compression mechanisms for IPv6 datagrams (LOWPAN\_HC1 and LOWPAN\_HC2) to reduce the relatively large IPv6 and UDP headers.
-However, LOWPAN\_HC1 and LOWPAN\_HC2 are insufficient for most practical uses of 6LoWPAN networks, since they are effective only for link-local unicast communication, where IPv6 addresses derived directly from IEEE 802.15.4 addresses.
+RFC 4944 \cite{rfc4944} defines a stateless header compression mechanism for IPv6 datagrams (LOWPAN\_HC1 and LOWPAN\_HC2) to reduce the relatively large IPv6 and UDP headers.
+However, LOWPAN\_HC1 and LOWPAN\_HC2 are insufficient for most practical uses of 6LoWPAN networks, since they are effective only for link-local unicast communication, where IPv6 addresses can be derived directly from IEEE 802.15.4 addresses.
 When communicating with devices external to the LoWPAN or in a route-over configuration where IP forwarding occurs within the LoWPAN, LOWPAN\_HC1 carries both IPv6 source and destination addresses in-line.
-Internet Draft document \cite{draft-hc-06} defines an encoding format, LOWPAN\_IPHC, for effective compression of unique local, global, and multicast IPv6 addresses based on shared contexts.
-In addition, it also defines an encoding format, LOWPAN\_NHC, for arbitrary next header compression.
+The internet draft \cite{draft-hc-06} defines an encoding format, LOWPAN\_IPHC, for effective compression of unique local, global, and multicast IPv6 addresses based on shared contexts.
+In addition, it defines an encoding format, LOWPAN\_NHC, for arbitrary next header compression.
 
 \subsection{HC1 Header Encoding}
-The HC1 compression format does not use any shared context and compresses headers relying on information pertaining to the entire link.
-The following IPv6 header values are expected to be common on 6LoWPAN network: Version is IPv6; both IPv6 source and destination addresses are link local;  the IPv6 interface identifiers for the source or destination addresses can be inferred from the layer two source and destination addresses; the packet length can be inferred either from layer two or from the "datagram\_size" field in the fragment header; both the Traffic Class and the Flow Label are zero; and the Next Header is UDP, ICMP or TCP.
+
+Devices in the same 6LoWPAN network share some state.
+This makes it possible to compress headers without explicitly building any compression context state.
+HC1 header compression makes use of this.
+The following IPv6 header values are expected to be common on 6LoWPAN networks: Version is IPv6; both IPv6 source and destination addresses are link local;  the IPv6 interface identifiers for the source or destination addresses can be inferred from the layer two source and destination addresses; the packet length can be inferred either from layer two or from the "datagram\_size" field in the fragment header; both the Traffic Class and the Flow Label are zero; and the Next Header is UDP, ICMP or TCP.
 The Hop Limit (8 bits) field is the only one that always needs to be carried.
@@
 The IPHC encoding format enables effective compression of the IPv6 header relying on information pertaining to the entire 6LoWPAN network.
-The encoding assumes the following common case parameters for 6LoWPAN communication: Version is $6$; Traffic Class and Flow Label are both zero; Payload Length can be inferred from lower layers from either the 6LoWPAN Fragmentation header or the IEEE 802.15.4 header; Hop Limit will be set to a well-known value by the source; addresses assigned to 6LoWPAN interfaces will be formed using the link-local prefix or a single routable prefix assigned to the entire 6LoWPAN network; addresses assigned to 6LoWPAN interfaces are formed with an IID derived directly from either the 64-bit extended or 16-bit short IEEE 802.15.4 addresses.
+The encoding assumes the following common case parameters for 6LoWPAN communication: Version is $6$; Traffic Class and Flow Label are both zero; Payload Length can be inferred from lower layers from either the 6LoWPAN Fragmentation header or the IEEE 802.15.4 header; Hop Limit will be set to a well-known value by the source; addresses assigned to 6LoWPAN interfaces will be formed using the link-local prefix or a single routable prefix assigned to the entire 6LoWPAN network; addresses assigned to 6LoWPAN interfaces are formed with an interface identifier derived directly from either the 64-bit extended or 16-bit short IEEE 802.15.4 addresses.
 The compression mechanism is adapted for these values of the header fields and in such scenario the LOWPAN IPHC can compress the IPv6 header down to two octets (1-octet dispatch, 1-octet LOWPAN\_IPHC encoding) with link-local communication.
@@
 The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$ and a 13 bit IPHC base encoding (shown in Figure \ref{fig:iphc.format}), describing how the IPv6 header is compressed, followed by compressed IPv6 header fields.
-The IPHC encoding may be extended by another octet to support additional contexts.
+The IPHC encoding may be extended by another octet to support additional contexts for stateful address compression described later in this Section.
 
@@
 Stateful address compression expects that a conceptual context is shared between the node that compresses a packet and the node(s) that need to expand it.
-However, the Internet Draft \cite{draft-hc-06} does not specify how the contexts are shared, maintained and what information is contained within a context.
+However, the internet draft \cite{draft-hc-06} does not specify how the contexts are shared, maintained and what information is contained within a context.
 If the Context Identifier (CID) field (see Figure \ref{fig:iphc.format}) is set to 1 then an additional 8-bit Context Identifier Extension field is used and immediately follows the IPHC Base Encoding.
@@
 The latter 3 bits represent an IPv6 Extension Header ID (EID) which identifies the IPv6 Extension Header immediately following the NHC octet.
-Currently, the following values of the EID field have been specified:  $0$ -- IPv6 Hop-by-Hop Options Header \cite{rfc2460}; $1$ -- IPv6 Routing Header \cite{rfc2460};  $2$ -- IPv6 Fragment Header \cite{rfc2460}; $3$ -- IPv6 Destination Options Header \cite{rfc2460}; $4$ -- IPv6 Mobility Header \cite{rfc3775}; $5$ -- Reserved; $6$ -- Reserved; $7$ -- IPv6 Header.
+The defined values of the EID field are shown in Table \ref{table:ipv6.ext.header}.
+\begin{table}[htp]
+\begin{center}
+        \begin{tabular}{|c|l|}
+          \hline
+          EID & Description\\
+          \hline
+          \hline          
+           0 & IPv6 Hop-by-Hop Options Header \cite{rfc2460}\\
+           1 & IPv6 Routing Header \cite{rfc2460}\\
+           2 & IPv6 Fragment Header \cite{rfc2460}\\
+           3 & IPv6 Destination Options Header \cite{rfc2460}\\
+           4 & IPv6 Mobility Header \cite{rfc3775}\\
+           5 & Reserved\\
+           6 & Reserved\\
+           7 & IPv6 Header\\
+          \hline
+        \end{tabular}
+\end{center}
+\caption{IPv6 Extension Header ID values}\label{table:ipv6.ext.header}
+\end{table}
+
 The last bit of the NHC encoding indicates whether the NHC compression is used for the following header.
-
 \begin{figure}[htp]
@@
 
-This format also allows to compress down well-known ports (0xF0Bx).	When the P field has the value $01$, all 16 bits for the Source Port are carried in-line, whereas first 8 bits of the Destination Port is 0xF0 and elided, and the latest 8 bits are carried in-line.
+This format also allows to compress down well-known ports (0xF0Bx).
+When the P field has the value $01$, all 16 bits for the Source Port are carried in-line, whereas first 8 bits of the Destination Port is 0xF0 and elided, and the latest 8 bits are carried in-line.
 The value $10$ of the P field indicates that all 16 bits for the Destination Port are carried in-line and the Source Port is encoded in the same way as the Destination port in the previous case.
@@
 
-\input{msc-nd}
-%%
+\input{msc-nd} %%
 %   Conclusion
@@
 \nocite{rfc3775}
+\nocite{rfc4191}
 \nocite{rfc4291}

### 2010-01-13 doc/msc-proposal/msc-routing.tex
proposal: routing

@@
 In this approach a 6LoWPAN is seen as a single IP link and the IPv6 link-local scope covers all nodes in the LoWPAN.
-In the route over configuration intermediate nodes become LoWPAN Routers and perform standard Layer 3 (IP) routing.
-Therefore, in this case the link-local scope includes the set of nodes within symmetric radio range of a node and routing can be accomplished over various types of interconnected links.
+In the route over configuration intermediate nodes become LoWPAN Routers and perform standard layer 3 (IP) routing.
+Therefore, in this case, the link-local scope includes a set of nodes within symmetric radio range of a node and routing can be accomplished over various types of interconnected links.
+
+The IETF Routing Over Low power and Lossy networks working group  (ROLL WG) is focused on routing issues for Low power and Lossy Networks (LLNs)  composed of many embedded devices with limited power, memory, and processing resources interconnected by a variety of links, such as IEEE 802.15.4 or Low Power WiFi.
+The aim of the group is to provide an IPv6 architectural framework for routing and path selection in LLNs.
+So far several internet drafts have been published, such as an Overview of Existing Routing Protocols for Low Power and Lossy Networks \cite{draft-protocols-07},  IPv6 Routing Protocol for Low power and Lossy Networks \cite{draft-rpl-04}, and Routing Metrics used for Path Calculation in Low Power and Lossy Networks \cite{draft-routing-metrics-04}.
 
@@
 
-This chapter discusses a list of requirements for 6LoWPAN routing \cite{draft-routing-04} in Section \ref{sec:rout.req}.
-Section \ref{sec:rout.rpl} describes the Internet Draft \cite{draft-rpl-04} which specifies the IPv6 Routing Protocol for Low power and Lossy networks (LLNs).
+Further in this chapter, the list of requirements for 6LoWPAN routing defined in the internet draft \cite{draft-routing-04} is discussed in Section \ref{sec:rout.req}.
+The evaluation of existing  routing protocol for LLN scenarious is presented in Section \ref{sec:rout.protocols}.
+And,  finally, Section \ref{sec:rout.rpl} describes the Internet Draft \cite{draft-rpl-04} which specifies the IPv6 Routing Protocol for LLNs.
 
@@
 Battery-operated devices need to last from several months to a few years with a single AA battery.
-Therefore 6LoWPAN routing protocols have to cause minimal power consumption by the efficient use of control packets, minimization of expensive IP multicast which causes link broadcast to the entire LoWPAN, and by the efficient routing of data packets.
-Control messages have to fit a single IEEE 802.15.4 frame size in order to avoid packet fragmentation and the overhead for reassembly.
-The design of 6LoWPAN routing protocols should be scalable to support from a few nodes to millions of them.
+Therefore 6LoWPAN routing protocols have to cause minimal power consumption by the efficient use of control packets, minimization of expensive IP multicast which causes link broadcast to the entire LoWPAN, and by  efficient routing of data packets.
+Control messages have to fit into a single IEEE 802.15.4 frame in order to avoid packet fragmentation and the overhead for reassembly.
+The design of 6LoWPAN routing protocols should be scalable to support networks ranging from a few nodes to millions of nodes.
 
-6LoWPAN devices are unreliable due to limited system capabilities and unpredictable environment where they can be deployed.
+6LoWPAN devices are unreliable due to limited system capabilities and an unpredictable environment where they can be deployed.
 6LoWPAN routing protocols have to be robust to dynamic loss caused by link failure or device unavailability.
@@
 
-\section{Existing Routing Protocols}
+\section{Existing Routing Protocols}\label{sec:rout.protocols}
 
 \section{Routing Protocol for Low power and Lossy networks}\label{sec:rout.rpl}
-The ROLL WG is now working on route over approaches for Low power and Lossy Networks (LLNs), not specifically for 6LoWPAN. 
+
 \subsection{Routing Metrics}

### 2010-01-14 doc/msc-proposal/msc-routing.tex
proposal: overview of existing routing protocols

@@
 \section{Existing Routing Protocols}\label{sec:rout.protocols}
+The internet draft \cite{draft-protocols-07} provides a survey of the strengths and weaknesses of existing routing protocols with respect to the 6LoWPAN routing requirements.
+The survey examines whether existing and mature IETF protocols can meet LLN requirements without modifications.
+The list of considered protocols is OSPF \cite{rfc2328}, IS-IS \cite{rfc1142}, RIP \cite{rfc2453}, OLSR \cite{rfc3626}, OLSv2 \cite{draft-manet-olsrv2}, TBRPF \cite{rfc3684}, AODV \cite{rfc3561}, DYMO \cite{draft-manet-dymo}, and DSR \cite{rfc4728}.
+
+The survey uses five criteria derived from a set of requirements for routing in low power and lossy networks:  routing state, loss response, control cost, link cost, and node cost.
+The routing state criterion indicates whether routing state scales reasonably within the memory resources of low-power nodes.
+Routing state that scales linearly with the size of the network or a node's neighbourhood fails and passes if scales with the number of destinations.
+The loss response indicates whether the protocol localizes responses to link failures with no triggering of global network re-optimization.
+Protocols which require many link changes to propagate across the entire network fail.
+The control cost criterion defines constraints on control traffic, in order to discover a topology.
+The link and node cost specify how a protocol chooses routes for data packets to take through the network.
+A protocol passes these criteria if it provides a mechanism allowing link and node properties to be considered when choosing routes.
+
+Table \ref{table:routing.prot.survey} summarizes the survey showing which of existing protocols meet the criteria described above.
+The detailed analysis is given in \cite{draft-protocols-07}.
+For each of these criteria, the value "pass" indicates that a protocol has satisfactory performance.
+The value "fail" corresponds to not acceptable performance, which means that the protocol does not meet the criterion.
+Finally, a "?" means that a protocol would require a supplementary document specifying how a protocol should behave.
+As can be seen from the table, no existing IETF protocol meets the described criteria.
+Therefore, the survey concludes that a new protocol specification documents have to be defined for a LLN routing protocol.
+
+\begin{table}[htp]
+\begin{center}
+        \begin{tabular}{|l|c|c|c|c|c|}
+          \hline
+          Protocol   &   State &  Loss & Control &  Link Cost & Node Cost\\
+          \hline
+          \hline
+     OSPF/IS-IS  &  fail  &  fail  &  fail   &   pass    &   fail\\
+     OLSRv2      &  fail  &   ?    &   ?     &   pass    &   pass\\
+     TBRPF       &  fail  &  pass  &  fail   &   pass    &    ?\\
+     RIP         &  pass  &  fail  &  pass   &    ?      &   fail\\
+     AODV        &  pass  &  fail  &  pass   &   fail    &   fail\\
+     DYMO        &  pass  &   ?    &  pass   &    ?      &    ?\\
+     DSR         &  fail  &  pass  &  pass   &   fail    &   fail\\
+          \hline
+        \end{tabular}
+\end{center}
+\caption{Routing protocol survey results}\label{table:routing.prot.survey}
+\end{table}
 

### 2010-01-15 trunk/doc/msc-proposal/msc-routing.tex
proposal: RPL terminology

@@
 In this approach a 6LoWPAN is seen as a single IP link and the IPv6 link-local scope covers all nodes in the LoWPAN.
-In the route over configuration intermediate nodes become LoWPAN Routers and perform standard Layer 3 (IP) routing.
-Therefore, in this case, the link-local scope includes the set of nodes within symmetric radio range of a node and routing can be accomplished over various types of interconnected links.
+In the route over configuration intermediate nodes become LoWPAN Routers and perform standard layer 3 (IP) routing.
+Therefore, in this case, the link-local scope includes a set of nodes within symmetric radio range of a node and routing can be accomplished over various types of interconnected links.
 
-The IETF Routing Over Low power and Lossy networks Working Group  (ROLL WG) is focused on routing issues for Low power and Lossy networks (LLNs)  composed of many embedded devices with limited power, memory, and processing resources interconnected by a variety of links, such as IEEE 802.15.4 or Low Power WiFi.
+The IETF Routing Over Low power and Lossy networks working group  (ROLL WG) is focused on routing issues for Low power and Lossy Networks (LLNs)  composed of many embedded devices with limited power, memory, and processing resources interconnected by a variety of links, such as IEEE 802.15.4 or Low Power WiFi.
 The aim of the group is to provide an IPv6 architectural framework for routing and path selection in LLNs.
-So far several Internet Drafts have been published, such as Overview of Existing Routing Protocols for Low Power and Lossy Networks \cite{draft-protocols-07},  IPv6 Routing Protocol for Low power and Lossy Networks \cite{draft-rpl-04}, and Routing Metrics used for Path Calculation in Low Power and Lossy Networks \cite{draft-routing-metrics-04}.
+So far several internet drafts have been published, such as an Overview of Existing Routing Protocols for Low Power and Lossy Networks \cite{draft-protocols-07},  IPv6 Routing Protocol for Low power and Lossy Networks \cite{draft-rpl-04}, and Routing Metrics used for Path Calculation in Low Power and Lossy Networks \cite{draft-routing-metrics-04}.
 
@@
 
-Further in this chapter, the list of requirements for 6LoWPAN routing defined in the Internet Draft \cite{draft-routing-04} is discussed in Section \ref{sec:rout.req}.
+Further in this chapter, the list of requirements for 6LoWPAN routing defined in the internet draft \cite{draft-routing-04} is discussed in Section \ref{sec:rout.req}.
 The evaluation of existing  routing protocol for LLN scenarious is presented in Section \ref{sec:rout.protocols}.
-And,  finally, Section \ref{sec:rout.rpl} describes the Internet Draft \cite{draft-rpl-04} which specifies the IPv6 Routing Protocol for LLNs.
+And,  finally, Section \ref{sec:rout.rpl} describes the internet draft \cite{draft-rpl-04} which specifies the IPv6 Routing Protocol for LLNs.
 
@@
 \section{Existing Routing Protocols}\label{sec:rout.protocols}
+The internet draft \cite{draft-protocols-07} provides a survey of the strengths and weaknesses of existing routing protocols with respect to the 6LoWPAN routing requirements.
+The survey examines whether existing and mature IETF protocols can meet LLN requirements without modifications.
+The list of considered protocols is OSPF \cite{rfc2328}, IS-IS \cite{rfc1142}, RIP \cite{rfc2453}, OLSR \cite{rfc3626}, OLSv2 \cite{draft-manet-olsrv2}, TBRPF \cite{rfc3684}, AODV \cite{rfc3561}, DYMO \cite{draft-manet-dymo}, and DSR \cite{rfc4728}.
 
-\section{Routing Protocol for Low power and Lossy networks}\label{sec:rout.rpl}
+The survey uses five criteria derived from a set of requirements for routing in low power and lossy networks:  routing state, loss response, control cost, link cost, and node cost.
+The routing state criterion indicates whether routing state scales reasonably within the memory resources of low-power nodes.
+Routing state that scales linearly with the size of the network or a node's neighbourhood fails and passes if scales with the number of destinations.
+The loss response indicates whether the protocol localizes responses to link failures with no triggering of global network re-optimization.
+Protocols which require many link changes to propagate across the entire network fail.
+The control cost criterion defines constraints on control traffic, in order to discover a topology.
+The link and node cost specify how a protocol chooses routes for data packets to take through the network.
+A protocol passes these criteria if it provides a mechanism allowing link and node properties to be considered when choosing routes.
+
+Table \ref{table:routing.prot.survey} summarizes the survey showing which of existing protocols meet the criteria described above.
+The detailed analysis is given in \cite{draft-protocols-07}.
+For each of these criteria, the value "pass" indicates that a protocol has satisfactory performance.
+The value "fail" corresponds to not acceptable performance, which means that the protocol does not meet the criterion.
+Finally, a "?" means that a protocol would require a supplementary document specifying how a protocol should behave.
+As can be seen from the table, no existing IETF protocol meets the described criteria.
+Therefore, the survey concludes that new protocol specification documents have to be defined for a LLN routing protocol.
+
+\begin{table}[htp]
+\begin{center}
+        \begin{tabular}{|l|c|c|c|c|c|}
+          \hline
+          Protocol   &   State &  Loss & Control &  Link Cost & Node Cost\\
+          \hline
+          \hline
+     OSPF/IS-IS  &  fail  &  fail  &  fail   &   pass    &   fail\\
+     OLSRv2      &  fail  &   ?    &   ?     &   pass    &   pass\\
+     TBRPF       &  fail  &  pass  &  fail   &   pass    &    ?\\
+     RIP         &  pass  &  fail  &  pass   &    ?      &   fail\\
+     AODV        &  pass  &  fail  &  pass   &   fail    &   fail\\
+     DYMO        &  pass  &   ?    &  pass   &    ?      &    ?\\
+     DSR         &  fail  &  pass  &  pass   &   fail    &   fail\\
+          \hline
+        \end{tabular}
+\end{center}
+\caption{Routing protocol survey results}\label{table:routing.prot.survey}
+\end{table}
+
+\section{IPv6 Routing Protocol for Low power and Lossy Networks}\label{sec:rout.rpl}
+\subsection{Terminology}
+The IPv6 Routing Protocol for Low power and Lossy Networks (RPL) introduces a notion of a Directed Acyclic Graph (DAG), which is a directed graph with no cycles exist.
+All edges of a DAG are oriented toward and terminating at one or more root nodes.
+A Destination Oriented DAG (DODAG) is a DAG rooted at a single destination, which is a node with no outgoing edges.
+Each root has a unique identifier, the DAGID.
+While constructing a DODAG various routing metrics and optimization objectives can be used, which are defined by Objective Function (OF) \cite{draft-routing-metrics-04}.
+The Objective Code Point (OCP) is used to indicate which OF is in use in a DODAG.
+A DAG Instance is a set of possibly multiple DODAGs.
+Each DAG instance constructs a routing topology optimized for a certain OF.
+A DAG Instance is uniquely identified by InstanceID.
+A network may have more than one DAG Instance.
+Each of them operates independently.
 

### 2010-01-15 doc/msc-proposal/msc-routing.tex
proposal: routing

@@
 The evaluation of existing  routing protocol for LLN scenarious is presented in Section \ref{sec:rout.protocols}.
-And,  finally, Section \ref{sec:rout.rpl} describes the Internet Draft \cite{draft-rpl-04} which specifies the IPv6 Routing Protocol for LLNs.
+And,  finally, Section \ref{sec:rout.rpl} describes the internet draft \cite{draft-rpl-04} which specifies the IPv6 Routing Protocol for LLNs.
 
@@
 As can be seen from the table, no existing IETF protocol meets the described criteria.
-Therefore, the survey concludes that a new protocol specification documents have to be defined for a LLN routing protocol.
+Therefore, the survey concludes that new protocol specification documents have to be defined for a LLN routing protocol.
 
@@
 
-\section{Routing Protocol for Low power and Lossy networks}\label{sec:rout.rpl}
+\section{IPv6 Routing Protocol for Low power and Lossy Networks}\label{sec:rout.rpl}
+\subsection{Terminology}
+The IPv6 Routing Protocol for Low power and Lossy Networks (RPL) introduces a notion of a Directed Acyclic Graph (DAG), which is a directed graph with no cycles exist.
+All edges of a DAG are oriented toward and terminating at one or more root nodes.
+A Destination Oriented DAG (DODAG) is a DAG rooted at a single destination, which is a node with no outgoing edges.
+Each root has a unique identifier, the DAGID.
+While constructing a DODAG various routing metrics and optimization objectives can be used, which are defined by Objective Function (OF) \cite{draft-routing-metrics-04}.
+The Objective Code Point (OCP) is used to indicate which OF is in use in a DODAG.
+A DAG Instance is a set of possibly multiple DODAGs.
+A DAG Instance is uniquely identified by InstanceID.
+
+\subsection{RPL Model}
+Each DAG instance constructs a routing topology optimized for a certain OF.
+Traffic that belongs to a specific DODAG Instance is marked in the flow label of the IPv6 header.
+A network may have more than one DAG Instance which operate independently.
+Applications may tag traffic to follow an appropriate DAG instance, i.e., optimized for low latency or low energy.
+
+RPL nodes construct and maintain DODAGs via exchange of DAG Information Object (DIO) messages.
+A DIO message identifies the DAG Instance, the DAGID, and the values used to compute the DAG Instance's objective function.
+The DIO also includes a measure derived from the position of the node within the DODAG, the rank, used to determine its position relative to each other and to avoid loops.
+The rate at which DIO messages are sent varies depending on the state of the DODAG.
+When a DODAG is detected to be inconsistent, the RPL sends DIO messages more frequently, and, otherwise, when the DODAG is in a stable state.
+
+\subsection{Messages}
 

### 2010-01-15 doc/msc-proposal/msc-routing.tex
proposal: routing

@@
 
-\subsection{RPL Model}
+\subsection{Model}
 Each DAG instance constructs a routing topology optimized for a certain OF.
@@
 \subsection{Messages}
+The draft \cite{draft-rpl-04} defines the RPL Control Message, which is a new ICMPv6 message. The format of this message is shown in Figure \ref{fig:rpl.control.message}.
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+   0                   1                   2                   3
+   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |     Type      |     Code      |          Checksum             |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |                                                               |
+   +                         Message Body                          +
+   |                                                               |
+\end{verbatim}
+\end{mylisting}
+\caption{RPL Control Message}\label{fig:rpl.control.message}
+\end{figure}
+
+The Type for the RPL Control message has the value of 155.
+The Code identifies the RPL Control Messages as follows:  0x01 -- DAG Information Solicitation; 0x02: DAG Information Object; 0x04: Destination Advertisement Object.
+The DAG Information Solicitation (DIS) message is used to solicit a DAG Information Object from a RPL node.
+The DAG Information Object carries a number of metrics and other information that allows a node to discover a DAG Instance, select its DAG parents, and avoid loops.
+The Destination Advertisement Object (DAO) is used to propagate destination information upwards along the DODAG.
+The format of these messages is specified in the internet draft \cite{draft-rpl-04}.
 

### 2010-01-16 trunk/doc/msc-proposal/msc-routing.tex
proposal: routing

@@
 \section{IPv6 Routing Protocol for Low power and Lossy Networks}\label{sec:rout.rpl}
-\subsection{Terminology}
 The IPv6 Routing Protocol for Low power and Lossy Networks (RPL) introduces a notion of a Directed Acyclic Graph (DAG), which is a directed graph with no cycles exist.
-All edges of a DAG are oriented toward and terminating at one or more root nodes.
-A Destination Oriented DAG (DODAG) is a DAG rooted at a single destination, which is a node with no outgoing edges.
-Each root has a unique identifier, the DAGID.
-While constructing a DODAG various routing metrics and optimization objectives can be used, which are defined by Objective Function (OF) \cite{draft-routing-metrics-04}.
-The Objective Code Point (OCP) is used to indicate which OF is in use in a DODAG.
-A DAG Instance is a set of possibly multiple DODAGs.
-Each DAG instance constructs a routing topology optimized for a certain OF.
-A DAG Instance is uniquely identified by InstanceID.
-A network may have more than one DAG Instance.
-Each of them operates independently.
+All edges of a DAG are oriented toward and terminating at one or more root nodes uniquely identified by the DAGID.
+A Destination Oriented DAG (DODAG) is a DAG rooted at a single destination.
+While constructing a DODAG various routing metrics and optimization objectives can be used to compute the shortest paths.
+The set of supported link and node metrics used in the RPL is specified in the internet draft \cite{draft-routing-metrics-04}.
 
-\subsection{Routing Metrics}
-Historically, IGP such as OSPF ([RFC2328]) and IS-IS ([RFC1195]) have used quantitative static link metric
+A DAG instance is a set of possibly multiple DODAGs.
+The routing metrics and objectives for a certain DAG instance are specified by an Objective Function (OF).
+Each DAG instance constructs a routing topology optimized for its OF.
+Traffic that belongs to a specific DODAG instance is marked in the flow label of the IPv6 header.
+A network may have more than one DAG instance which operate independently.
+This allows to applications to tag traffic to follow an appropriate DAG instance, i.e., optimized for low latency or low energy.
+
+\subsection{Messages}
+The protocol defines the RPL Control Message, which is a new ICMPv6 message. The format of this message is shown in Figure \ref{fig:rpl.control.message}.
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+   0                   1                   2                   3
+   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |     Type      |     Code      |          Checksum             |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |                                                               |
+   +                         Message Body                          +
+   |                                                               |
+\end{verbatim}
+\end{mylisting}
+\caption{RPL Control Message}\label{fig:rpl.control.message}
+\end{figure}
+
+The Type for the RPL Control message has the value of 155.
+The Code identifies the RPL Control Messages as follows:  0x01 -- DAG Information Solicitation; 0x02: DAG Information Object; 0x04: Destination Advertisement Object.
+
+DAG Information Object (DIO) message exchanges are used by RPL nodes to construct and maintain DODAGs.
+A DIO message identifies the DAG instance, the DAGID, and the values used to compute the DAG instance's objective function.
+The DIO also includes a measure derived from the location of the node within the DODAG, the rank, used to determine its position relative to other nodes and avoid loops.
+The exact calculation of the rank is left to the Objective Function and  depends on parents of the node.
+The DAG Information Solicitation (DIS) message is used to solicit a DAG Information Object from a RPL node.
+
+DIO messages allow to establish upward routes, i.e., the routes in the direction from leaf nodes towards DODAG roots.
+In order to establish downward routes, Destination Advertisement Object (DAO) messages are used.
+The DAO propagates destination information upwards along the DODAG and includes prefix information to identify destinations.
+Nodes that are capable to maintain routing state may aggregate routes from DAO messages that they receive before transmitting them.
+Nodes that are not capable to maintain routing state attaches a next-hop address to the Reverse Route Stack contained within the DAO message.
+The Reverse Route Stack contains piecewise routes over regions of the LLN that are incapable of storing routing state.
+
+The detailed format of the messages is specified in the internet draft \cite{draft-rpl-04}.
+
+\subsection{Basic Operation}
+A node's neighbor set is the nodes that it can reach with a link-local multicast.
+A node that is not a DODAG root may maintain multiple DAG parents for a single DAG instance.
+DAG parents of a node is a subset of the set of       its neighbors.
+A node may implement different policies for selecting its DAG parents.
+
+DAG discovery allows a node to join a DODAG by discovering neighbors that are members of the DODAG.
+Some nodes are preconfigured to be DODAG roots.
+Each node maintains a timer that governs when to advertise node presence by sending link-local multicast DIO messages.
+The rate at which DIO messages are sent varies in response to stability or detection of routing inconsistencies.
+Nodes listen for DIOs and use their information to join a new DODAG, or to maintain an existing DODAG, according to the specified Objective Function.
+DODAG discovery may avoid loops by constraining how and when nodes can increase their rank.

### 2010-01-17 doc/msc-proposal/msc-routing.tex
proposal: corrected ieee 802.15.4

--- 2010-01-14 doc/msc-proposal/msc-routing.tex
@@
 The routing state criterion indicates whether routing state scales reasonably within the memory resources of low-power nodes.
-Routing state that scales linearly with the size of the network or a node's neighbourhood fails and passes if scales with the number of destinations.
+Routing state that scales linearly with the size of the network or a node's neighbourhood fails, and passes if scales with the number of destinations.
 The loss response indicates whether the protocol localizes responses to link failures with no triggering of global network re-optimization.
 Protocols which require many link changes to propagate across the entire network fail.
-The control cost criterion defines constraints on control traffic, in order to discover a topology.
+The control cost criterion defines constraints on control traffic required to discover a topology.
 The link and node cost specify how a protocol chooses routes for data packets to take through the network.
-A protocol passes these criteria if it provides a mechanism allowing link and node properties to be considered when choosing routes.
+A protocol passes these two criteria if it provides a mechanism allowing both link and node properties to be considered when choosing routes.
 
 Table \ref{table:routing.prot.survey} summarizes the survey showing which of existing protocols meet the criteria described above.
-The detailed analysis is given in \cite{draft-protocols-07}.
+The detailed analysis is given in the document \cite{draft-protocols-07}.
 For each of these criteria, the value "pass" indicates that a protocol has satisfactory performance.
@@
 Finally, a "?" means that a protocol would require a supplementary document specifying how a protocol should behave.
-As can be seen from the table, no existing IETF protocol meets the described criteria.
-Therefore, the survey concludes that a new protocol specification documents have to be defined for a LLN routing protocol.
+As can be seen from the table, no existing IETF protocol meets all described criteria.
+Therefore, the survey concludes that new protocol specification documents have to be defined for a LLN routing protocol.
 
@@
 
-\section{Routing Protocol for Low power and Lossy networks}\label{sec:rout.rpl}
+\section{Overview of IPv6 Routing Protocol for LLNs}\label{sec:rout.rpl}
+The IPv6 Routing Protocol for Low power and Lossy Networks (RPL) introduces a notion of a Directed Acyclic Graph (DAG), which is a directed graph with no cycles exist.
+All edges of a DAG are oriented toward and terminating at one or more root nodes uniquely identified by the DAGID.
+A Destination Oriented DAG (DODAG) is a DAG rooted at a single destination.
+While constructing a DODAG various routing metrics and optimization objectives can be used to compute the shortest paths.
+The set of supported link and node metrics used in the RPL is specified in the internet draft \cite{draft-routing-metrics-04}.
 
-\subsection{Routing Metrics}
-Historically, IGP such as OSPF ([RFC2328]) and IS-IS ([RFC1195]) have used quantitative static link metric
+A DAG instance is a set of possibly multiple DODAGs.
+The routing metrics and objectives for a certain DAG instance are specified by an Objective Function (OF).
+Each DAG instance constructs a routing topology optimized for its OF.
+Traffic that belongs to a specific DODAG instance is marked in the flow label of the IPv6 header.
+A network may have more than one DAG instance which operate independently.
+This allows to applications to tag traffic to follow an appropriate DAG instance, i.e., optimized for low latency or low energy.
+
+\subsection{Messages}
+The protocol defines the RPL Control Message, which is a new ICMPv6 message. The format of this message is shown in Figure \ref{fig:rpl.control.message}.
+\begin{figure}[htp]
+\begin{mylisting}
+\begin{verbatim}
+   0                   1                   2                   3
+   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |     Type      |     Code      |          Checksum             |
+   +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
+   |                                                               |
+   +                         Message Body                          +
+   |                                                               |
+\end{verbatim}
+\end{mylisting}
+\caption{RPL Control Message}\label{fig:rpl.control.message}
+\end{figure}
+
+The Type for the RPL Control message has the value of 155.
+The Code identifies the RPL Control Messages as follows:  0x01 -- DAG Information Solicitation; 0x02: DAG Information Object; 0x04: Destination Advertisement Object.
+
+DAG Information Object (DIO) message exchanges are used by RPL nodes to construct and maintain DODAGs.
+A DIO message identifies the DAG instance, the DAGID, and the values used to compute the DAG instance's objective function.
+The DIO also includes a measure derived from the location of the node within the DODAG, the rank, used to determine the position of the node relative to other nodes and avoid loops.
+The exact calculation of the rank is left to the Objective Function and  depends on the parents of the node.
+The DAG Information Solicitation (DIS) message is used to solicit a DAG Information Object from a RPL node.
+
+DIO messages allow to establish upward routes, i.e., the routes in the direction from leaf nodes towards DODAG roots.
+In order to establish downward routes Destination Advertisement Object (DAO) messages are used.
+The DAO propagates destination information upwards along the DODAG.
+
+The detailed format of the messages is specified in the internet draft \cite{draft-rpl-04}.
+
+\subsection{Basic Operation}
+A node's neighbor set is the nodes that it can reach with a link-local multicast.
+A node that is not a DODAG root may maintain multiple DAG parents for a single DAG instance.
+DAG parents of a node is a subset of the set of       its neighbors.
+A node may implement different policies for selecting its DAG parents.
+
+DAG discovery allows a node to join a DODAG by discovering neighbors that are members of the DODAG.
+Some nodes are preconfigured to be DODAG roots.
+Each node maintains a timer that governs when to advertise node presence by sending link-local multicast DIO messages.
+The rate at which DIO messages are sent varies in response to stability or detection of routing inconsistencies.
+Nodes listen for DIOs and use their information to join a new DODAG, or to maintain an existing DODAG, according to the specified Objective Function.
+DODAG discovery avoids loops by constraining how and when nodes can increase their rank.
+
+The destination advertisement mechanism supports the dissemination of routing state required to support traffic flows down along the DODAG, from the DODAG root toward nodes.
+Destinations disseminated with the destination advertisement mechanism may be prefixes, individual hosts, or multicast listeners.
+The mechanism supports nodes with varying capabilities.
+Nodes that are capable to maintain routing state may inspect destination advertisements and learn hop-by-hop routing state toward destinations populating their routing tables with the learned routes.
+They may also learn necessary piecewise source routes to traverse regions of the LLN that do not maintain routing state.
+Route aggregation on known destinations can be accomplished by such nodes before emitting Destination Advertisements.
+The nodes that are incapable of storing routing state forward destination advertisements, attaching a next-hop address to the reverse route used to support the construction of piecewise source routes.
+
+A comprehensive description of the RPL is given in the internet draft \cite{draft-rpl-04}

### 2010-01-17 trunk/doc/msc-proposal/msc-proposal.tex
proposal: corrected ieee 802.15.4

@@
 \chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
-The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks.
-The standard intends to offer the fundamental lower network layers of a type of wireless personal area network which focuses on low-cost, low-speed communication between tiny devices (motes).
-Such networks are typically limited to a personal operating space (POS) of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.
+The IEEE 802.15.4 standard \cite{ieee802.15.4} developed by the 802.15.4 Task Group within the IEEE specifies the physical layer and media access control for low-rate wireless personal area networks providing node-to-node frame delivery between devices within reachable distance from each other.
+The standard is targeted at low cost, low speed and low power consumption devices.
+This chapter provides an overview of the IEEE 802.15.4 standard.
 
+\section{Device types}
 The IEEE 802.15.4 standard distinguishes between two types of nodes, reduced-function devices (RFDs) and full-function devices (FFDs).
-FFDs typically have more resources and may be mains powered.
-Therefor, FFDs aid RFDs by providing functions such as network coordination and packet forwarding.
+FFDs typically have more resources, implement the complete protocol set and may be mains powered.
+An FFD can talk to RFDs or other FFDs, while an RFD can talk only to an FFD.
+FFDs aid RFDs by providing functions such as network coordination and packet forwarding.
+An RFD is intended for applications that are extremely simple, which do not have the need to send large amounts of data and may only associate with a single FFD at a time.
+Therefore, RFDs implement minimal subset of the IEEE 802.15.4 protocol and can be implemented using minimal resources and memory capacity.
 
+Two or more devices within a personal operating space communicating on the same physical channel constitute a wireless personal area network (WPAN). However, a network has to include at least one FFD.
+
+\section{Network topologies}
 An 802.15.4 network may operate in either the star or the peer-to-peer topology.
-In the star topology devices communicate with a single central                           a personal area network (PAN) coordinator.
-In the peer-to-peer topology a device can communicate with any other device as long as they are in range of one another.
+In the star topology devices communicate with a single central personal area network (PAN) coordinator.
+Only an FDD device can be a PAN coordinator.
+The PAN coordinator is usually mains powered, while the other devices are most likely battery operated.
+In the peer-to-peer topology a device can communicate with any other device as long as they are in range of each other.
 Based on the peer-to-peer topology more complex network formations may be constructed, such as mesh networking topology.
-However, the standard does not define a network layer, therefor routing is not directly supported, but such an additional layer can add support for multihop communications.
+However, the standard does not define a network layer, therefore, routing is not directly supported, but such an additional layer can add support for multihop communications.
 All devices operating on a network of either topology have unique 64 bit extended addresses.
-This address can be used for direct communication within the PAN, or it can be exchanged for a short 16 bit address allocated by the PAN coordinator.
+This address can be used for direct communication within the PAN.
+Each independent PAN selects a unique identifier.
+This PAN identifier allows communication between
+devices within a network using short 16 bit addresses and enables transmissions between devices across independent networks.
 
-The basic unit of data transport is a frame. The standard defines four frame structures: a beacon frame, used by a coordinator to transmit beacons;
-a data frame, used for all transfers of data; an acknowledgment frame, used for confirming successful frame reception; a MAC command frame, used for handling all MAC control transfers.
+\section{Frame structure}
+The basic unit of data transport is a frame.
+The standard defines four frame structures: a beacon frame, used by a coordinator to transmit beacons; a data frame, used for all transfers of data;  frame, used for confirming successful frame reception; a MAC command frame, used for handling all MAC control transfers.
 Additionally, a superframe structure may be defined by the coordinator.
@@
 An important aspect of 802.15.4 is its limitation on the frame size, which is specified by the frame length 7 bit field (0-127 bytes).
-Taking into account the frame header, which is up to 25 octets, this leaves 102 bytes for the payload of the higher layers.
+Taking into account the frame header, which is up to 25 octets (disabled security), this leaves 102 bytes for the payload of the higher layers.
 
-The standard also defines several security services such as maintaining an access control list (ACL) and
-using symmetric-key cryptography to protect transmitted frames. However, key management is not specified and must be provided by the higher layers.
+\section{Security}
+The standard defines several security services such as maintaining an access control list (ACL) and
+using symmetric-key cryptography to protect transmitted frames.
+Access control is a security service that provides the ability for a device to select the other devices with which it is willing to communicate.
+If the access control service is enabled, a device maintains a list of devices in its ACL from which it expects to receive frames.
+In order to protect data from being read by parties without permission, data may be encrypted using a key shared by a group of devices or using a key shared between two peers.
+However, key management is not specified in the standard and must be provided by the higher layers.
+Security services also specify frame integrity that uses a message integrity code to protect data from being modified by parties without the cryptographic key.
+This provides assurance that data came from a party with the cryptographic key.
 
@@
 As described in Chapter \ref{ch:ieee802.15.4}, the IEEE 802.15.4-2003 standard specifies the physical layer and media access control for low-rate wireless personal area network providing node-to-node frame delivery between the devices within reachable distance from each other.
-On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be defined that would provide end-to-end packet delivery including routing through intermediate hosts.
+On top of IEEE 802.15.4, a network layer (layer 3 of the seven-layer OSI model) has to be defined that provides end-to-end packet delivery including routing through intermediate nodes.
 

### 2010-01-17 doc/msc-proposal/msc-proposal.tex
proposal: prof. corrections
proposal: 802.15.4

@@
 \chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
-The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks providing nodo-to-node frame dilivery between devices within reachable distance from each other.
-Such networks are typically limited to a personal operating space of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.
+The IEEE 802.15.4 standard developed by the 802.15.4 Task Group within the IEEE specifies the physical layer and media access control for low-rate wireless personal area networks providing node-to-node frame delivery between devices within reachable distance from each other.
+The standard is targeted at low cost, low speed and low power consumption devices.
+Several versions of the standard have been published, such as 802.15.4-2003 \cite{ieee802.15.4}, 802.15.4-2006 \cite{ieee802.15.4-2006}, 802.15.4a-2007 \cite{ieee802.15.4a-2007}, 802.15.4c-2009 \cite{ieee802.15.4c-2009} and 802.15.4d-2009 \cite{ieee802.15.4d-2009} .
+This chapter provides an overview of the IEEE 802.15.4 standard.
 
+\section{Device types}
 The IEEE 802.15.4 standard distinguishes between two types of nodes, reduced-function devices (RFDs) and full-function devices (FFDs).
-FFDs typically have more resources and may be mains powered.
-Therefor, FFDs aid RFDs by providing functions such as network coordination and packet forwarding.
+FFDs typically have more resources, implement the complete protocol set and may be mains powered.
+An FFD can talk to RFDs or other FFDs, while an RFD can talk only to an FFD.
+FFDs aid RFDs by providing functions such as network coordination and packet forwarding.
+An RFD is intended for applications that are extremely simple, which do not have the need to send large amounts of data and may only associate with a single FFD at a time.
+Therefore, RFDs implement minimal subset of the IEEE 802.15.4 protocol and can be implemented using minimal resources and memory capacity.
 
+Two or more devices within a personal operating space communicating on the same physical channel constitute a wireless personal area network (WPAN). However, a network has to include at least one FFD.
+
+\section{Network topologies}
 An 802.15.4 network may operate in either the star or the peer-to-peer topology.
 In the star topology devices communicate with a single central personal area network (PAN) coordinator.
-In the peer-to-peer topology a device can communicate with any other device as long as they are in range of one another.
+Only an FDD device can be a PAN coordinator.
+The PAN coordinator is usually mains powered, while the other devices are most likely battery operated.
+In the peer-to-peer topology a device can communicate with any other device as long as they are in range of each other.
 Based on the peer-to-peer topology more complex network formations may be constructed, such as mesh networking topology.
-However, the standard does not define a network layer, therefor routing is not directly supported, but such an additional layer can add support for multihop communications.
+However, the standard does not define a network layer, therefore, routing is not directly supported, but such an additional layer can add support for multihop communications.
 All devices operating on a network of either topology have unique 64 bit extended addresses.
-This address can be used for direct communication within the PAN, or it can be exchanged for a short 16 bit address allocated by the PAN coordinator.
+This address can be used for direct communication within the PAN.
+Each independent PAN selects a unique identifier.
+This PAN identifier allows communication between
+devices within a network using short 16 bit addresses and enables transmissions between devices across independent networks.
 
-The basic unit of data transport is a frame. The standard defines four frame structures: a beacon frame, used by a coordinator to transmit beacons;
-a data frame, used for all transfers of data; an acknowledgment frame, used for confirming successful frame reception; a MAC command frame, used for handling all MAC control transfers.
+\section{Frame structure}
+The basic unit of data transport is a frame.
+The standard defines four frame structures: a beacon frame, used by a coordinator to transmit beacons; a data frame, used for all transfers of data;  frame, used for confirming successful frame reception; a MAC command frame, used for handling all MAC control transfers.
 Additionally, a superframe structure may be defined by the coordinator.

### 2010-01-20 trunk/doc/msc-proposal/msc-proposal.tex
proposal: added title

@@
 
-\title{IPv6 over Low-Power Wireless Personal Area Networks\\ MSc. Thesis Proposal}
+\title{Implementation and Evaluation of the Simple Network Management Protocol over IEEE 802.15.4 Radios under the Contiki Operating System\\ MSc. Thesis Proposal}
 \author{Siarhei Kuryla}
@@
 
-\huge IPv6 over Low-Power Wireless Personal Area Networks\\ [0.5cm]
+\huge Implementation and Evaluation of the Simple Network Management Protocol over IEEE 802.15.4 Radios under the Contiki Operating System\\ [0.5cm]
 
@@
 \chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
-The IEEE 802.15.4 standard \cite{ieee802.15.4} developed by the 802.15.4 Task Group within the IEEE specifies the physical layer and media access control for low-rate wireless personal area networks providing node-to-node frame delivery between devices within reachable distance from each other.
+The IEEE 802.15.4 standard developed by the 802.15.4 Task Group within the IEEE specifies the physical layer and media access control for low-rate wireless personal area networks providing node-to-node frame delivery between devices within reachable distance from each other.
 The standard is targeted at low cost, low speed and low power consumption devices.
+Several versions of the standard have been published, such as 802.15.4-2003 \cite{ieee802.15.4}, 802.15.4-2006 \cite{ieee802.15.4-2006}, 802.15.4a-2007 \cite{ieee802.15.4a-2007}, 802.15.4c-2009 \cite{ieee802.15.4c-2009} and 802.15.4d-2009 \cite{ieee802.15.4d-2009} .
 This chapter provides an overview of the IEEE 802.15.4 standard.

### 2010-01-20 doc/msc-proposal/msc-proposal.tex
proposal: added abstract

@@
 \begin{abstract}
-Abstract
+IEEE 802.15.4 is a standard which specifies the physical layer and media access control for low-rate wireless personal area networks providing node-to-node frame delivery between devices within reachable distance.
+The standard is targeted at low-cost, low-speed and low-power devices.
+Simple Network Management Protocol (SNMP) is a widely deployed application protocol for network management and data retrieval.
+It can be very lightweight and may fit resource-constrained devices very well.
+The focus of the proposed work will be implementation and evaluation of the SNMP protocol for devices supporting IEEE 802.15.4 Radios and the Contiki Embedded Operating System.
+This document provides background information on several related areas, such as 802.15.4 standard,                          the 6LoWPAN approach for transmitting IPv6 traffic over 802.15.4 and the SNMP protocol.
+
 \end{abstract}

### 2010-01-20 doc/msc-proposal/msc-proposal.tex
proposal: added introduction

@@
 \chapter{Introduction}
-* There are a lot of applications where we need many small devices + examples; why it's important;
-* LoWPAN\\
-* The purpose of the proposed work will be to investigate something. The proposal describes problems associated with enabling IP communication with devices in a LoWPAN.\\
-* As preparation for this work, several related areas have been studied in more
-depth and the collected background information; Structure of the proposal.\\
+A low-power wireless personal area network (LoWPAN) is a simple low cost communication network that allows wireless connectivity in applications with limited power and relaxed throughput requirements.
+LoWPANs conform to the IEEE 802.15.4 standard \cite{ieee802.15.4} and typically include devices limited in their computational power, memory, and energy availability.
+Such networks can benefit from IP and, in particular, IPv6 networking.
+IP-based devices can be easily connected to other IP-based networks by reusing existing infrastructure.
+However, due to limited nature of IEEE 802.15.4 devices the IPv6 protocol \cite{rfc2460} needs to be adapted for its use in LoWPANs.
+Devices within LoWPANs are expected to be deployed in exceedingly large numbers.
+Furthermore, they are expected to have limited display and input capabilities.
+This makes network management functionality critical for LoWPANs.
+However, deployment of the IP infrastructure in 802.15.4 networks allows to reuse existing protocols.
+The Simple Network Management Protocol (SNMP) is a widely deployed application protocol for network management and data retrieval.
+SNMP is datagram-oriented and the implementations of SNMP can be very lightweight.
+It may fit LoWPAN applications very well.
+
+The purpose of the proposed work will be to investigate whether the SNMP functionality may be reused "as is" in LoWPANs.
+The implementation and evaluation of the SNMP protocol for devices supporting IEEE 802.15.4 and the Contiki Embedded Operating System \cite{contiki} will be accomplished.
+
+As preparation for this work, several related areas have been studied in more depth to collect background information.
+The rest of the document is organised in six chapters.
+Chapter \ref{ch:ieee802.15.4} provides an introduction to the IEEE 802.15.4 standard.
+Chapter \ref{ch:6lowpan} describes the adaptation layer that allows the transmission of IPv6 packets over 802.15.4 links.
+In Chapter \ref{ch:routing} 6LoWPAN routing problems are discussed.
+The 6LoWPAN Neighbor Discovery Protocol is covered in Chapter \ref{ch:nd}.
+Chapter \ref{ch:snmp} gives an overview of the Simple Network Management Protocol.
+Chapter \ref{ch:proposal} describes details of the proposed work to be done in the scope of the Master Thesis.
 
@@
 
-\input{msc-nd} %%
+\input{msc-nd}
+
+\input{msc-snmp} %%
+%%
 %   Conclusion
 %%
-\chapter*{Proposed work}
+\chapter{Proposed work}\label{ch:proposal}
 \addcontentsline{toc}{chapter}{Proposed work}
@@
 \nocite{draft-routing-metrics-04}
-
+\nocite{rfc1157}
+\nocite{rfc1901}
+\nocite{rfc1908}
+\nocite{rfc3411}
+\nocite{rfc3418}
 \end{document}

### 2010-01-20 trunk/doc/msc-proposal/msc-proposal.tex
proposal: proposed work

@@
 
-\input{msc-nd} %%
+\input{msc-nd}
+
+\input{msc-snmp} %%
+%%
 %   Conclusion
 %%
-\chapter*{Proposed work}
-\addcontentsline{toc}{chapter}{Proposed work}
+\chapter{Proposed work}\label{ch:proposal}
+
+The proposed Master Thesis is aimed at investigating whether the SNMP protocol may be reused "as is" in 802.15.4 networks.
+For this purpose, its implementation and evaluation for devices supporting IEEE 802.15.4 and running the Contiki Embedded Operating System \cite{contiki} will be accomplished.
+The target hardware platform will be the AVR Raven board which has a 2.4 GHz transceiver and uses an 8-bit ATmega1284P micro-controller with 16KB of RAM and 128KB of flash memory.
+The implementation will be done on top of Contiki 6LoWPAN implementation.
+
+The work will be accomplished over several phases.
+In the first phase, the basic SNMPv1 protocol will be implemented.
+In the second phase, the required security primitives will be addressed and implemented, in case there will not be available ready-to-use solutions.
+Finally, a security layer will be added to the SNMP implementation.
 

### 2010-01-22 doc/msc-proposal/msc-proposal.tex
proposal: snmp chaper added, intoduction

@@
 \chapter{Proposed work}\label{ch:proposal}
-\addcontentsline{toc}{chapter}{Proposed work}
+
+The proposed Master Thesis is aimed at investigating whether the SNMP protocol may be reused "as is" in 802.15.4 networks.
+For this purpose, its implementation and evaluation for devices supporting IEEE 802.15.4 and running the Contiki Embedded Operating System \cite{contiki} will be accomplished.
+The target hardware platform will be the AVR Raven board which has a 2.4 GHz transceiver and uses an 8-bit ATmega1284P micro-controller with 16KB of RAM and 128KB of flash memory.
+The implementation will be done on top of the Contiki 6LoWPAN implementation.
+
+The work will be accomplished over several phases.
+In the first phase, the basic SNMPv1 protocol will be implemented.
+Then, the focus will be shifted to creation of a secure network layer for Contiki.
+Finally, a security layer will be added to the SNMPv1 implementation.
+
+%Network Management: One of the points of transmitting IPv6
+%       packets is to reuse existing protocols as much as possible.
+%       Network management functionality is critical for LoWPANs.
+%       However, management solutions need to meet the resource
+%       constraints as well as the minimal configuration and self-healing
+%       functionality described in Section 4.4. The Simple Network
+%       Management Protocol (SNMP) [RFC3410] is widely used for
+%       monitoring data sources and sensors in conventional networks.
+%       SNMP functionality may be translated "as is" to LoWPANs with the
+%       benefit to utilize existing tools.  However, due to the memory,
+%       processing, and message size constraints, further investigation
+%       is required to determine if the use of SNMPv3 is suitable, or if
+%       an appropriate adaptation of SNMPv3 or use of different protocols
+%       is in order.
 

### 2010-01-22 trunk/doc/msc-proposal/msc-snmp.tex
proposal: snmp chaper added, intoduction

\chapter{Simple Network Management Protocol}\label{ch:snmp}
Large networks has too many components to be managed by human agents alone. Network devices have to maintain a large amount of management data such as configuration information,  operational state and statistics. Management information can be used to understand how a network performs, how devices in a network are configured and to change their configuration. The Simple Network Management Protocol (SNMP) \cite{rfc3410} is an application layer protocol that facilitates the exchange of management information between network devices. It exposes management data in the form of variables on the managed systems, which describe the system configuration. Since its first publication in 1988, the SNMP protocol has become the most widely-used network management tool for IP-based networks.

Currently, there are three versions of SNMP defined. The first version (SNMPv1) is nowadays a historical IETF standard, although it is still widely supported by many vendors. SNMPv1 has been criticized for its poor security based on a community string, which is a type of password transmitted in cleartext. SNMPv2 extended the functionality of SNMPv1 and includes a number of improvements, such as additional protocol operations. A new security system was proposed, however, it was too complex and was not widely accepted. SNMPv3 addresses the security problems of the previous versions. The SNMPv3 architecture introduces a well defined extensible architecture and the User-based Security Model (USM) for message security.

The SNMP protocol is datagram-oriented and its implementation can be very lightweight \cite{draft-6lowpan-snmp}. It can fit resource-constrained devices very well. This makes it a perfect candidate as a management protocol for 6LoWPAN applications. In this chapter an overview of the SNMP protocol is given.

\section{SNMP Concepts}
* SNMPv1 and SNMPv2 have two kinds of entities: managers and agents.

* SNMPv3 makes a big change to the basic concepts and abandons the notion of managers and agents, both are now called SNMP entities.

\section{The Structure of Management Information (SMI)}

\section{Management Information Base}

\section{Protocol Architecture}

\subsection{SNMPv3 Engine}

\subsection{The SNMPv3 Applications}

\section{SNMP Operations}

\section{Security Models}

### 2010-01-23 trunk/doc/msc-proposal/msc-snmp.tex
proposal: added SNMP architecture image

@@
 
-\section{SNMP Concepts}
-* SNMPv1 and SNMPv2 have two kinds of entities: managers and agents.
+\section{Protocol Architecture}
+The SNMP protocol consists of four basic components: an entity which provides remote access to management instrumentation (typically called as an agent); an entity running management applications (typically called a manager); a management protocol to convey management information between the entities; and management information.
 
-* SNMPv3 makes a big change to the basic concepts and abandons the notion of managers and agents, both are now called SNMP entities.
+According to the SNMP architecture \cite{rfc3411}, an SNMP entity consists of an SNMP engine and one or more associated applications.
 
-\section{The Structure of Management Information (SMI)}
-
-\section{Management Information Base}
-
-\section{Protocol Architecture}
-
-\subsection{SNMPv3 Engine}
-
-\subsection{The SNMPv3 Applications}
+\begin{figure}	
+\begin{center}
+    \includegraphics[scale = 0.6]{img/snmp-arch.pdf}
+    \caption{Structure of an SNMP entity}   
+	\label{fig:snmparch}
+\end{center}
+\end{figure}
 

### 2010-01-23 doc/msc-proposal/msc-snmp.tex
proposal: SNMP architecture

--- 2010-01-23 trunk/doc/msc-proposal/msc-snmp.tex
@@
 \section{Protocol Architecture}
-The SNMP protocol consists of four basic components: an entity which provides remote access to management instrumentation (typically called as an agent); an entity running management applications (typically called a manager); a management protocol to convey management information between the entities; and management information.
+The SNMP protocol is designed to have a modular architecture, which allows the evolution of the protocol and enables protocol extensions.
+The architecture includes several subsystems communicating via defined abstract service interfaces.
+Abstract service interfaces describe the conceptual interfaces between various subsystems.
+While the subsystems can be changed and extended over time, the interfaces are fixed.
 
-According to the SNMP architecture \cite{rfc3411}, an SNMP entity consists of an SNMP engine and one or more associated applications.
-
-\begin{figure}	
+\begin{figure}[htp]	
 \begin{center}
@@
 
-\section{SNMP Operations}
+According to the SNMP architecture defined in \cite{rfc3411}, an SNMP entity consists of an SNMP engine and one or more associated applications.
+The SNMP engine contains  a dispatcher, a message processing subsystem, a security subsystem, an access control subsystem, and a transport subsystem (Figure \ref{fig:snmparch}).
+Each subsystem may include multiple concrete models implementing different services.
+The dispatcher is responsible for controlling the data flow within an SNMP entity.
+Extracting data from received messages and preparing messages for sending are accomplished in the message processing subsystem.
+Each message processing model defines the format of a particular version of an SNMP message.
+Typically, the message processing subsystem supports three models for SNMPv1, SNMPv2c, and SNMPv3.
+The security subsystem provides security services such as authentication and privacy of messages.
+Authorization services are provided by the access control subsystem.
+The transport subsystem \cite{rfc5590} allows for multiple transport protocols to be used.
+
+There are several types of defined applications (Figure \ref{fig:snmparch}), such as a command generator which monitor and manipulate management data, a command responder providing access to management data, a notification originator initiating asynchronous messages, a notification receiver processing these messages, and a proxy forwarder which forwards messages between entities.
+Applications may use the services provided by the SNMP engine.
+An SNMP entity which includes one or more command generator and notification receiver applications is called an SNMP manager.
+An entity containing one or more command generator and notification receiver applications is called an SNMP manager.
+
+\section{Operations}
 

### 2010-01-23 trunk/doc/msc-proposal/msc-snmp.tex
proposal: SNMP architecture

@@
 
-\section{SNMP Operations}
+According to the SNMP architecture defined in \cite{rfc3411}, an SNMP entity consists of an SNMP engine and one or more associated applications.
+The SNMP engine is composed of a dispatcher, a message processing subsystem, a security subsystem, an access control subsystem, and a transport subsystem (Figure \ref{fig:snmparch}).
+Each subsystem may include multiple concrete models implementing different services.
+The dispatcher is the key part of an SNMP engine.
+It's responsible for controlling the data flow within an SNMP entity.
+When an SNMP message needs to be prepared or when data needs to be extracted from an SNMP message, the dispatcher delegates these tasks to a version-specific message processing model within the message processing subsystem.
+It also dispatches SNMP  protocol data units (PDUs) to SNMP applications.
+Each message processing model defines the format of a particular version of an SNMP message.
+Typically, the message processing subsystem supports three models for SNMPv1, SNMPv2c, and SNMPv3.
+The security subsystem provides security services such as authentication and privacy of messages.
+Authorization services are provided by the access control subsystem.
+The transport subsystem \cite{rfc5590} allows for multiple transport protocols to be used.
+
+There are several types of defined applications (Figure \ref{fig:snmparch}), such as a command generator which monitor and manipulate management data, a command responder providing access to management data, a notification originator initiating asynchronous messages, a notification receiver processing these messages, and a proxy forwarder which forwards messages between entities.
+Applications may use the services provided by the SNMP engine.
+An SNMP entity which includes one or more command generator and notification receiver applications is called an SNMP manager.
+An entity containing one or more command generator and notification receiver applications is called an SNMP manager.
+
+\section{Operations}
 
 \section{Security Models}
+
+\section{SMI and MIB}

### 2010-01-28 trunk/doc/msc-proposal/msc-snmp.tex
proposal: snmp message format

@@
 The first version (SNMPv1) is nowadays a historical IETF standard, although it is still widely supported by many vendors.
-SNMPv1 has been criticized for its poor security based on a community string, which is a type of password transmitted in cleartext.
+One of the most important weaknesses of SNMPv1 is the lack of adequate mechanisms for securing the management function.
+Its security is based on a community string, which is a type of password transmitted in cleartext.
 SNMPv2 extended the functionality of SNMPv1 and includes a number of improvements, such as additional protocol operations.
@@
 This makes it a perfect candidate as a management protocol for 6LoWPAN applications.
-In this chapter an overview of the SNMP protocol is given.
+In this chapter an overview of the SNMPv3 protocol is given.
 
@@
 The SNMP protocol is designed to have a modular architecture, which allows the evolution of the protocol and enables protocol extensions.
-The architecture includes several subsystems communicating via defined abstract service interfaces.
+The protocol architecture is composed of an SNMP engine and applications.
+An SNMP entity is an implementation of this architecture.
+The SNMP engine includes several subsystems communicating via defined abstract service interfaces.
 Abstract service interfaces describe the conceptual interfaces between various subsystems.
@@
 
-According to the SNMP architecture defined in \cite{rfc3411}, an SNMP entity consists of an SNMP engine and one or more associated applications.
-The SNMP engine is composed of a dispatcher, a message processing subsystem, a security subsystem, an access control subsystem, and a transport subsystem (Figure \ref{fig:snmparch}).
+According to the SNMP architecture defined in \cite{rfc3411}, the SNMP engine is composed of a dispatcher, a message processing subsystem, a security subsystem, an access control subsystem, and a transport subsystem (Figure \ref{fig:snmparch}).
 Each subsystem may include multiple concrete models implementing different services.
@@
 An entity containing one or more command generator and notification receiver applications is called an SNMP manager.
+\section{Message Format}
+The SNMPv3 message (Figure \ref{fig:snmp-format}) contains the SNMP version, fields for global data (such as the message identifier, the maximum message size, the security model and the level of security), fields for the security model information and naming scope (context identifier and name), and finally the protocol data unit (PDU).
+\begin{figure}[htp]	
+\begin{center}
+    \includegraphics[scale = 0.60]{img/snmpv3-message.pdf}
+    \caption{SNMPv3 message format}   
+	\label{fig:snmp-format}
+\end{center}
+\end{figure}
 
-\section{Operations}
+The SNMP Version field identifies the version of the message processing.
+The message identifier (msgID) is used between two SNMP entities to coordinate request messages and responses and to coordinate the message processing by different subsystem models.
+The maximum message size (Max Size) is the maximum message size supported by the sender of the message, i.e., the maximum message size that the sender can accept when another SNMP engine sends an SNMP message.
+The Flag field contains several bit fields which control processing of the message.
+The security model field enables the concurrent support of multiple security models identifying which security model was used by the sender to generate the message.
+The security parameters field is used for communication between the security model modules in the sending and receiving SNMP engines.
+The contents and format of the data in this field are defined by the security model.
+
+The naming scope contains a PDU and the context in which it has to be processed.
+The context engineID identifies the engine  and the context name defines the particular context associated with the management information contained in the PDU of the message.
+These fields are followed by the PDU data.
+
+\section{Protocol Operations}
 
 \section{Security Models}
-
-\section{SMI and MIB}

### 2010-01-29 trunk/doc/msc-proposal/msc-snmp.tex
proposal: snmp operations

@@
 An entity containing one or more command generator and notification receiver applications is called an SNMP manager.
-\section{Message Format}
+
+\section{Message Format}\label{sec:messformat}
 The SNMPv3 message (Figure \ref{fig:snmp-format}) contains the SNMP version, fields for global data (such as the message identifier, the maximum message size, the security model and the level of security), fields for the security model information and naming scope (context identifier and name), and finally the protocol data unit (PDU).
@@
 The naming scope contains a PDU and the context in which it has to be processed.
-The context engineID identifies the engine  and the context name defines the particular context associated with the management information contained in the PDU of the message.
+The context engineID identifies the engine which realizes the managed objects referenced within the PDU, and the context name defines the particular context associated with the management information contained in the PDU of the message.
 These fields are followed by the PDU data.
@@
 \section{Protocol Operations}
+As described in Section \ref{sec:messformat}, an SNMP message encapsulates a PDU which specifies the operations performed by the receiving SNMP engine.
+The SNMP protocol defines PDUs for the following operations: Get, GetNext, GetBulk, Response, Set, Trap, Inform.
+
+In addition, the Report protocol operation can be used internally to allow SNMP engines to communicate with each other for error notifications and clock synchronization.
 

### 2010-01-30 doc/msc-proposal/msc-snmp.tex
proposal: snmp PDU&operations

@@
 This makes it a perfect candidate as a management protocol for 6LoWPAN applications.
-In this chapter an overview of the SNMP protocol is given.
+In this chapter an overview of the SNMPv3 protocol is given.
+
+\section{Management Information Base}
+SNMP itself does not specify which information is offered by a managed system.
+Management information which can be accessed via SNMP is viewed as a collection of managed objects in a virtual information store, called the Management Information Base (MIB).
+MIBs describe the structure of the device's management information using a hierarchical namespace containing object identifiers (OID).
+Each OID identifies a variable that can be read or set via SNMP.
+Collections of related objects are defined in MIB modules written in the Structure of Management Information (SMI) data definition language, which is an adapted subset of the Abstract Syntax Notation One (ASN.1) language.
+A number of specific MIBs has been defined as part of the SNMP suite specifying managed objects for the most common network management subjects.
 
@@
 
-According to the SNMP architecture defined in \cite{rfc3411}, an SNMP entity consists of an SNMP engine and one or more associated applications.
-The SNMP engine contains  a dispatcher, a message processing subsystem, a security subsystem, an access control subsystem, and a transport subsystem (Figure \ref{fig:snmparch}).
+According to the SNMP architecture defined in \cite{rfc3411}, the SNMP engine is composed of a dispatcher, a message processing subsystem, a security subsystem, an access control subsystem, and a transport subsystem (Figure \ref{fig:snmparch}).
 Each subsystem may include multiple concrete models implementing different services.
-The dispatcher is responsible for controlling the data flow within an SNMP entity.
-Extracting data from received messages and preparing messages for sending are accomplished in the message processing subsystem.
+The dispatcher is the key part of an SNMP engine.
+It's responsible for controlling the data flow within an SNMP entity.
+When an SNMP message needs to be prepared or when data needs to be extracted from an SNMP message, the dispatcher delegates these tasks to a version-specific message processing model within the message processing subsystem.
+It also dispatches SNMP  protocol data units (PDUs) to SNMP applications.
 Each message processing model defines the format of a particular version of an SNMP message.
@@
 An SNMP entity which includes one or more command generator and notification receiver applications is called an SNMP manager.
-An entity containing one or more command generator and notification receiver applications is called an SNMP manager.
+An SNMP entity containing one or more command responder and notification originator applications is called an SNMP agent.
 
-\section{Operations}
+\section{Message Format}\label{sec:messformat}
+The SNMPv3 message (Figure \ref{fig:snmp-format}) contains the SNMP version, fields for global data (such as the message identifier, the maximum message size, the security model and the level of security), fields for the security model information and naming scope (context identifier and name), and finally the protocol data unit (PDU).
+\begin{figure}[htp]	
+\begin{center}
+    \includegraphics[scale = 0.60]{img/snmpv3-message.pdf}
+    \caption{SNMPv3 Message Format}   
+	\label{fig:snmp-format}
+\end{center}
+\end{figure}
+
+The SNMP Version field identifies the version of the message processing.
+The message identifier (msgID) is used between two SNMP entities to coordinate request messages and responses and to coordinate the message processing by different subsystem models.
+The maximum message size (Max Size) is the maximum message size supported by the sender of the message, i.e., the maximum message size that the sender can accept when another SNMP engine sends an SNMP message.
+The Flag field contains several bit fields which control processing of the message.
+The security model field enables the concurrent support of multiple security models identifying which security model was used by the sender to generate the message.
+The security parameters field is used for communication between the security model modules in the sending and receiving SNMP engines.
+The contents and format of the data in this field are defined by the security model.
+
+The naming scope contains a PDU and the context in which it has to be processed.
+The context engineID identifies the engine which realizes the managed objects referenced within the PDU, and the context name defines the particular context associated with the management information contained in the PDU of the message.
+These fields are followed by the PDU data.
+
+The common format of protocol data units is shown in Figure \ref{fig:snmp-pdu}.
+This format is used for all operations described in Section \ref{sec:opetations}, except for the GetBulk request.
+The PDU Type field has an integer value which identifies the type of the PDU.
+The Request Identifier is used to match requests with replies.
+It is generated by the device that sends a request and copied into this field in a response PDU by the responding SNMP entity.
+The Error Status tells the requesting SNMP entity the result of its request.
+A value of zero indicates that no error occurred; the other values indicate what sort of error happened.
+When Error Status is non-zero, the Error Index field points to the object generated the error, and has the value of zero in a request.
+The variable bindings is a set of name-value pairs identifying the MIB objects in the PDU, and in the case of messages other than requests, containing their values.
+\begin{figure}[htp]	
+\begin{center}
+    \includegraphics[scale = 0.60]{img/snmp-pdu.pdf}
+    \caption{SNMP Common PDU Format}   
+	\label{fig:snmp-pdu}
+\end{center}
+\end{figure}
+
+\section{Protocol Operations}\label{sec:opetations}
+As described in Section \ref{sec:messformat}, an SNMP message encapsulates a PDU which specifies the operations performed by the receiving SNMP engine.
+The SNMP protocol defines PDU formats for the following operations: Get, GetNext, GetBulk, Response, Set, Trap, Inform and Report (Figure \ref{fig:snmp-operations}).
+
+\begin{figure}[htp]	
+\begin{center}
+	\label{fig:snmp-operations}
+    \includegraphics[scale = 0.80]{img/snmp-operations.pdf}
+    \caption{SNMP Operations}   
+\end{center}
+\end{figure}
+
+The Get request is initiated by a manager. Upon receipt of the Get request, the agent processes each variable binding in the variable-binding list and produces a Response.
+
+The Report protocol operation can be used internally to allow SNMP engines to communicate with each other for error notifications and clock synchronization.
 

### 2010-01-30 doc/msc-proposal/msc-snmp.tex
proposal: snmp operations

@@
 This format is used for all operations described in Section \ref{sec:opetations}, except for the GetBulk request.
-The PDU Type field has an integer value which identifies the type of the PDU.
-The Request Identifier is used to match requests with replies.
+The PDU type field has an integer value which identifies the type of the PDU.
+The request identifier is used to match requests with replies.
 It is generated by the device that sends a request and copied into this field in a response PDU by the responding SNMP entity.
-The Error Status tells the requesting SNMP entity the result of its request.
+The error status tells the requesting SNMP entity the result of its request.
 A value of zero indicates that no error occurred; the other values indicate what sort of error happened.
-When Error Status is non-zero, the Error Index field points to the object generated the error, and has the value of zero in a request.
+When error status is non-zero, the error index field points to the object generated the error, and has the value of zero in a request.
 The variable bindings is a set of name-value pairs identifying the MIB objects in the PDU, and in the case of messages other than requests, containing their values.
@@
 
-The Get request is initiated by a manager. Upon receipt of the Get request, the agent processes each variable binding in the variable-binding list and produces a Response.
+The Get request initiated by a manager allows to access and retrieve the value of one or more instances of management information.
+The agent processes each variable binding in the variable-binding list of the received Get PDU by resolving their values and produces a Response PDU.
+The response PDU is sent back to the manager.
+If the processing of any variable binding fails the details of the error are reflected in the the error status and error index fields in the Response PDU.
+The GetNext operation retrieves the value of the next existing MIB instances in lexicographic order.
+Successive GetNext operations can be used to walk the MIB instances without prior knowledge about the MIB structure.
+The GetBulk operation is a generalization of the GetNext operation where the agent performs a series of GetNext operations internally.
+This makes it easier to acquire large amounts of related information without initiating repeated GetNext operations.
+The manager can set values for the managed objects by using the Set operation.
 
-The Report protocol operation can be used internally to allow SNMP engines to communicate with each other for error notifications and clock synchronization.
+The Trap is sent by an agent to a manager and indicates that an event has occurred or a certain condition is present.
+There is no confirmation associated with this notification delivery mechanism for this operation.
+The Inform operation provides a confirmed notification delivery mechanism.
+Finally, the Report operation can be used internally to allow SNMP engines to communicate with each other for error notifications and clock synchronization.
 

### 2010-01-31 doc/msc-proposal/msc-snmp.tex
proposal: snmp operations

@@
 The Trap is sent by an agent to a manager and indicates that an event has occurred or a certain condition is present.
-There is no confirmation associated with this notification delivery mechanism for this operation.
-The Inform operation provides a confirmed notification delivery mechanism.
+There is no confirmation associated with this notification delivery mechanism.
+The confirmed notification delivery mechanism is provided by the Inform operation.
 Finally, the Report operation can be used internally to allow SNMP engines to communicate with each other for error notifications and clock synchronization.

### 2010-01-31 trunk/doc/msc-proposal/msc-proposal.tex
proposal: snmp operations

@@
 The target hardware platform will be the AVR Raven board which has a 2.4 GHz transceiver and uses an 8-bit ATmega1284P micro-controller with 16KB of RAM and 128KB of flash memory.
-The implementation will be done on top of Contiki 6LoWPAN implementation.
+The implementation will be done on top of the Contiki 6LoWPAN implementation.
 
@@
 In the first phase, the basic SNMPv1 protocol will be implemented.
-In the second phase, the required security primitives will be addressed and implemented, in case there will not be available ready-to-use solutions.
-Finally, a security layer will be added to the SNMP implementation.
+Then, the focus will be shifted to the creation of a secure network layer for Contiki.
+Finally, a security layer will be added to the SNMPv1 implementation.
+
+%Network Management: One of the points of transmitting IPv6
+%       packets is to reuse existing protocols as much as possible.
+%       Network management functionality is critical for LoWPANs.
+%       However, management solutions need to meet the resource
+%       constraints as well as the minimal configuration and self-healing
+%       functionality described in Section 4.4. The Simple Network
+%       Management Protocol (SNMP) [RFC3410] is widely used for
+%       monitoring data sources and sensors in conventional networks.
+%       SNMP functionality may be translated "as is" to LoWPANs with the
+%       benefit to utilize existing tools.  However, due to the memory,
+%       processing, and message size constraints, further investigation
+%       is required to determine if the use of SNMPv3 is suitable, or if
+%       an appropriate adaptation of SNMPv3 or use of different protocols
+%       is in order.
 

### 2010-01-31 trunk/doc/msc-proposal/msc-snmp.tex
proposal: SNMP security models

@@
 In this chapter an overview of the SNMPv3 protocol is given.
+
+\section{Management Information Base}
+SNMP itself does not specify which information is offered by a managed system.
+Management information which can be accessed via SNMP is viewed as a collection of managed objects in a virtual information store, called the Management Information Base (MIB).
+MIBs describe the structure of the device's management information using a hierarchical namespace containing object identifiers (OID).
+Each OID identifies a variable that can be read or set via SNMP.
+Collections of related objects are defined in MIB modules written in the Structure of Management Information (SMI) data definition language, which is an adapted subset of the Abstract Syntax Notation One (ASN.1) language.
+A number of specific MIBs has been defined as part of the SNMP suite specifying managed objects for the most common network management subjects.
 
@@
 An SNMP entity which includes one or more command generator and notification receiver applications is called an SNMP manager.
-An entity containing one or more command generator and notification receiver applications is called an SNMP manager.
+An SNMP entity containing one or more command responder and notification originator applications is called an SNMP agent.
 
@@
     \includegraphics[scale = 0.60]{img/snmpv3-message.pdf}
-    \caption{SNMPv3 message format}   
+    \caption{SNMPv3 Message Format}   
 	\label{fig:snmp-format}
@@
 
-\section{Protocol Operations}
+The common format of protocol data units is shown in Figure \ref{fig:snmp-pdu}.
+This format is used for all operations described in Section \ref{sec:opetations}, except for the GetBulk request.
+The PDU type field has an integer value which identifies the type of the PDU.
+The request identifier is used to match requests with replies.
+It is generated by the device that sends a request and copied into this field in a response PDU by the responding SNMP entity.
+The error status tells the requesting SNMP entity the result of its request.
+A value of zero indicates that no error occurred; the other values indicate what sort of error happened.
+When error status is non-zero, the error index field points to the object generated the error, and has the value of zero in a request.
+The variable bindings is a set of name-value pairs identifying the MIB objects in the PDU, and in the case of messages other than requests, containing their values.
+\begin{figure}[htp]	
+\begin{center}
+    \includegraphics[scale = 0.60]{img/snmp-pdu.pdf}
+    \caption{SNMP Common PDU Format}   
+	\label{fig:snmp-pdu}
+\end{center}
+\end{figure}
+
+\section{Protocol Operations}\label{sec:opetations}
 As described in Section \ref{sec:messformat}, an SNMP message encapsulates a PDU which specifies the operations performed by the receiving SNMP engine.
-The SNMP protocol defines PDUs for the following operations: Get, GetNext, GetBulk, Response, Set, Trap, Inform.
+The SNMP protocol defines PDU formats for the following operations: \texttt{Get}, \texttt{GetNext}, \texttt{GetBulk}, \texttt{Response}, \texttt{Set}, \texttt{Trap}, \texttt{Inform} and \texttt{Report} (Figure \ref{fig:snmp-operations}).
 
-In addition, the Report protocol operation can be used internally to allow SNMP engines to communicate with each other for error notifications and clock synchronization.
+\begin{figure}[htp]	
+\begin{center}
+	\label{fig:snmp-operations}
+    \includegraphics[scale = 0.80]{img/snmp-operations.pdf}
+    \caption{SNMP Operations}   
+\end{center}
+\end{figure}
+
+The \texttt{Get} request initiated by a manager allows to access and retrieve the value of one or more instances of management information.
+The agent processes each variable binding in the variable-binding list of the received \texttt{Get} PDU by resolving their values and produces a \texttt{Response} PDU.
+The \texttt{Response} PDU is sent back to the manager.
+If the processing of any variable binding fails the details of the error are reflected in the the error status and error index fields in the \texttt{Response} PDU.
+The \texttt{GetNext} operation retrieves the value of the next existing MIB instances in lexicographic order.
+Successive \texttt{GetNext} operations can be used to walk the MIB instances without prior knowledge about the MIB structure.
+The \texttt{GetBulk} operation is a generalization of the \texttt{GetNext} operation where the agent performs a series of \texttt{GetNext} operations internally.
+This makes it easier to acquire large amounts of related information without initiating repeated \texttt{GetNext} operations.
+The manager can set values for the managed objects by using the \texttt{Set} operation.
+
+The \texttt{Trap} is sent by an agent to a manager and indicates that an event has occurred or a certain condition is present.
+There is no confirmation associated with this notification delivery mechanism.
+The confirmed notification delivery mechanism is provided by the \texttt{Inform} operation.
+Finally, the \texttt{Report} operation can be used internally to allow SNMP engines to communicate with each other for error notifications and clock synchronization.
 
 \section{Security Models}
+SNMPv1 and SNMPv2c  provide only trivial security based on a clear-text community string and lack of adequate mechanisms for securing the management function.
+The security problems are addressed in SNMPv3 where the User-based Security Model (USM) \cite{rfc3414} is defined that implements the security services for authentication and privacy.
+The USM provides three levels of security:
+\begin{enumerate}
+\item[1)] \texttt{noAuthNoPriv} mode provides no authentication and no encryption services.
+\item[2)] \texttt{authNoPriv} mode provides message authentication, message integrity, and timeliness checking services but no encryption.
+\item[3)] \texttt{authPriv} mode message provides authentication, message integrity, timeliness checking services and encryption of the payload.
+\end{enumerate}
+
+The document \cite{rfc5591} introduces a Transport Security Model for the SNMP protocol that makes use of existing and commonly deployed security infrastructures and Transport Models providing lower-layer, secure transports.

### 2010-02-04 doc/msc-proposal/msc-proposal.tex
proposal: final corrections

@@
 Simple Network Management Protocol (SNMP) is a widely deployed application protocol for network management and data retrieval.
-It can be very lightweight and may fit resource-constrained devices very well.
+Its implementation can be very lightweight and may fit resource-constrained devices very well.
 The focus of the proposed work will be implementation and evaluation of the SNMP protocol for devices supporting IEEE 802.15.4 Radios and the Contiki Embedded Operating System.
@@
 In Chapter \ref{ch:routing} 6LoWPAN routing problems are discussed.
-The 6LoWPAN Neighbor Discovery Protocol is covered in Chapter \ref{ch:nd}.
+The 6LoWPAN Neighbour Discovery Protocol is covered in Chapter \ref{ch:nd}.
 Chapter \ref{ch:snmp} gives an overview of the Simple Network Management Protocol.
-Chapter \ref{ch:proposal} describes details of the proposed work to be done in the scope of the Master Thesis.
+Chapter \ref{ch:proposal} describes the details of the proposed work to be done in the scope of the Master Thesis.
 
@@
 
-Packets with a multicast IPv6 destination address are sent to the 16-bit 802.15.4 address obtained by concatenating the 3-bit multicast prefix 101, the last 5 bits in the 15-th octet and the whole 16-th octet of the multicast IPv6 address.
+Packets with a multicast IPv6 destination address are sent to the 16-bit 802.15.4 address obtained by concatenating the 3-bit multicast prefix 101, the last 5 bits in the 15th octet and the whole 16th octet of the multicast IPv6 address.
 
@@
 In the first phase, the basic SNMPv1 protocol will be implemented.
-Then, the focus will be shifted to creation of a secure network layer for Contiki.
+Then, the focus will be shifted to the creation of a secure network layer for Contiki.
 Finally, a security layer will be added to the SNMPv1 implementation.

### 2010-03-03 trunk/src/snmpd.c
source code initiat commit

@@ -1,0 +1,71 @@
+#include "contiki.h"
+#include "contiki-lib.h"
+#include "contiki-net.h"
+#include <string.h>
+
+#define UDP_IP_BUF   ((struct uip_udpip_hdr *)&uip_buf[UIP_LLH_LEN])
+
+char *help_buf = "aaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaa\n"
+				 "bbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbb\n"
+				 "cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc\n"
+				 "dddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddd\n"
+				 "aaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaa\n"
+				 "bbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbb\n"
+				 "cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc\n"
+				 "dddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddd\n"
+				 "aaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaa\n"
[…]

### 2010-03-03 doc/msc-thesis/msc-nd.tex
initial msc thesis report

--- 2009-12-19 doc/msc-proposal/msc-nd.tex
@@
 
-The Internet Draft \cite{draft-nd-07} specifies an optimized neighbor discovery mechanism sufficient for LoWPAN operation.
+The internet draft \cite{draft-nd-07} specifies an optimized neighbor discovery mechanism sufficient for LoWPAN operation.
 The 6LoWPAN Neighbor Discovery (6LoWPAN-ND) introduces a node registration mechanism optimizing the node-router interface, which requires no flooding and reduces link-local multicast frequency.
-The concept of a LoWPAN Whiteboard located at Edge Routers (routers that connect a LoWPAN to another IP network) is introduced, which allows for Duplicate Address Detection for the entire LoWPAN.
+The concept of a LoWPAN Whiteboard located at Edge Routers (ERs - routers that connect a LoWPAN to another IP network) is introduced, which allows for Duplicate Address Detection for the entire LoWPAN.
 The solution supports both Mesh Under and route over configurations for multihop forwarding.
@@
 Therefore,  knowledge of the 64-bit address of a node is enough to derive its link-local address and reach it on the same link.
-In order to receive RAs from routers, a node joins the all-nodes multicast address and, if the node is a router, the all-routers multicast address.
-Once the interfaces have been initialized, a node listens for Router Advertisements (RA), or broadcasts a Router Solicitation (RS).
+In order to receive Router Advertisements (RAs) from routers, a node joins the all-nodes multicast address and, if the node is a router, the all-routers multicast address.
+Once the interfaces have been initialized, a node listens for RAs, or broadcasts a Router Solicitation (RS).
 Upon receipt of the RA, the node forms an optimistic global unique address with stateless address autoconfiguration and chooses one or more default routers (routers are present only in a route over configuration).
@@
 Once the node has received the NC message, it is capable to send packets to any IPv6 address inside or outside the LoWPAN.
-Each router maintains a set of information about nodes that are currently registered through it, called the binding table.
+Each router maintains information about nodes that are currently registered through it, called the binding table.
 Nodes send periodic registration messages in order to maintain their bindings in the router binding tables and Edge Router Whiteboard.
@@
 RAs are sent either to all-nodes multicast or to a link-local unicast address as a response to an RS.
-RA messages use additional 6LoWPAN Information and 6LoWPAN Summary Options described in Sections \ref{nd.option.info} and \ref{nd.option.summary} respectively.
+RA messages use additional 6LoWPAN Information and 6LoWPAN Summary Options described in Section \ref{nd.option.info}.
 
 \subsection{Neighbor Solicitation/Advertisement Message}
-Neighbor Solicitation(NS) and Neighbor Advertisement(NA) messages are employed only between ERs on the backbone link in Extended LoWPAN.
+Neighbor Solicitation(NS) and Neighbor Advertisement(NA) messages are employed only between ERs on the backbone link in an Extended LoWPAN.
 Extended LoWPAN (shown in Figure \ref{fig:ext.lowpan})) allows to form a single subnet out of multiple LoWPANs interconnected by a backbone link via Edge Routers.
-The format of NS/NA messages is identical to the classic NS/NA messages \cite{rfc4861} with additional use of Owner Interface Identifier Option, described in Section \ref{sec:nd.oiio}.
+The format of NS/NA messages is identical to the classic NS/NA messages \cite{rfc4861} with additional use of the Owner Interface Identifier Option, described in Section \ref{sec:nd.oiio}.
 
@@
 The Status field is used only in NC messages and specifies the result of the registration procedure.
-A 4-bit unsigned Code provides additional details for the message, e.g.
-it may indicate whether the NR/NC is sent directly or relayed by a router.
+A 4-bit unsigned Code provides additional details for the message, e.g., it may indicate whether the NR/NC is sent directly or relayed by a router.
 The Checksum field carries the ICMP checksum.
@@
 \subsubsection{6LoWPAN Address Option}\label{nd.option.address}
-The 6LoWPAN Address Option (6AO) is used in NR and NC messages to indicate the address which a node wants to register with an ER and to get the result of registration back.
+The 6LoWPAN Address Option (6AO) is used in NR and NC messages to indicate the address which a node wants to register with an ER and to get the result of the registration back.
 A NR/NC message can include multiple Address Options.
@@
 The entries of the list are associated with the sequence numbers last advertised in the 6LoWPAN Summary Option, which allows to detect the changes in prefix information.
-In order to support stateful address compression, a node holds a list of context.
+In order to support stateful address compression, a node holds a list of contexts.
 As in the case of prefixes, each context entry is associated with the 6LoWPAN Summary Option sequence number.
@@
 
-In addition to the mentioned data structures, a node also keeps a set of conceptual variables for each interface, such as the default hop limit and the binding lifetime.
+In addition to the mentioned data structures, a node also keeps a set of conceptual variables for each interface, e.g., the default hop limit and the binding lifetime.
 
@@
 \subsection{LoWPAN Edge Router}
-Edge Routers implement LoWPAN Router features and extend them with Whiteboard registration for LoWPAN Nodes within their subnets.
-For every registered node an ER maintains the IPv6 address, Owner Interface Identifier, .
+Edge Routers implement LoWPAN Router features and extend them with whiteboard registration for LoWPAN Nodes within their subnets.
+An ER maintains information about every registered node, e.g., the IPv6 address and Owner Interface Identifier of the node.
 There is no link-layer information stored in the Whiteboard.

### 2010-03-03 doc/msc-thesis/msc-routing.tex
initial msc thesis report

--- 2010-01-17 doc/msc-proposal/msc-routing.tex
@@
 The value "fail" corresponds to not acceptable performance, which means that the protocol does not meet the criterion.
-Finally, a "?" means that a protocol would require a supplementary document specifying how a protocol should behave.
+Finally, a question mark"?" means that a protocol would require a supplementary document specifying how a protocol should behave.
 As can be seen from the table, no existing IETF protocol meets all described criteria.
@@
 \section{Overview of IPv6 Routing Protocol for LLNs}\label{sec:rout.rpl}
-The IPv6 Routing Protocol for Low power and Lossy Networks (RPL) introduces a notion of a Directed Acyclic Graph (DAG), which is a directed graph with no cycles exist.
+Typically, the traffic patterns for LLNs are not simply unicast, but in many cases point-to-multipoint or multipoint-to-point.
+The IPv6 Routing Protocol for Low power and Lossy Networks (RPL) \cite{draft-rpl-04} is designed to meet the routing requirements for LLNs and supports multipoint-to-point traffic from devices inside the LLN towards a central control point, as well as point-to-multipoint traffic from the central control point to the devices inside the LLN.
+
+The RPL introduces a notion of a Directed Acyclic Graph (DAG), which is a directed graph with no cycles.
 All edges of a DAG are oriented toward and terminating at one or more root nodes uniquely identified by the DAGID.
 A Destination Oriented DAG (DODAG) is a DAG rooted at a single destination.
-While constructing a DODAG various routing metrics and optimization objectives can be used to compute the shortest paths.
+A DODAG can be constructed with various routing metrics and optimization objectives in use to compute the shortest paths.
 The set of supported link and node metrics used in the RPL is specified in the internet draft \cite{draft-routing-metrics-04}.
 
-A DAG instance is a set of possibly multiple DODAGs.
+A DAG instance may consist of multiple DODAGs.
 The routing metrics and objectives for a certain DAG instance are specified by an Objective Function (OF).
@@
 \subsection{Basic Operation}
-A node's neighbor set is the nodes that it can reach with a link-local multicast.
+A set of nodes that can be reached with a link-local multicast forms the node's neighbor set.
 A node that is not a DODAG root may maintain multiple DAG parents for a single DAG instance.
-DAG parents of a node is a subset of the set of       its neighbors.
-A node may implement different policies for selecting its DAG parents.
+DAG parents of a node is a subset of the set of its neighbors and can be selected using different policies.
 

### 2010-03-03 src/logging.h
added logging header

@@ -1,0 +1,8 @@
+/**
+ * \file
+ *         Logging facilites for the SNMP server
+ * \author
+ *         Siarhei Kuryla <kurilo@gmail.com>
+ *
+ * @(#)$Id: mtarch.h,v 1.2 2006/12/22 16:55:53 barner Exp $
+ */

### 2010-03-03 trunk/src/snmpd.c
added logging header

@@ -5,22 +5,11 @@
 
 #define UDP_IP_BUF   ((struct uip_udpip_hdr *)&uip_buf[UIP_LLH_LEN])
 
-char *help_buf = "aaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaa\n"
-				 "bbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbb\n"
-				 "cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc\n"
-				 "dddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddd\n"
-				 "aaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaa\n"
-				 "bbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbb\n"
-				 "cccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccccc\n"
-				 "dddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddddd\n"
-				 "aaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaaa\n"
-				 "bbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbbb\n"
[…]

### 2010-03-05 src/snmpd-logging.h
implemented sceleton

@@ -1,0 +1,25 @@
+/**
+ * \file
+ *         Logging facilites for the SNMP server
+ * \author
+ *         Siarhei Kuryla <kurilo@gmail.com>
+ */
+
+#ifndef __SNMPD_LOGGING_H__
+#define __SNMPD_LOGGING_H__
+
+/** \brief indicates whether debug is enabled */
+#define DEBUG 1
+
+/**
+ * Log a message.
+ *
+ * \param m A pointer to a string containing a message.
+ *
+ * \sa log()
+ *
+ * \hideinitializer
+ */
+void snmp_log(char* format, ...);
+
+#endif /* __SNMPD_LOGGING_H__ */

### 2010-03-05 src/snmpd-utils.h

@@ -1,0 +1,16 @@
+/**
+ * \file
+ *         Utility facilites for the SNMP server
+ * \author
+ *         Siarhei Kuryla <kurilo@gmail.com>
+ */
+
+#ifndef __SNMPD_UTILS_H__
+#define	__SNMPD_UTILS_H__
+
+#define min(a,b) ((a>b) ? b : a)
+#define max(a,b) ((a>b) ? a : b)
+
+
+#endif	/* __SNMPD_UTILS_H__ */
+

### 2010-03-05 trunk/src/logging.h
implemented sceleton

@@ -1,8 +1,0 @@
-/**
- * \file
- *         Logging facilites for the SNMP server
- * \author
- *         Siarhei Kuryla <kurilo@gmail.com>
- *
- * @(#)$Id: mtarch.h,v 1.2 2006/12/22 16:55:53 barner Exp $
- */

### 2010-03-05 trunk/src/snmp-protocol.c

@@ -1,0 +1,112 @@
+#include "snmp-protocol.h"
+#include "snmpd-logging.h"
+
+/*-----------------------------------------------------------------------------------*/
+/*
+ * Decode a BER encoded SNMP message.
+ */
+
+static u8_t fetch_type(const u8_t* const request, const u16_t* const len, u16_t* pos, u8_t* type)
+{
+    if (*pos < *len) {
+        switch (request[*pos]) {
+            case BER_TYPE_BOOLEAN:
+            case BER_TYPE_INTEGER:
+            case BER_TYPE_BIT_STRING:
+            case BER_TYPE_OCTET_STRING:
+            case BER_TYPE_NULL:
+            case BER_TYPE_OID:
+            case BER_TYPE_SEQUENCE:
+            case BER_TYPE_IPADDRESS:
+            case BER_TYPE_COUNTER:
+            case BER_TYPE_GAUGE:
+            case BER_TYPE_TIME_TICKS:
+            case BER_TYPE_OPAQUE:
+            case BER_TYPE_NASAPADDRESS:
+            case BER_TYPE_COUNTER64:
+            case BER_TYPE_UINTEGER32:
+            case BER_TYPE_NO_SUCH_OBJECT:
+            case BER_TYPE_NO_SUCH_INSTANCE:
+            case BER_TYPE_END_OF_MIB_VIEW:
+            case BER_TYPE_SNMP_GET:
+            case BER_TYPE_SNMP_GETNEXT:
+            case BER_TYPE_SNMP_RESPONSE:
+            case BER_TYPE_SNMP_SET:
+            case BER_TYPE_SNMP_GETBULK:
+            case BER_TYPE_SNMP_INFORM:
+            case BER_TYPE_SNMP_TRAP:
+                *type = request[*pos];
+                *pos = *pos + 1;
+                break;
+            default:
+                //snmp_log("unsupported BER type %02X\n", request[*pos]);
+                return -1;
+        }
+    } else {
+            //snmp_log("malformed SNMP request\n");
+            return -1;
+    }
+    return 0;
+}
+
+
+/*-----------------------------------------------------------------------------------*/
+/*
+ * Decode a BER encoded SNMP message.
+ */
+u8_t snmp_decode_request(const u8_t* const request, const u16_t* const len)
+{
+    static u16_t pos = 0;
+    static u8_t type, res;
+
[…]

### 2010-03-05 trunk/src/snmp-protocol.h

@@ -1,0 +1,45 @@
+/**
+ * \file
+ *         Implementation of the SNMP protocol
+ * \author
+ *         Siarhei Kuryla <kurilo@gmail.com>
+ */
+
+#ifndef __SNMP_PROTOCOL_H__
+#define	__SNMP_PROTOCOL_H__
+
+#include "contiki.h"
+/*
+ * BER identifiers for ASN.1 implementation of SNMP.
+ */
+#define BER_TYPE_BOOLEAN                                0x01
+#define BER_TYPE_INTEGER                                0x02
+#define BER_TYPE_BIT_STRING                             0x03
+#define BER_TYPE_OCTET_STRING                           0x04
+#define BER_TYPE_NULL                                   0x05
+#define BER_TYPE_OID                                    0x06
+#define BER_TYPE_SEQUENCE                               0x30
+#define BER_TYPE_IPADDRESS                              0x40
+#define BER_TYPE_COUNTER                                0x41
+#define BER_TYPE_GAUGE                                  0x42
+#define BER_TYPE_TIME_TICKS				0x43
+#define BER_TYPE_OPAQUE                                 0x44
+#define BER_TYPE_NASAPADDRESS                           0x45
+#define BER_TYPE_COUNTER64                              0x46
+#define BER_TYPE_UINTEGER32                             0x47
+#define BER_TYPE_NO_SUCH_OBJECT                         0x80
+#define BER_TYPE_NO_SUCH_INSTANCE                       0x81
+#define BER_TYPE_END_OF_MIB_VIEW                        0x82
+#define BER_TYPE_SNMP_GET                               0xA0
+#define BER_TYPE_SNMP_GETNEXT                           0xA1
+#define BER_TYPE_SNMP_RESPONSE                          0xA2
+#define BER_TYPE_SNMP_SET                               0xA3
+#define BER_TYPE_SNMP_GETBULK                           0xA5
+#define BER_TYPE_SNMP_INFORM                            0xA6
+#define BER_TYPE_SNMP_TRAP                              0xA7
+#define BER_TYPE_SNMP_REPORT                            0xA8
+
+u8_t snmp_decode_request(const u8_t* const data,  const u16_t* const len);
+
+#endif	/* __SNMP_PROTOCOL_H__ */
+

### 2010-03-05 trunk/src/snmpd-logging.c

@@ -1,0 +1,57 @@
+/**
+ * \file
+ *         UDP implementatio of logging facilites for the SNMP server
+ * \author
+ *         Siarhei Kuryla <kurilo@gmail.com>
+ */
+#include "snmpd-logging.h"
+
+#ifdef DEBUG
+#include <stdio.h>
+#include <stdarg.h>
+#include <string.h>
+
+#include "contiki-net.h"
+
+/** \brief port number where debug messages are sent */
+#define LOGGING_PORT 12345
+
+/** \brief length of the buffer used for debugging messages */
+#define BUF_LEN 100
+
+/*--------------------------------------------------------------------------*/
+/*
+ * Log a debug message by sending it within a UDP message.
+ */
+void snmp_log(char* format, ...)
+{  
+    static struct uip_udp_conn *udp_con = NULL;
+    if (udp_con == NULL) {
+        udp_con = udp_new(NULL, HTONS(LOGGING_PORT), NULL);
+        uip_ip6addr(&udp_con->ripaddr, 0xaaaa, 0x0000, 0x0000, 0x0000, 0x0000, 0x0000, 0x0000, 0x0001);
+	udp_bind(udp_con, HTONS(3000));
+    }
+
+/*    static char buf[BUF_LEN];
+    memset(buf, 0, BUF_LEN);
+    va_list args;
+    va_start(args, format);
+    sprintf(buf, format, args);
+    va_end(args);*/
+
+    va_list args;
+    va_start(args, format);
+    char buf[100];
+    memset(buf, 0, 100);
+    vsprintf(buf, "type: %02x, pos: %02x, type: %02x\n", args);
+    va_end(args);
+
+    //uip_udp_packet_send(udp_con, format, strlen(format));
+    uip_udp_packet_send(udp_con, buf, strlen(buf));
+
+}
+#else
+void snmp_log(char* format, ... )
+{
+}
+#endif /* DEBUG */

### 2010-03-05 trunk/src/snmpd-utils.c

@@ -1,0 +1,1 @@
+#include "snmpd-utils.h"

### 2010-03-05 trunk/src/snmpd.c

@@ -1,65 +1,43 @@
+#include <stdio.h>
+#include <string.h>
+
 #include "contiki.h"
 #include "contiki-lib.h"
 #include "contiki-net.h"
-#include <string.h>
 
-#define UDP_IP_BUF   ((struct uip_udpip_hdr *)&uip_buf[UIP_LLH_LEN])
-
-#define PRINT6ADDR(addr) "%02x%02x:%02x%02x:%02x%02x:%02x%02x:%02x%02x:%02x%02x:%02x%02x:%02x%02x", ((u8_t *)addr)[0], ((u8_t *)addr)[1], ((u8_t *)addr)[2], ((u8_t *)addr)[3], ((u8_t *)addr)[4], ((u8_t *)addr)[5], ((u8_t *)addr)[6], ((u8_t *)addr)[7], ((u8_t *)addr)[8], ((u8_t *)addr)[9], ((u8_t *)addr)[10], ((u8_t *)addr)[11], ((u8_t *)addr)[12], ((u8_t *)addr)[13], ((u8_t *)addr)[14], ((u8_t *)addr)[15]
-				 
-#define min(a,b) ( (a>b) ? b : a )
-#define max(a,b) ( (a>b) ? a : b )
-
-#define UDP_DATA_LEN 10000
+#include "snmpd.h"
+#include "snmp-protocol.h"
+#include "snmpd-utils.h"
+#include "snmpd-logging.h"
 
 static struct uip_udp_conn *udpconn;
 
-PROCESS(udp_process_receiver, "UPD test receiver");
-AUTOSTART_PROCESSES(&udp_process_receiver);
+PROCESS(snmpd_process, "SNMP daemon process");
 
 /*-----------------------------------------------------------------------------------*/
-void send(char *r) 
+/*
+ * UDP handler.
+ */
+static void udp_handler(process_event_t ev, process_data_t data)
 {
-	uip_ipaddr_copy(&udpconn->ripaddr, &UDP_IP_BUF->srcipaddr);
-	udpconn->rport = UDP_IP_BUF->srcport;
-
-	u16_t cur = 0;
-	while (cur < strlen(r)) {
-		uip_udp_packet_send(udpconn, r + cur, min(strlen(r + cur), UDP_DATA_LEN));
-		cur += UDP_DATA_LEN;
-	}
-
-	memset(&udpconn->ripaddr, 0, sizeof(udpconn->ripaddr));
-	udpconn->rport = 0;
[…]

### 2010-03-05 trunk/src/snmpd.h

@@ -1,0 +1,10 @@
+#ifndef __SNMPD_H__
+#define __SNMPD_H__
+
+#include "contiki-net.h"
+
+#define LISTEN_PORT 161
+
+PROCESS_NAME(snmpd_process);
+
+#endif /* __SNMPD_H__ */
